### 2024-05-05 dissertation/content/p1_intro.tex
added dissertation

\chapter{The Relevance of Time-Frequency Analyses}

### 2024-05-05 dissertation/content/p2_intro.tex
added dissertation

\chapter{Feature Extraction with Wavelet Transforms}

### 2024-05-05 dissertation/content/p3_intro.tex
added dissertation

\chapter{Wavelet Scattering in Higher Dimensions}

### 2024-05-05 dissertation/content/paper1/main.tex
added dissertation

\chapter{Wavelet Transforms as a Fourier Alternative}

\articleinfo{M.W. Rademan, D.J.J Verself, J.A. Du Preez}{Ecological Informatics}{15 May 2024}{doi}{Underwater passive acoustic monitoring systems record many hours of audio data for marine research, 
making fast and reliable non-causal signal detection paramount. Such detectors assist in reducing the amount of 
labor required for signal annotations, which often contain large portions devoid of signals.\par
Cetacean vocalization detection based on spectral entropy is investigated as a means of vocalization discovery. 
Previous techniques using spectral entropy mostly consider time-frequency enhancement of the entropy measure, and 
utilize the short time Fourier transform (STFT) as its time-frequency (TF) decomposition. Spectral entropy methods 
also requires the user to set a detection threshold manually, which call for knowledge of the produced entropy measures.\par
This paper considers median filtering as a simple, effective way to provide temporal stabilization to the entropy measure, 
and considers the continuous wavelet transform (CWT) as an alternative TF decomposition. K-means clustering is used to 
determine the threshold required to accurately separate the signal/no-signal entropy measures, resulting in a  one-dimensional, 
two-class classification problem. The class means are used to perform pseudo-probabilistic soft class assignment, which is a 
useful metric in algorithmic development. The effect of median filtering, signal-to-noise ratio and the chosen TF decomposition are investigated.\par 
The accuracy and specificity measures of the proposed detection technique are simulated using a pulsed frequency modulated sweep, 
corrupted by a sample of ocean noise. The results show that median filtering is particularly effective for low signal-to-noise ratios. 
Both the STFT and CWT prove to be effective TF analyses for signal detection purposes, each presenting with different advantages and drawbacks. 
The simulated results provide insight into configuring the proposed detector, which is compared to a conventional STFT-based spectral entropy 
detector using manually annotated humpback whale (\textit{Megaptera novaeangliae}) songs recorded in False Bay, South Africa, July 2021.\par
The proposed method shows a significant improvement in detection accuracy and specificity, while also providing a more interpretable 
detection threshold setting via soft class assignment, providing a detector for use in development of adaptive algorithms.%
}

\section{Introduction}

Underwater passive acoustic monitoring (PAM) systems consisting of hydrophones are often 
used to record marine activity \citep{hydrophone, internalsignaldetection, dolphindetection}. 
PAM is a very effective tool which may aid in ecological monitoring, preservation and
research. However, PAM systems present their own of set challenges relating 
to the field of digital signal processing. The recorded vocalizations are often 
from distant sources, which may result in poor signal-to-noise ratios. 
PAM systems record hundreds of hours of audio from multiple hydrophones, 
making it time-consuming to find and label vocalizations manually. Thus, 
reliable and computationally efficient signal detection is of great 
importance for data gathering purposes.

Many cetaceans, such as the humpback whale (\textit{Megaptera novaeangliae}) and southern right whale (\textit{Eubalaena australis}), produce vocalizations that are periodic in nature. These signals may be modelled as amplitude-frequency modulated (AMFM) sinusoids \citep{whalemodel}. Recordings of these vocalizations are useful for conservation studies, like determining migration patterns or population estimates. It has become critical to monitor the effect of industry on marine mammals, and to take measures for ecological preservation. Numerous detection algorithms and monitoring systems have been developed to aid in this venture \citep{glider, rightdetection}.

Recent techniques which detect and/or classify cetacean calls often make use of machine learning (ML) models, 
using deep learning \citep{MLcnn, MLalexnet} or hidden Markov models (HMM) \citep{hmm1, hmm2}. 
Due to the difficulty of obtaining large amounts of annotated data, neural network ML models often 
need to use transfer learning as in \citep{MLalexnet}. ML techniques often require additional feature 
extraction methods to reduce the dimensionality of the audio data, thus adding to computational 
complexity \citep{hmm3}. ML models cannot necessarily operate well outside the conditions of their training data. 
Training ML models may also be a cumbersome task \citep{unsupervised}, making simpler, unsupervised techniques 
which are more computationally efficient the preferred method of sifting through large amounts of data to 
identify points of interest. These points of interest allow biologists to ignore extensive portions of noise in the 
audio, thus greatly reducing manual work, before classifying the audio.

An example of such an efficient detector which is prevalent in PAM audio analysis 
software packages (PAMGuard \citep{PAMGuard} and Raven Pro \citep{raven}), %PAMGUARD CITATION CAUSES ISSUES!
is the band-limited energy detector (BLED) \citep{internalsignaldetection}. 
This detector is simple to compute, making it a popular choice for finding 
points of interest in long audio recordings. BLED has also seen use as part of multi-stage classifier systems \citep{unsupervised}.

Spectral entropy (SE) is another popular and simple method derived from Shannon's 
information entropy \citep{shannon} to perform narrow-band signal detection, and has 
demonstrated success in both speech and cetacean vocalization detection, consistently 
outperforming BLED \citep{entropyJASA, entropyOCEANS, entropysubband}.  At its core, 
spectral entropy utilizes a time-frequency (TF) decomposition and treats each time-slice as a pseudo-probability density. 
SE will decrease in the presence of narrow-band frequency content. This method of narrow-band signal detection was first 
proposed by Powell and Percival in \citep{entropyorig}.

In real data, the SE measure may fluctuate due to changing noise conditions, resulting in less accurate detection. Various methods for speech detection aim to stabilize/enhance the SE as in \citep{entropyOCEANS, entropysubband}, which utilize weighting factors in the SE calculation 
as well as temporal averaging. For signal detection, the researcher must 
determine the appropriate SE threshold manually, which again increases 
labor, and requires some understanding of how to set the detection threshold properly. Furthermore, \citep{entropyOCEANS, entropysubband} only consider the short-time Fourier transform (STFT) as a TF decomposition.

The continuous wavelet transform (CWT) has shown success in many applications, 
including feature extraction in ML models for electrocardiogram classification \citep{ecgcwt}, 
analysis of periodic patterns in lake water levels \citep{lakelevelcwt} and detecting pachyderm 
movements using seismic sensors \citep{seismiccwt}. The CWT has previously been used in structural 
engineering applications as a TF decomposition for SE \citep{cwtentropy1, cwtentropy2}, and may be 
considered as an alternative to the STFT for SE calculation.

Erbe and King \citep{entropyJASA} used SE to detect  cetacean calls in the Canadian Arctic. They demonstrate its effectiveness compared to BLED, which SE outperforms on all accounts. Erbe and King used the simplest form of SE with the STFT, without any enhancement of the entropy measure, which will serve as a baseline comparison for this paper.

\begin{figure}[h!]
	\centering
	\begin{tikzpicture}

		\node (x) at (0,0) {\scriptsize $x[n]$};

		\node[draw,
		fill=white,
		minimum width=0.8cm,
		minimum height=1.2cm,
		align=center,
		right = 0.3cm of x
		] (TF) {\scriptsize TF\\ \scriptsize  Spectrum};

		\node[draw,
		fill=white,
		minimum width=0.8cm,
		minimum height=0.7cm,
		right= 0.5cm of TF
		] (SE) {\scriptsize SE};

		\node[draw,
		fill=white,
		minimum width=0.8cm,
		minimum height=0.7cm,
		right= 0.5cm of SE
		] (MF) {\scriptsize MF};

		\node[draw,
		fill=white,
		minimum width=0.8cm,
		minimum height=0.7cm,
		right= 0.9cm of MF
		] (Sigmoid) {\scriptsize $\sigma$};

		\node[draw,
		fill=white,
		minimum width=0.8cm,
		minimum height=0.7cm,
		below= 0.7cm of MF
		] (KM) {\scriptsize K-Means};

		\node (cs)[right=0.3cm of Sigmoid] {\scriptsize $c_s(H[m])$};

		\draw[->] (x) -- (TF.west);
		\draw[->] (TF.east) -- (SE.west);
		\draw[->] (SE.east) -- (MF.west);
		\draw[->] (MF.east) -- (Sigmoid.west)node[above, midway]{\scriptsize $H[m]$};
		\draw[->] (Sigmoid.east) -- (cs.west);
		\draw[->] (MF.south) -- (KM.north);
		\draw[<-]  (Sigmoid.south) |- (KM.east) node[above, near end]{\scriptsize $\mu_S$, $\mu_{NS}$};

	\end{tikzpicture}
	\caption{Block diagram of the proposed soft-output detector.}\label{fig:systdiag}
\end{figure}

This paper compares the STFT and CWT as TF decompositions for the SE calculation, and proposes median filtering (MF) as a simple, yet effective, method of providing temporal stabilization to the SE measure. Additionally, a means of converting the SE to a pseudo-probability is proposed, using the k-means algorithm as a computationally simple way to distinguish between signal/noise SE segments. The pseudo-probability measure is more interpretable than SE, and is automatically scaled, making it robust against changes in ambient ocean conditions across multiple recorded files, which will typically require manually resetting the SE threshold.  Refer to figure \ref{fig:systdiag} for a description of the proposed solution.

To test the accuracy and specificity of our proposed solution compared to the baseline SE detector, 
this paper measures detector performance on a per-sample basis on the audio. Simulation is used to verify 
and design the proposed technique, and gain some insight on how SNR and hyper-parameters influence the detector in a 
controlled test environment. The proposed detector is also verified with manually annotated data, to demonstrate that 
it performs well in a practical setting. Per-sample accuracy is a necessary measure to test detectors, since software packages (PAMGuard; Raven Pro) 
allow the user to discard detections of short length. Thus, detecting only part of the signal is not suitable for all applications.

We demonstrate empirically that the L1-normalized CWT is a superior 
TF decomposition for similar practical applications for signal detection using SE. 
Median filtering is shown to be a simple and viable form of SE stabilization to improve 
signal detection accuracy and specificity. Additionally, we propose the K-means algorithm 
to provide appropriate scaling for the SE to be converted into a pseudo-probabilistic measure 
of signal presence, which may be used to develop adaptive algorithms and provide more intuitive 
threshold settings for non-experts. To the authors' knowledge, this paper is the first to introduce a 
soft-classification procedure for SE detectors through the k-means algorithm. Additionally, to the authors' knowledge, 
it is the first time the CWT is used in a SE detector for cetacean detection applications.

Improving SE-based detectors are of great importance for performing narrow-band signal detection. 
SE detectors offer good accuracy in identifying periodic signals, while requiring no training, and offering 
some robustness against fluctuation noise profiles. Depending on the TF decomposition parameters used, 
SE allows for accurate time-localized detections, which yields more detailed analysis on vocalization statistics, 
such as call length, and distance between calls, and may assist in creating accurate datasets via data augmentation for supervised ML models.

\section{False Bay Data}
The data used in this paper was recorded via a hydrophone at False Bay, on 1 July 2021 at midnight. Two humpback whale songs were extracted in the time frame between 00:00 and 00:30 hours. The hydrophones were anchored 6 kilometers off the coast. The hydrophone capsule recorded at 15 metres below sea level. The audio is sampled 32 kHz, which is resampled to 2 kHz. This is adequate for preserving the observed Humpback whale vocalizations, which consist of songs in the 150-1000 Hz band. The audio is normalized to have an average power of 1.

\begin{figure}[h!]
	\centering
	\includegraphics[width=0.6\textwidth]{fbdata1}
	\caption{Audio snippet and spectrogram of a humpback song recorded in False Bay. The waveform shown is high-passed at 100 Hz, resampled and normalized. The signal annotations are shown on the spectrogram. The spectrogram is calculated before high-passing the signal.}
	\label{fig:fbdata1}
\end{figure}

Figure \ref{fig:fbdata1} shows a snippet of a recorded whale song. Many noise types can be observed. Low frequency anchor noise, caused by the swaying of the hydrophone in the water due to the anchoring system, contains the majority of signal power. This significantly corrupts all frequency content below 100 Hz. Thus, this noise is filtered out, since it does not influence the signals of interest, and is highly dependent on the anchoring system and the tides.

The remaining noise types are background noise, and impulsive noise, caused by snapping shrimp and fish. All other noise sources, such as waves, fish and boats are classified as background noise. %Figure \ref{fig:fbdata2} shows the complete segments of the two recorded humpback whale songs, which have been normalized, high-passed at 100 Hz to remove the anchor noise and resampled.

Due to the dominating anchor noise, the signal-to-noise ratio (SNR) is only calculated for frequencies above 150 Hz to the Nyquist limit (1 kHz). The recorded whale calls are estimated to have an average SNR of approximately 0 dB, which is calculated in the time-frequency regions of signal presence, utilizing a power spectrogram. The instantaneous SNR is estimated to range from -20 to 30 dB.

% \begin{figure}[h!]
% 	\centering
% 	\includegraphics[width=0.48\textwidth]{realdata_songs}
% 	\caption{High-passed, resampled and normalized audio segments of recorded humpback vocalizations.}
% 	\label{fig:fbdata2}
% \end{figure}

The songs were manually annotated. The starting and ending time of a vocalization is recorded, which therefore only indicates signal presence. Overlapping calls are annotated as one vocalization. The two songs have a combined length of 369 seconds, containing 120 annotated calls. The signals compose 32\% of the recorded samples within the song segments. The remaining 68\% of samples are considered as noise.

Figure \ref{fig:fbhist} show the combined statistics of the humpback songs. Call length and the distance between calls are calculated from the manual annotations. The distance is measured between the endpoint of a vocalization and the start of the next vocalization. These statistics are useful in determining median filter window lengths limits.

\begin{figure}[h!]
	\centering
	\includegraphics[width=0.7\textwidth]{real_data_hist}
	\caption{Histograms indicating the combined call length and distance between calls of the two humpback songs, for use in determining median filter window length limits.}
	\label{fig:fbhist}
\end{figure}

\section{Spectral Entropy}
\subsection{STFT}
Given a discrete-time signal $x[n]$, the windowed STFT may be expressed as

\begin{equation*}
	X[k, m] = \sum_{n=m}^{m + N - 1} w[m-n] x[n] e^{-j 2 \pi k n / N },
\end{equation*}

\noindent where $w$ is the windowing function, $N$ is the window size, $k \in \{0, 1, ..., N-1\}$ is the frequency index and $m$ is the time index. The total number of frequency bins may be artificially increased by padding each window with additional zeros.

The power spectrum $S[k, m]$ may be derived from $X[k, m]$ as

\begin{equation*}
	S[k,m] = \frac{1}{N}X[k,m] X^*[k,m],
\end{equation*}

\noindent which may then be converted to a pseudo-probability density $P[k,m]$, referred to as the spectral distribution:
\begin{equation}
\label{eqn:specdist}
	P[k,m] = \frac{S[k,m]}{\sum_{j=0}^{N-1} S[j,m]}.
\end{equation}

The spectral distribution is used to construct a time-varying entropy measure $H[m]$, based on Shannon's entropy:
\begin{equation}
\label{eqn:entropy}
	H[m] = - \sum_{k=0}^{N/2} P[k,m] \log (P[k,m]).
\end{equation}

It is well known that the uniform distribution is the maximum entropy distribution on a finite interval \citep{maxentropy}.  For spectral entropy, this corresponds to the spectral distribution of white noise. In a sense, it is a measure of certainty for discrete probability distributions: a flat spectral distribution will have high entropy, whereas the spectral impulse from a sinusoid will have low entropy. Note that the summation only occurs to $N/2$ (even $N$ assumed), due to the symmetry of the Fourier transform.

\begin{figure}[!t]
	\centering
	\includegraphics[width=0.6\textwidth]{entropy1}
	\caption{Spectral entropy (STFT) calculated for different frequency bin ranges.}
	\label{fig:entropy1}
\end{figure}

If a sinusoidal signal is known to be within a certain frequency range $f_0 \in (f_1, f_2)$, it may be beneficial to restrict the entropy measure between frequency indices:
\begin{equation*}
	H[m] = - \sum_{k=k_1}^{k_2} P[k,m] \ln (P[k,m]),
\end{equation*}

\noindent where $k_1 = \lfloor N f_1/f_s\rfloor$ and $k_2 = \lfloor N f_2/f_s\rfloor$, with $f_s$ representing the sample frequency. This will reduce the risk of signals within other bands contaminating the entropy measure. However, small frequency ranges are not advised, since a smaller frequency range will decrease the apparent ``flatness" of the measurement, hence lowering the entropy difference between signal/no-signal segments.

% https://ieeexplore-ieee-org.ez.sun.ac.za/document/1041025 MORSE WAVELETS
% https://www.researchgate.net/publication/1900314_On_the_Analytic_Wavelet_Transform

\subsection{CWT}
The CWT is a form of TF decomposition which falls under the theory of multi-resolution analysis (MRA). Wavelet decomposition is a topic of many facets and contain a great deal of mathematical depth. Mallat \citep{waveletbook} describes wavelet theory as an amalgamation of ideas from a variety of scientific fields, which were developed simultaneously.

From a signal processing point of view, the numerically calculated CWT may be interpreted as a set of finite impulse response (FIR) band-pass filters constructed to have a certain bandwidth and time support. This allows for more flexibility in the time-frequency resolution, depending on the chosen filter-bank parameters. The mother wavelet $\psi$ may be considered as the main impulse response (IR) of the filters, which is scaled in time such that its centre frequency coincides with a frequency of interest.

The analysing wavelet used in this paper is the generalized Morse wavelet, described by Olhede and Walden in \citep{morsewavelets, morsewavelets2}, which is defined by its Fourier transform:
\begin{equation*}
	\mathcal{F}\{\psi_{\beta, \gamma} (t) \} = \Psi_{\beta, \gamma}(\omega) = U(\omega) \alpha_{\beta, \gamma} \omega ^ \beta e^{-\omega \gamma},
\end{equation*}

\noindent with $\alpha_{\beta, \gamma} = (\frac{2e \gamma}{\beta})^{\beta/\gamma}$ as a normalizing constant, such that 
\begin{equation}
	\label{eqn:cwt:bpmag}
\Psi(\omega_{\beta,\gamma}) = 2,
\end{equation}

\noindent where $\omega_{\beta,\gamma} = (\frac{\beta}{\gamma})^{1/\gamma}$ is the peak frequency. The time-bandwidth product of the wavelet $P^2 = \beta \gamma$ determines the wavelet time support, which is proportional to $P = \sqrt{\beta\gamma}$. $U(\omega)$ represents the Heaviside step function:
\begin{equation*}
	U(\omega) = \begin{cases}
		1, & \omega > 0 \\
		0, & \omega \le 0
	\end{cases},
\end{equation*}

\noindent which removes all negative frequency components, making the Morse wavelet analytic (contains one-sided frequency information).

The frequency symmetry of the wavelet is controlled by $\gamma$. A symmetric Morse wavelet has $\gamma = 3$. $\beta$ further controls the time support of the wavelet, or equivalently, the filter bandwidth and IR length. \textcolor{black}{Any analytic wavelet with a tunable bandwidth may also be used to yield similar results. The bandwidth must be set narrow enough so that the filters do not significantly overlap given a number of filters per octave. A filterbank that adequately covers the entire frequency range of interest without significant overlap between filters provides optimal results since a sharper frequency resolution creates a lower entropy.}

The CWT of a signal $x(t)$ is defined as:
\begin{equation}
	\label{eqn:cwt}
	X(s,b) = \frac{1}{\sqrt{|s|}} \int_{-\infty}^{\infty} x(t) \psi^*\left(\frac{t-b}{s}\right) dt,
\end{equation}

\noindent where $s$ is the scale parameter which expands/contracts the analyzing wavelet and $b$ is the time-shift parameter.  $\psi^*$ denotes the complex conjugate of $\psi$. Equation (\ref{eqn:cwt}) corresponds to band-pass-filtering $x(t)$ with the impulse response described by the scaled wavelet. The factor $|s|^{-1/2}$ provides L2 normalization, ensuring that energy is preserved with scaling.

%https://www.mathworks.com/help/wavelet/ref/cwtfilterbank.html
In this paper, the MATLAB wavelet toolbox is used to perform CWT and frequency analysis. More information on how this toolbox performs CWT calculation may be found at \citep{cwtmatlab}. The number of band-pass filters is specified by the number of voices per octave, which result in a discrete set of scale parameters $\{s_1, s_2, ..., s_N\}$ which are related to one another by a factor $s_{n+1} / s_n = \sqrt[p]{2}$. This toolbox limits the ratio of the time/bandwidth parameters to $\beta_{max} = 40\gamma$.

Given a starting frequency $f_1$, then
\begin{equation*}
	s_1 =  \frac{\omega_{\beta,\gamma}}{2\pi  f_1 } .
\end{equation*}

When calculating the CWT numerically, the MATLAB wavelet toolbox instead uses L1 normalization of the wavelets, thus preserving amplitude and not energy. The  scaling factor correction $|s|^{-\frac{1}{2}}$ in (\ref{eqn:cwt}) is therefore replaced by a factor of $|s|^{-1}$ in the MATLAB implementation. As such, all band-pass filters have a maximum gain of $2$ which compensates for the analytic nature of the Morse wavelet, since the combined positive and negative frequency components of a real sinusoid captures the amplitude.

The L1 normalized CWT is defined as
\begin{equation}
	X(s,b) = \frac{1}{|s|} \int_{-\infty}^{\infty} x(t) \psi^*\left(\frac{t-b}{s}\right) dt.
\end{equation}

More information on the fast numerical calculation of the CWT may be found in \citep{fcwt}. Figure \ref{fig:cwt_fb_example} shows an example of the resulting MATLAB Morse wavelet filterbank (L1-normalized), and its L2-normalized version.

\begin{figure}[t]
	\centering
	\includegraphics[width=0.7\textwidth]{cwt_fb_example}
	\caption{An example of a CWT filterbank, shown as the L1 and L2 normalized versions , with $\gamma = 10$, $\beta=50$ and 4 voices per octave.}
	\label{fig:cwt_fb_example}
\end{figure}

Given then the discretized/numerical calculation of the CWT with L1 normalization, $X[k,m]$, where $m$ denotes the discrete-time index of the transform, and $k$ denotes the index corresponding to $s$ which refers the centre frequencies ($f_c$) of the band-pass filters, a new pseudo-power spectrum may be utilized for entropy calculation. For a sinusoid with amplitude $A$ and a frequency corresponding to any $k_0$ of the band-pass filters' centre frequencies, the resulting amplitude of $|X[k_0,m]|$ will be $A$ from (\ref{eqn:cwt:bpmag}). Thus, the power of a sinusoid observed within a band is
\begin{equation}
	\label{eqn:cwtps}
	S[k,m] = \frac{1}{2} X[k,m] X^*[k,m]
\end{equation}

\noindent since the average power of a sinusoid with amplitude $A$ is $A^2/2$. For the CWT, (\ref{eqn:cwtps}) refers to the equivalent average power of a sinusoid at time index $m$ and frequency band $k$.

The bandwidth of the filter influences the observed power, especially in the presence of noise. However, the equivalent sinusoidal power may be more useful in most applications. When a sinusoid is present, the equivalent sinusoidal power does not reduce as frequency increases (as with L2 normalization). However, higher frequency wavelets will be more sensitive to noise, due to their increased bandwidth.

SE is calculated by substituting the CWT equivalent sinusoidal power from equation (\ref{eqn:cwtps}) into equations (\ref{eqn:specdist}) and (\ref{eqn:entropy}).

\section{Signal detection}
\subsection{K-means}
Many variants of the k-means algorithm were independently developed by researchers in the 1950's \citep{kmeansoverview}. The name was first coined by MacQueen in \citep{kmeans}. It has a rich history as a method of unsupervised classification, used in a wide set of applications. Its simplicity and interpretability make it a popular choice for many classification problems. \changes{In HMM classifiers, the k-means clustering algorithm is often used to initialize Gaussian mixture models (GMM) prior to data fitting \citep{hmm1, hmm2, hmm3}.} In this paper, k-means is used to determine the class means of the entropy measure given signal/no-signal conditions. This eliminates the need for a user-set threshold on the entropy and provides class means which may be used to construct a pseudo-probabilistic measure of signal presence. \changes{Refer to appendix \ref{app:kmeans} for details on the implementation for this application.}

\subsection{Soft Output}\label{sec:softoutput}

Given the class means $\mu_S$ (signal), $\mu_{NS}$ (no signal) determined by k-means clustering, the k-means decision boundary $\beta$ is the midpoint:
\begin{equation*}
	\beta = \mu_S + \frac{\mu_{NS} - \mu_S}{2}
\end{equation*}

In classical K-means, the assigned class $c_h(H[m])$ results in a hard output:
\begin{equation*}
	c_h(H[m]) = \begin{cases}
		1, & H[m] \leq \beta \\
		0, & H[m] > \beta
	\end{cases}
\end{equation*}

\noindent where a value of $1$ corresponds to signal presence and a value of $0$ representing no signal.

To provide a soft output, a function which smoothly transitions from one class to another in the interval $H[m] \in [\mu_S, \mu_{NS}]$ is required. Since an entropy measurement which is lower than $\mu_S$ is more probable to contain signal and, conversely, an entropy measurement which is greater than $\mu_{NS}$ is more probable to be noise, the transition function $\alpha(t)$ should satisfy the following criteria:
\begin{enumerate}
	\item  $\alpha(t)$ is monotonically decreasing.
	\item Approaches hard classification in the limit: $\lim\limits_{t \rightarrow \infty} \alpha(t) = 0$ and $\lim\limits_{t \rightarrow -\infty} \alpha(t) = 1$.
	\item Equal class probability at the boundary: $\alpha(t - \beta) |_{t=\beta} = \alpha(0) =  \frac{1}{2}$ .
	\item Class assignment symmetry: $\alpha(t) = 1 - \alpha(-t)$.
\end{enumerate}

The sigmoid function $\sigma(x)$ evaluated at $x = -t$ satisfies all the above criteria, defined as
\begin{equation*}
	\sigma(x) = \frac{1}{1 + e^{-x}}.
\end{equation*}

$\sigma(x)$ is proposed in the style of logistic regression \citep{logit}. A piecewise-linear function may also satisfy the above criteria.

The shape of the sigmoid function may be set by specifying the gain $g > 0$ as a hyper parameter. A higher gain will result in a steeper shape, thus attributing more certainty to entropy measurements closer to the class means, and vice versa for lower gains. It may also be beneficial to normalize the entropy measure such that $H[m] = \mu_S$ and $H[m] = \mu_{NS}$ map to $\pm1$. The normalization will ensure consistency in the behaviour of the hyper parameter $g$ over a range of class means. Figure \ref{fig:sigmoid} shows the effect of $g$ on the classification output.

The soft-output classification function $c_{s}(H[m])$ is therefore
\begin{equation*}
	c_{s}(H[m]) = \sigma\left(	-2g	\frac{H[m] - \beta}{\mu_{NS} - \mu_S}	\right).
\end{equation*}

To assign a specific pseudo-probability $c_{s}(H[m]) = p$ that corresponds to signal presence when $H[m] = \mu_S$, then the gain hyper parameter $g$ is calculated as
\begin{equation}
	\label{eqn:sigmoidg}
	g = \frac{1}{2} \frac{\mu_{NS} - \mu_S}{\mu_S - \beta} \ln\left(\frac{1}{p} - 1\right)
\end{equation}

For $g>0$, equation (\ref{eqn:sigmoidg}) requires $p > \frac{1}{2}$.

\begin{figure}[!t]
	\centering
	\includegraphics[width=0.6\textwidth]{sigmoid}
	\caption{Example of the sigmoid function for soft output classification for various gains ($\mu_S=2$, $\mu_{NS}=4$).}
	\label{fig:sigmoid}
\end{figure}

\subsection{Median Filtering}
Median filtering (MF) is often used to suppress ``salt-and-pepper" noise in image processing, and shown past success as a simplistic method of denoising images and signals \citep{mf1, mf2}.

Median filtering  may assist in suppressing the fluctuations of instantaneous spectral entropy measures prior to K-means, but preserve the ``step responses" present in the entropy measure when transitioning from signal to no-signal and vice versa. The entropy of background noise in practical data may not be as consistent as the entropy of GWN. It will therefore help suppress any outliers within the filtering window.

If the window is chosen large enough, it will suppress the entropy fluctuations of signals that are shorter than the half the window length. This may be useful if a restriction on signal length is imposed.

Given a discrete-time signal $x[n]$, and an window of odd length $M$, the median-filtered signal $y[n]$ is
\begin{equation*}
	y[n] = \text{med}(x[n - R], ..., x[0], ..., x[n + R] ),
\end{equation*}

\noindent where $R = \frac{M-1}{2}$ is the half-window length, and $\text{med}(x_1, ..., x_M)$ refers to the median of a sequence of $M$ numbers $x_1$ to $x_M$. Note that if $M=1$, then no median filtering is applied.

\begin{figure}[h!]
	\centering
	\includegraphics[width=0.6\textwidth]{entropymf}
	\caption{Median filtered entropy (STFT) with $M=355$.}
	\label{fig:entropymf}
\end{figure}

Figure \ref{fig:entropymf} shows an example of applying median filtering ($M=355$) on the STFT spectral entropy.

\begin{figure}[h!]
	\centering
	\includegraphics[width=0.6\textwidth]{kmeans1}
	\caption{The resulting class means from the K-means algorithm on median filtered entropy with $M=355$.}
	\label{fig:kmeans1}
\end{figure}

An example of the resulting class means applied to the entropy measure is shown for the STFT in figure \ref{fig:kmeans1}.

\section{Simulation}

\subsection{Purpose}
Measuring classification accuracy on a per-sample basis is difficult to do with practical data, since the exact start and endpoints of a signal cannot be known exactly. Furthermore, the instantaneous SNR of a whale vocalization must be estimated, and does not remain constant.

As such, it is preferable to corrupt known test signals with noise for sample-accurate measurements. An exact SNR also allows for detailed investigation on the effect of SNR on the algorithm, which grants insight into designing a detector for practical data with estimated parameters.

Simulation is used to determine the best TF-decomposition for the practical data with estimated instaneous SNRs and provide insight on how to choose the median filter window length.

\subsection{Setup}
A pulsed FM sweep ($x[n]$) from 150 - 800 Hz is mixed with a sample of noise ($w[n]$) collected in the same time frame as the humpback songs from False Bay. The SNR is calculated only with respect to relevant indices used for entropy calculation, which ranges from 150 - 1000 Hz. \changes{As such, the SNR described in this section is higher than what would be observed practically.} This eliminates anchor noise, which contains most of the signal power, from the SNR calculation. The SNR is calculated using the segments where $x[n]$ is present.

For this simulation, $x[n]$ is non-zero in 10\% of noise samples, pulsed evenly over a 25 second duration, sampled at 2 kHz. \changes{The STFT of the test signal is shown in figure \ref{fig:testsigstft}.}

All STFT's are performed with a Hamming window of length 256 with an overlap of 255. No zero-padding is applied for the FFTs. Figure \ref{fig:testsigstft} shows the STFT of the test signal mixed with the ocean noise with a SNR of 0 dB.

 All CWTs are L1-normalized and are performed with a Morse wavelet with parameters $\gamma = 50$, $\beta = 2000$, over the same frequency range as the STFT (150 - 1000 Hz). 40 filters per octave are used, corresponding to a total of 110 band-pass filters, which is a comparable number of frequency bins used by the STFT entropy calculation (109 bins). The high time-bandwidth product ensures that the filters are narrow enough to reduce contamination between frequency bins. The hyper parameter $g$ is set such that $p = 0.99$.

  \begin{figure}
 	\centering
 	\includegraphics[width=0.6\textwidth]{testsigstft}
 	\caption{Spectrogram of the test signal (0 dB SNR).}
 	\label{fig:testsigstft}
 \end{figure}

 %   \begin{figure}
 % 	\centering
 % 	\includegraphics[width=0.6\textwidth]{testsigcwt}
 % 	\caption{CWT of the test signal used to determine discovery rates (SNR = 0 dB). The CWT L1-normalized and is configured with $\gamma=50$ and $\beta=2000$, with 40 filters per octave.}
 % 	\label{fig:testsigcwt}
 % \end{figure}

\subsection{Measuring Soft-Output Accuracy}

Given an observed signal $y[n] = w[n] + x[n]$ of length $N$, where $w[n]$ is noise and $x[n]$ is a sporadically appearing signal. The presence of $x[n]$ is encoded by the set $S_x = \{ {n_1, n_2, ..., n_L} \}$, where $n_i \in \{0, 1, ..., N -1\}$ is the indices at which the signal is present. The cardinality (size) of the set is denoted as $|S_x|$.

The soft-output classification function $c_s(H[m])$ is used to determine the soft true positive (STPR) and true negative (STNR) discovery rates:
\begin{equation*}
	\text{STPR} = \frac{1}{|S_x|} \sum_{n \in S_x} c_s(H[m])
\end{equation*}
\begin{equation*}
	\text{STNR} = \frac{1}{N - |S_x|} \sum_{n \notin S_x} 1 - c_s(H[m])
\end{equation*}

This is equivalent to calculating the conventional TPR and TNR discovery rates if $c_h(H[m])$ is used instead of $c_s(H[m])$.

 \subsection{Results}
 \label{sec:fmsweep}

Figure \ref{fig:res_snr_mf} shows the STPR and STNR measures for various SNRs and median filter window lengths.

 \begin{figure}[h!]
 	\centering
 	\includegraphics[width=0.6\textwidth]{res_snr_mf}
 	\caption{STFT and CWT soft discovery rates over a range of SNR's and median filter window lengths. The left-hand side of the figure shows the STPR and STNR of the STFT respectively. The CWT soft discovery rates are on the right-hand side.}
 	\label{fig:res_snr_mf}
 \end{figure}

Figure \ref{fig:res_mf_snr} shows the STPR and STNR measures for median filter window lengths for SNRs of -10, -5 and 0 dB. As shown in tables \ref{tab:stpr} and \ref{tab:stnr}, the discovery rates for SNRs above 0 dB are greater than 90\%, with median filtering only showing minor improvements of the measures. Median filtering is therefore critical for lower power signals, which may include the fringes of a whale call waveform, due to ramping in volume.

 \begin{figure}[h!]
	\centering
	\includegraphics[width=0.6\textwidth]{res_mf_snr}
	\caption{STFT and CWT soft discovery rates over a range median filter window lengths for low SNRs. The left and right sides of the figure show the results for the STFT and CWT respectively.}
	\label{fig:res_mf_snr}
\end{figure}

Median filtering increases the STPR for low SNRs. Below -10 dB, it has a negative effect on the STNR. This is acceptable, since most practical signals are expected to operate above -10 dB SNR.

Figure \ref{fig:res_classmeans} shows how the class mean separation is affected by SNR and median filter window length for the STFT and L1-normalized CWT. Median filtering has no significant influence on class separation above 0 dB SNR, but negatively impacts the separation below 0 dB SNR. The STFT exhibits better class mean separation compared to the CWT. Despite this, tables \ref{tab:stpr} and \ref{tab:stnr} show superior CWT performance, due to the better time localization properties of the CWT.

\begin{figure}[h!]
	\centering
	\includegraphics[width=0.6\textwidth]{res_classmeans}
	\caption{STFT and CWT difference in class means ($\mu_{NS} - \mu_S$) over a range of SNRs and median filter window lengths.}
	\label{fig:res_classmeans}
\end{figure}

\begin{table}[h!]
	\caption{Comparing the STPR of the STFT and CWT at varying SNR's and MF Window Lengths\label{tab:stpr}}
	\centering
\begin{tabular}{cclllll}
\hline
	\multirow{3}{*}{} & \multicolumn{1}{c}{\multirow{3}{*}{M}} & \multicolumn{5}{c}{\textbf{Soft True Positive Rate (\%)}}                   \\
	& \multicolumn{1}{c}{} & \multicolumn{5}{c}{SNR}                  \\ \cline{3-7} 
	& \multicolumn{1}{c}{} & -10  & -5           & 0    & 5   & 10   \\ \hline
\multirow{3}{*}{STFT} & 1                    & 45.5          & 74.5          & 89.9          & 94.5          & 96.3          \\
& 301                  & 59.3          & 84.3          & 92.7          & 95.7          & 96.6          \\
& 501                  & \textbf{66.4} & \textbf{88.4} & 94.0          & 96.0          & 96.7          \\ \hline
\multirow{3}{*}{CWT}  & 1                    & 40.8          & 62.0          & 86.7          & 95.5          & 98.1          \\
& 301                  & 52.4          & 75.0          & 93.2          & 97.8          & 98.7          \\
& 501                  & 56.9          & 79.1          & \textbf{95.1} & \textbf{98.0} & \textbf{98.7} \\ \hline
\end{tabular}
\end{table}

\begin{table}[h!]
	\caption{Comparing the STNR of the STFT and CWT at varying SNR's and MF Window Lengths\label{tab:stnr}}
	\centering
	\begin{tabular}{cclllll}
		\hline
		\multirow{3}{*}{} & \multicolumn{1}{c}{\multirow{3}{*}{M}} & \multicolumn{5}{c}{\textbf{Soft True Negative Rate (\%)}}                   \\
		& \multicolumn{1}{c}{} & \multicolumn{5}{c}{SNR}                           \\ \cline{3-7} 
	& \multicolumn{1}{c}{} & -10  & -5           & 0    & 5   & 10   \\ \hline
\multirow{3}{*}{STFT} & 1                    & 87.9          & 95.3          & 96.6          & 96.5          & 96.2          \\
& 301                  & 91.2          & 97.2          & 96.9          & 96.6          & 96.3          \\
& 501                  & \textbf{92.9} & \textbf{97.4} & 96.9          & 96.6          & 96.3          \\ \hline
\multirow{3}{*}{CWT}  & 1                    & 71.0          & 89.0          & 97.1          & 98.5          & 98.5          \\
& 301                  & 70.8          & 95.6          & 98.7          & 98.8          & 98.7          \\
& 501                  & 66.8          & 96.4          & \textbf{98.7} & \textbf{98.8} & \textbf{98.7} \\ \hline
	\end{tabular}
\end{table}

The CWT exhibits poorer STNR and STPR performance for SNRs below 0 dB. Utilizing the equivalent sinusoidal power from equation (\ref{eqn:cwtps}) unevenly weights some of the frequency bins according to filter bandwidth, which lowers the CWT's performance for low SNRs, since the noise power becomes significant compared to the signal power. This makes the L1-normalized CWT ineffective for low SNRs. The reduction in class mean separation compared to the STFT in figure \ref{fig:res_classmeans} may also be explained by this phenomenon.

For SNRs below 0 dB, it is recommended to use the STFT with a large MF window ($M \in [500, 1000]$). Note that this will reduce the capability of detecting shorter signals. For SNRs above 0 dB, the CWT performs best, with a recommended window size greater than 300.

Median filtering has little to no significant effect above 10 dB SNR. If no median filtering is to be used ($M=1$), the STFT is recommended for SNRs below 5 dB and the CWT for SNRs above 5 dB. For a change in sampling frequency, the MF window lengths may be adjusted accordingly.

\subsection{A Note on L2 Normalization}

Figure \ref{fig:res_snr_mf_L2} compares the performance of the L1 and L2-normalized CWTs. The L2-CWT shows better STNR performance than the L1-CWT, since the power scaling effect is not present. However, the L2-CWT does not perform as well on STPR, due to the suppression of sinusoidal power at higher frequencies.

The the STFT outperforms the L2-CWT on all accounts, making the L1-CWT the superior option, since it performs better than the STFT for higher SNRs.

\begin{figure}[h!]
	\centering
	\includegraphics[width=0.6\textwidth]{res_snr_mf_L2}
	\caption{Comparing L1 and L2 CWT normalization for spectral entropy calculation. The left side of the figure shows the results without MF ($M=1$). The right side of the figure shows with MF ($M=501$). Note the large differences between L1 and L2 normalization for STNR and STPR.}
	\label{fig:res_snr_mf_L2}
\end{figure}

\section{Application to the False Bay Data}

From simulation, it is determined that a larger median filter length and the L1-normalized CWT performs best for SNRs above 0 dB. As such, a median filter window lengths of 201 and 501 are tested for the proposed soft-classification technique, and compared to the baseline STFT-based SE (Hamming window of 256 samples; 255 sample overlap) used in \citep{entropyJASA}. No median filter is also tested on the CWT so that the STFT and CWT may be directly compared without enhancements. The detectors operate in the 130-1000 Hz frequency region, with the CWT having 40 filters per octave, $\gamma = 50$ and $\beta = 2000$.

The manual annotations are considered as the ground truth. However, it should be noted that these annotations may suffer from slight inaccuracies on a per-sample basis, which would reflect as a constant offset in the results, thus not contaminating this experiment significantly.

 \begin{figure}[h!]
	\centering
	\includegraphics[width=1\textwidth]{res_realdata_snippet1}
	\caption{A snippet showing the measures produced by the entropy and the proposed soft-output signal detector.}
	\label{fig:res_realdata_snippet1}
\end{figure}

A receiver operating characteristic (ROC) curve is used to show the performance of the proposed technique, as is done in \citep{entropyJASA}. This curve plots the true positive rate (TPR) and the false positive rate (FPR) on the same axes, determined by sweeping over a detection threshold parameter. For the baseline SE, this threshold is swept linearly between the maximum and minimum entropy values. The proposed technique sweeps the pseudo-probability threshold from 0 to 1. The sigmoid gain is set such that $p=0.95$. Figure \ref{fig:res_realdata_snippet1} shows the output the baseline SE and the proposed detector on a snippet of the recorded humpback song.

The ROC curve illustrates the trade-off between probability of detection (TPR) and probability of false alarms (FPR) for a varying detection threshold. Curves which are closer to the upper left corner of the ROC plot indicate greater detection capability, since it reduces the risk of a false detection while remaining accurate (true detections). Accuracy is measured on a per-sample basis, since accurate start and endpoints are important for many applications.

The TPR and FPR measures are calculated using:
\begin{equation*}
	\text{TPR} = \frac{\text{\# True Positives}}{\text{\# Signal Samples}};
\end{equation*}
\begin{equation*}
	\text{FPR} = \frac{\text{\# False Positives}}{\text{\# No-Signal Samples}}.
\end{equation*}

The ROC curve shown in figure \ref{fig:res_realdata_roc} is constructed from the two recording segments containing the humpback songs. The thresholds which produce the ROC curve is shown in figure \ref{fig:res_realdata_thresh}. \changes{For reference, the ROC curve of BLED is also shown, since it is widely used as a comparison in relevant literature \citep{internalsignaldetection, entropyJASA}. BLED is configured to calculate the energy in the 130-1000 Hz region via summation of the STFT produced for the baseline SE detector.}

 \begin{figure}[h!]
	\centering
	\includegraphics[width=0.5\textwidth]{res_realdata_roc}
	\caption{The ROC curves comparing the proposed technique and the baseline STFT SE. Median filter window lengths of 1, 201 and 501 are shown. The sigmoid gain is set such that $p=0.95$. FPR is plotted on a log scale, which better illustrates the TPR differences for small FPR values.}
	\label{fig:res_realdata_roc}
\end{figure}

 \begin{figure}[h!]
	\centering
	\includegraphics[width=0.55\textwidth]{res_realdata_thresh}
	\caption{The thresholds used to produce the ROC curve and the TPR/FPR of the baseline SE and the proposed technique. Median filter window length of 1, 201 and 501 are shown on the right. The baseline entropy is shown on the left. }
	\label{fig:res_realdata_thresh}
\end{figure}

From figure \ref{fig:res_realdata_roc}, it is clear that the proposed technique which uses a median filtering window of 501 samples and the CWT as a TF decomposition outperforms the conventional STFT-based SE. The proposed method shows an increase of around 10\% in sensitivity (TPR) for FPRs  below 3\%, and remains a better detector overall. \changes{As shown previously \citep{entropyJASA}, SE greatly outperforms a standard BLED detector.}

In practice, the detector should be tuned such that the FPR is below 5\%. From figure \ref{fig:res_realdata_thresh}, this corresponds to a threshold of $c_s(H[m]) > 0.1$ for $p=0.95$. This threshold will vary depending on the choice of $p$. Also note that the CWT SE without median filtering ($M=1$) outperforms the STFT in the $\text{FPR} < 5\%$ region, indicating that it is the best TF-decomposition for similar practical applications.

\section{Discussion}
\changes{The proposed detector based on the CWT with median filtering outperforms the traditional SE detector as described in \citep{entropyJASA} for practical applications, verified by real and simulated experiments. The comparison method is done on a per-sample basis, which also indicates increased time localization capabilities of the proposed detector. This is to be expected, since the CWT improves time localization for higher frequencies as a trade-off for filter bandwidth, as opposed to the STFT which has a fixed time-localization as a function of window length \citep{waveletbook}. Median filtering significantly decreases the FPR by suppressing fluctuations in the SE measure. Overall, the proposed detector indicates an approximate 10\% increase of TPR for FPRs below 3\%. As shown by \citep{entropyJASA}, we have again demonstrated that SE detectors outperform BLED on all accounts.}

\changes{It should be noted that the results shown by the ROC curve from figure \ref{fig:res_realdata_roc} are lower compared to Erbe and King's results \citep{entropyJASA}. Our paper's method of measuring performance differs from much of classification literature, since it constrains the window overlap to produce a TF decomposition for every sample, whereas classification accuracy is usually measured on a per-window basis \citep{hmm1, hmm2, hmm3, MLcnn, detectionGLRT} with less overlap. For simple classification problems, accurate audio segmentation and time localization is not necessarily a priority. In \citep{entropyJASA}, the window overlap is not maximal, thus detections are limited to the time localization accuracy determined by the overlap length. The obtained results are more similar to \citep{entropysubband}, which also measure accuracy on a per-sample basis.}

\changes{This paper introduces the k-mean clustering algorithm as a method to provide a pseudo-probabilistic measure of signal presence. This idea may be adapted to any binary signal detector, such as BLED \citep{internalsignaldetection}, to assist in providing a more intepretable measure of signal presence which does not depend on the magnitude of test statistic produced by the detector. The pseudo-probabilistic measure yields a threshold setting which may be more intuitive to non-experts, while also providing a value to be used to create adaptive algorithms which deal with sporadically appearing signals. Possible fields of application include denoising \citep{denoising, denoiseSSN}, instantaneous frequency estimation \citep{frequencyestimation} and classification (for example, as an additional input to a deep neural network) \citep{MLcnn}. Additionally, the CWT may offer better computational efficiency compared to the STFT if per-sample signal segmentation accuracy is required \citep{fcwt}, which could be relevant in preparing datasets for supervised learning.}

\changes{The addition of the k-means algorithm presents another advantage: utilizing the class-mean separation ($\mu_{NS} - \mu_S$) as metric to quickly determine whether there are points of interest within an audio segment. Furthermore, the k-means algorithm may be restricted to converge within a set number of iterations. Whether the algorithm converged indicates if there are points of interest within an audio segment. By segmenting the long recording and testing for appropriate class mean separation and k-means algorithm convergence, the proposed technique provides a fast way of discarding large portions of noise in long audio recordings. This method may be robust to changes in noise conditions, as long as the changes in noise are not experienced in a given segment of a long audio recording. The use of such a system remains to be explored.}

\changes{Automatic threshold setting is a possible extension of this study. This may be achieved by fitting a 2-component GMM to the SE distribution, with k-means clustering acting as the seed values of the fitting procedure, similar to HMM literature \citep{hmm1, hmm2, hmm3}. If a good approximation of the SE distribution is achieved, an appropriate threshold can be automatically determined for a fixed FPR specified by the user \citep{vantrees}.}

\changes{A SE-based detector may not be appropriate for detection in all whale species, since it is restricted to detecting vocalizations with periodic oscillations. This implies that some broadband impulsive noises, such as very short clicks, may not be detected. However, the detector will be insensitive to impulse noise disturbances, such as snapping shrimp \citep{denoiseSSN, internalsignaldetection}.}

\section{Conclusion}

\changes{The performance of whale vocalization detection in the presence of noise based on spectral entropy has been evaluated in this paper. Through simulation and comparison with manual annotations, the effectiveness of the CWT as a TF-decomposition for SE calculation is demonstrated. The use of median filtering also indicates significant improvement of the conventional STFT-based SE detector. Through simulation, this paper shows that the L1-normalized CWT outperforms the STFT and L2-normalized CWT for practical SNRs.}

\changes{The performance of the proposed detector shows an increase of approximately 5-10\% in accuracy (TPR), for false alarm rates (FPR) below 5\%, which is ideally what the detector should be tuned to. Furthermore, we propose a method of converting the SE entropy measure to a pseudo-probability using k-means clustering and automatic scaling, which is both more interpretable to non-experts while providing a foundation for developing adaptive algorithms for sporadically appearing cetacean vocalizations, which must take signal presence into account.}

\section{Acknowledgements}
\changes{This research is funded by the National Research Foundation (NRF) of South Africa (grant numbers: MND210609609887; 129224). The findings of this study and the opinions reflected therein are those of the authors and do not necessarily represent the NRF.}

\begin{subappendices}
	\section{K-means Implementation}
	\label{app:kmeans}
	For a known signal and noise profile, such as white noise, it is possible to estimate the expected values of the spectral entropy measure $H[m]$ given signal/no-signal conditions. However, in a real-world setting, the noise profile cannot be assumed to be a stationary process such as Gaussian white noise (GWN) or colored white noise (CWN). The variability of AMFM signals also further complicates this estimation. Setting appropriate thresholds on the entropy measure have to be done either manually, or via an automatic process.

	K-means is used to estimate these expected values in real-world conditions. A spectral entropy measurement $H[m]$ may belong to 2 classes: no signal ($C_{NS}$); signal ($C_S$). Note that it is required to have signal segments from both classes present in $H[m]$, otherwise this process will fail. It is assumed that the data satisfies the condition $\mathbb{E}\{H[m] | C_{S}\} < \mathbb{E}\{H[m] | C_{NS}\}$, where $\mathbb{E}$ denotes the expected value, implying that signal presence will always lower the entropy measure.

	Class-membership is assigned using a hard class-assignment function:

	\begin{equation*}
		w[m] = \begin{cases}
			1, & H[m] \in C_S\\
			0, & H[m] \in C_{NS}
		\end{cases}.
	\end{equation*}

	Given the class means $\mu_S$, $\mu_{NS}$ and the entropy measure $H[m]$ of length $N$, the assignment function assigns the class with the closest mean:

	\begin{equation} 
	\label{eqn:classassign}
		w[m] = \begin{cases}
			1, & (H[m] - \mu_S)^2 \leq (H[m] - \mu_{NS})^2 \\
			0, & \text{otherwise}
		\end{cases}.
	\end{equation}

	The K-means algorithm for this application is shown below:
	\begin{enumerate}
		\item Assign initial means:
		\begin{equation*}
			\mu_S = \min\limits_{m}\left(H[m]\right);
		\end{equation*}
		\begin{equation*}
			\mu_{NS} = \max\limits_{m}\left(H[m]\right).
		\end{equation*}
		\item Perform class assignment using equation (\ref{eqn:classassign}).
		\item Update the means according to:
			\begin{equation*}
				\mu_S = \frac{\sum_{m=0}^{N-1} w[m] H[m]}{\sum_{m=0}^{N-1} w[m] };
			\end{equation*}
			\begin{equation*}
				\mu_{NS} = \frac{\sum_{m=0}^{N-1} (1-w[m]) H[m]}{\sum_{m=0}^{N-1} 1 - w[m] }.
			\end{equation*}
		\item Repeat steps 2 and 3 until the maximum change in means converges to a specified value $\epsilon$. Terminate if it exceeds a predefined number of iterations.
	\end{enumerate}

	Step 1 guarantees the means converge to their correct classes, and do not swap around in the process. It assumes that the maximum and minimum are not too extreme in that only a single point is associated with the initialization, and that signal/no-signal segments exist. The difference in class means may be used to determine whether a signal is present in the audio clip at all, since the class means will converge to similar values if no signal segments exist.
\end{subappendices}

### 2024-05-05 dissertation/content/paper2/main.tex
added dissertation

\chapter{Introducing Wavelet Scattering}

\articleinfo{M.W. Rademan, D.J.J Verself, J.A. Du Preez}{Journal of the Acoustical Society of America (JASA)}{15 May 2024}{doi}{Current literature trends for whale classification in passive acoustic monitoring systems are mainly concerned with high accuracies obtained via deep learning approaches. This requires preparation of large and accurate annotated datasets, which are often not available when surveying new regions, nor are they wholly accurate when they are. In contrast to classification, many efforts are still being made in improving generalised detectors for the purpose of disregarding noise. As such, there is a disparity between detectors and classification models.

This paper aims to bridge the gap between detection and classification, while exploring the application of wavelet scattering features not yet popularised in the field of passive acoustic monitoring. This paper improves on the classical spectral entropy detector regarding noise adaptivity and uses wavelet scattering for entropy calculation. The same scattering decomposition is further fed to a classification stage: a linear classifier using scattering features provided by windows identified by the entropy detector.

The proposed method works well for small datasets and aims to form the foundation of wavelet scattering classification/detection using non-neural network approaches. The purpose of this work is to provide a practical classification and detection system for new regions in which very little annotated data is available.  }

\section{Introduction}
Underwater \ac{pam} systems are an effective non-invasive tool to monitor marine activity, with many research endeavours to automate cetacean monitoring and detection over the last decades. \Ac{ml} is a particularly prevalent approach to detecting and classifying cetacean calls from \ac{pam} audio \citep{MLreview}, although some recent publications remain focused on traditional approaches that seem to oppose \ac{ml} trends \citep{jacques1, jacques2}. The disparity between the approaches are often due to a lack of data or a high uncertainty of the data quality \citep{casey2019}, which are detrimental to \ac{nn} and deep learning models.

Although the dataset and the species of interest for classification is not the focus of this paper, we apply the proposed method to a dataset containing annotated Antarctic blue whales (\textit{Balaenoptera musculus intermedia}) and fin whale calls (\textit{B. physalus}). Blue and fin whales are considered globally endangered species of whale following whaling activities in the 1900s \citep{blue_whale, fin_whale}. This launched an effort of gathering many hours of \ac{pam} data in the southern Antarctic ocean \citep{sorp_sohn} for the purpose of evaluating detectors and classifiers, which has been made publicly available \citep{casey2017}.

Large annotated datasets may not necessarily be accessible to automate detection and classification in conservation studies, which motivates the further improvement of ``white-box" strategies as opposed to using \acp{nn}. For \ac{poi} detection purposes, \ac{bled}, the \ac{se} detector and the spectrogram correlator \citep{casey2017} are mostly used in practise, often with software such as PAMGuard \citep{PAMGuard}. These detectors require the user to choose many different hyperparameters for the purpose of adaptivity to noise conditions and signal level. They do not necessarily classify signals, and instead pinpoint \acp{poi}, which suffer from many false positive detections. On the other hand, very accurate \ac{nn} approaches exist for a variety of species \citep{alexnet_killer_whales, cnn_multiple_whale_classes, narw_cn_denoising, cnn_fin_whale}, but require large datasets and careful data preparation.

Non-\ac{nn} approaches have also obtained very accurate results, most of which utilise \acp{mfcc} as features \citep{MFCC_HMM_birds, mfcc_hmm_brydes}. A \ac{hmm} is deployed in the aforementioned studies, thereby allowing the modelling of time-dependent sound characteristics without requiring a \ac{nn}.

At the center of the recent work revolving around \ac{pam} classification and detection lies a form of \ac{tf}-decomposition, whether it be for feature extraction or improved detection. Both \ac{mfcc} features and sound detectors (\ac{bled}, \ac{se}) typically use the \ac{stft} as a base for \ac{tf}-decomposition. However, \acp{cwt} may be be used a drop-in replacement for the \ac{stft}, as is shown by the recent work of \citet{mypaper}. \Ac{ws} is an extended form of the \ac{cwt} and may be used to extract features that provide more information than \acp{mfcc}, while also sharing many similarities with the structure of \acp{cnn} \citep{ws}. In fact, \ac{ws} may be considered as a manual definition of a \ac{cnn} front-end, while the time-decimated \ac{cwt} amplitude coefficients (equivalent to first level scattering) mimic \ac{mfcc} features.

\Ac{ws} has not yet been well established in the field of \ac{pam}, but has seen use in many medical \citep{ws_ecg, ecg_ws_svm}, fault diagnosis \citep{ws_fault_diag} and other audio \citep{ws_speech, ws_audio2, ws_audio} applications. A recent study by Michau et. al. has shown exceptionally accurate results in birdsong classification via \ac{ws} features with a \ac{svm} classifier \citep{scattering_birdsong}.

Automated classification is often the end goal of many \ac{pam} systems, but identifying \acp{poi} without classification is a necessary step to assist in the annotation and sound exploration process, especially when surveying new and unknown oceanic regions. As such, generalised detection is often performed with \acp{bled} and \ac{se} detectors \citep{se_dolphin}. \Ac{se} detectors have consistently shown better performance than energy detectors \citep{se_erbe_king}. The previous approach by \citet{mypaper} modifies \ac{se} detectors to stabilize the entropy measure and provide pseudo-probabilistic output, thereby improving endpoint detection accuracy and the interpretability of the detector's output. \ac{se} detectors naturally motivate extension by including feature extraction and classification using the same \ac{tf}-decomposition used for \ac{se} calculation, ultimately reducing computational resources.

A closely related study performed indicates \ac{sota} results in blue and fin whale call detection within given audio segments using a \ac{lstm} \ac{nn} with scattering features \citep{otherScattering}. Our work differs vastly from that of Sattar as it aims to incorporate detectors as well as a non-\ac{nn} model for use in small datasets. Our work classifies on a per-call basis, which is more comparable to other \ac{ml} techniques. We evaluate different blue whale calls and fin whale calls separately and our method distinguishes itself by focusing on the applicability to very small datasets. We evaluate the proposed method in the context of a fully annotated dataset and show it is effective for a variety of call types. Our evaluation method does not include dataset cleaning, thereby allowing for results that reflect on the full spectrum of \ac{snr}, annotation errors, very large noise variations ($>60$ dB) and various vocalisation types. Although we do not compensate for analyst errors as done by \citet{casey2019}, the accuracy of the ``ground truth" labels is discussed at length in this study.

In this paper, we introduce a method which combines an improved traditional \ac{se} detector with \ac{ws} to train a linear classifier. This paper aims to introduce \ac{ws} to the world of \ac{pam} classification and detection, as it is a feature extraction method which has not permeated into this field as of yet. Since \ac{ws} is a very flexible \ac{tf} decomposition, it may serve as input to a variety of traditional detectors, while also providing advantages in the visualisation of waveforms compared to the \ac{stft}. We show that the proposed method is effective for very small datasets (10s of samples) and apply simple chi-square feature selection \cite{chi2_fs} to further improve results for smaller datasets. Refer to Figure \ref{fig:system} for a visualisation of the processes employed in this study.

\begin{figure}[h]
    \centering
    \includegraphics[width=0.9\textwidth]{system_diagram_v2.png}
    \caption{The detection and classifying system. The audio database constists of 1-hour long recordings, which are segmented for computational efficiency. The detection system provides input to the feature extraction and classification steps. The classifier and detector are evaluated separately in this study.}
    \label{fig:system}
\end{figure}

\clearpage

\section{Data}
\label{sec:data}
In this paper, we use an open source dataset recorded over the period of one year in the Casey Islands of Antarctica in 2017, as detailed in \citep{casey2017}. This dataset contains 187 audio files of approximately 1 hour duration each. Recordings took place over the entire year. The hour-long audio recordings are spaced at intervals of approximately 42 hours apart, sampled at 1 kHz. This dataset was chosen instead of the other datasets presented in {\citep{casey2017} due to its reliability, abundance of blue whale annotations and annotation end-point accuracy. Other datasets contained many annotations which were not significantly above the noise level, have inaccurate time localisation regarding the starting and endpoints of the annotation, and/or did not contain enough annotations.

We do not perform any annotation cleaning, as this will unfairly bias our results towards high \ac{snr} detection and classification. However, inaccurate analyst annotations are a significant concern \citep{casey2019}, where some studies discard annotations based on estimated \ac{snr} to improve the ``ground truth" used for accuracy measurements \citep{blue_fin_detection}. Analyst annotation inaccuracies are discussed in greater detail in the later sections of this paper.

For this study, we split each hour-long recording into segments of $2^{18}$ samples after re-sampling to 250 Hz, with the end segment often having less samples. Each segment is approximately 17 minutes in length. Segmentation is performed for memory conservation during computation.

Annotations are provided that give the time-frequency bounding boxes of blue and fin whale vocalisations, among others. In this paper, we mainly focus on the blue whale vocalisations as this is the class with the greatest abundance in this dataset. We also include fin whale vocalisations during detector evaluation. However, due to low \ac{snr} and annotation overlap, very few fin whale samples reach the classification stage. The Casey Islands dataset contains a total of 2988 blue whale annotations, separated as \textit{Bm-A}, \textit{Bm-B}, \textit{Bm-Z} and \textit{Bm-D} types. Figure \ref{fig:Bm_Ex} shows a 1-st level \ac{ws} decomposition ($\mathcal{S}_1$) of high SNR examples of these vocalisation types. Table \ref{tab:numannotations} summarises the number of annotations within this dataset.

\begin{figure}
     \centering
     \begin{subfigure}[]{0.45\textwidth}
         \centering
         \includegraphics[width=\textwidth]{Bm_Ant-A_11.jpg}
     \end{subfigure}
     \hfill
     \begin{subfigure}[]{0.45\textwidth}
         \centering
         \includegraphics[width=\textwidth]{Bm_Ant-B_54.jpg}
     \end{subfigure}
     \hfill
     \begin{subfigure}[]{0.45\textwidth}
         \centering
         \includegraphics[width=\textwidth]{Bm_Ant-Z_98.jpg}
     \end{subfigure}
     \hfill
     \begin{subfigure}[]{0.45\textwidth}
         \centering
         \includegraphics[width=\textwidth]{Bm_D_173.jpg}
     \end{subfigure}
        \caption{High SNR examples of the first level scattering transform to illustrate the characteristics of blue whale calls. Note the logarithmic frequency scale.  \textit{Bm\_Ant\_A} (a) is a constant frequency pulse which occurs around 20 Hz. \textit{Bm\_Ant\_B} (b) and \textit{Bm\_Ant\_Z} (c) are similar to \textit{Bm\_Ant\_A}, but with the addition of frequency modulated tails. \textit{Bm\_D} (d) is a frequency modulated downsweep which typically starts around the 80 Hz and sweeps to 30 Hz.}
        \label{fig:Bm_Ex}
\end{figure}

% \begin{table}[h!]
% \caption{Total number of annotations in the Casey 2017 dataset.}
% \label{tab:numannotations}
% \begin{ruledtabular}  
% \renewcommand{\arraystretch}{0.8}
% \begin{tabular}{lrr} 
% \textbf{Vocalisation}      & \textbf{Number of annotations} & \textbf{Estimated False positives (\%)} \\ \hline
% Bm\_Ant\_A & 1741  & 5 - 10               \\
% Bm\_Ant\_B & 558  & 5 - 10                 \\
% Bm\_Ant\_Z & 119    & 1 - 2               \\
% Bm\_D      & 553     & 8 - 15              \\
% Bp\_20Hz & 78        & 1 - 3             \\
% Bp\_20Plus & 214     & 1 - 3               \\ 
% Total & 3263 & 5 - 10\\ 
% \end{tabular}
% \end{ruledtabular}
% \end{table}

We estimate the number of false positives by investigating a subset of annotations (at least 100 when possible) and determining which are more probable to be noise than a vocalisation according to their $\mathcal{S}_1$ coefficients as portrayed by Figures \ref{fig:Bm_Ex} and \ref{fig:det_example_1}. This task is difficult to perform with absolute precision, so we provide a lower and upper bound for these estimates such that they are a fair reflection of the uncertainty. The number of missed detections require more investigation and expertise and is out of scope for this study. The false positive proportions are indicated in Table \ref{tab:numannotations}, with the total proportion of false positives calculated by considering the proportion of annotations per class.

\begin{table}[h!]
    \centering
\caption{Estimated erroneous (false positives) proportion of annotations per class that could be considered noise. This table does not reflect on whether annotations are of the correct class (i.e., misclassifications). The total proportion of false positives are calculated according to the number of annotations per class.}
\label{tab:errannotations}
% \begin{ruledtabular}    
\begin{tabular}{lr} 
\textbf{Vocalisation}      & \textbf{False positives (\%)} \\ \hline
Bm\_Ant\_A & 5 - 10                 \\
Bm\_Ant\_B & 5 - 10                  \\
Bm\_Ant\_Z & 1 - 2                  \\
Bm\_D      & 8 - 15                  \\
Bp\_20Hz & 1 - 3                   \\
Bp\_20Plus & 1 - 3                   \\
Total & 5 - 10\\ 
\end{tabular}
% \end{ruledtabular}
\end{table}

\clearpage

\section{Wavelet Scattering}
\subsection{Filter Bank Construction}

This paper utilises a modified version of the wavelet scattering transform, as detailed by \citet{ws}. Given a mother wavelet $\uppsi(t)$ with energy concentrated at 1 rad/s, the wavelet dilated by a factor $\lambda$ is notated by
\begin{equation}
\label{eqn:waveletscaling}
    \uppsi_\lambda(t) = \lambda \uppsi(\lambda t).
\end{equation}

The multiplicative scaling of $\lambda$ ensures that dilation results in constant peak Fourier amplitude in the frequency domain, therefore resulting in L1-normalisation. When $\uppsi$ is centered at 1 rad/s, then $\lambda$ is also the center frequency of the filter (in rad/s) specified by $\uppsi$.

 A typical choice of $\uppsi$, also the choice used in this paper, is the Morlet wavelet, given by
\begin{equation}
\label{eqn:morlet}
    \uppsi(t) = (e^{jt} - \beta) \theta_\sigma(t),
\end{equation}
with $\sigma$ indicating the time support of $\uppsi$ and $j$ is the imaginary unit. The shifting factor $\beta$ ensures that $\uppsi(t)$ satisfies the admissibility criterion (zero-mean). A Gaussian with zero mean and standard deviation $\sigma$ is denoted by $\theta_\sigma(t) = \frac{1}{\sigma\sqrt{2 \pi}} e^{-\frac{1}{2}\left(\frac{t}{\sigma}\right)^2}$, thus resulting in $\beta = \frac{\theta_{1/\sigma}(-1)}{\theta_{1/\sigma}(0)}$.

A filter bank is constructed with $Q$ wavelets per octave. To ensure adequate overlap with adjacent filters, we set $\sigma = Q$. The set of dilation factors $\Lambda$ in the filter bank is constructed with $\lambda_{n+1} = 2^{\frac{1}{Q}} \lambda_n$, where $\lambda_0$ sets the first center frequency in the filter bank, and $\lambda_k \le \omega_\text{max}$ is the last filter specified by the maximum frequency $\omega_\text{max}$.

Discretised filter impulse responses are constructed using
\begin{equation}
    \uppsi_\lambda[n] = \begin{cases}
        \uppsi_\lambda\left(\frac{n}{f_s}\right), \ n \in \{-N, ..., N\} \\
        0, \ \text{otherwise}
    \end{cases} , \ \lambda \in \Lambda.
\end{equation}
$N = \left\lceil\frac{5 Q}{\lambda_0} \right\rceil$ provides an adequate coverage of the lowest-frequency wavelet -- 5 standard deviations of  $\uppsi_{\lambda_0}[n]$.

The operator $\mathcal{U}$ is used to define the scalogram of a discrete signal $x[n]$ sampled at a frequency of $f_s$:
\begin{equation}
    \mathcal{U}x[n, \lambda] = \left|x[n] * \uppsi_\lambda[n]\right|,\  \lambda \in \Lambda.
\end{equation}
The scalogram extracts the band-limited amplitudes of each frequency band in $\Lambda$. Since the modulus operator $|\cdot|$ is effectively amplitude demodulating a band-limited analytic signal, the maximum bandwidth resulting from $\mathcal{U}$ is the bandwidth of $\uppsi_{\lambda_\text{max}}$. We define this bandwidth as 2 standard deviations of $\Psi_{\lambda_\text{max}}(\omega)$, corresponding to $2\frac{\lambda_\text{max}}{Q}$ rad/s. This allows the critical down-sampling of $\mathcal{U}x$ by a factor of $d = \left\lfloor\frac{\pi f_s Q}{2\lambda_\text{max}}\right\rfloor$. We denote the down-sampled scalogram as $\mathcal{U}_{\downarrow d}x$.  %should I downsample by 2 before entering the filterbank??

Additional to the scalogram, a further filtering and decimation step is performed to obtain the scattering coefficients. This is accomplished by utilising a Gaussian low-pass filter $\varphi(t)$ with a user-specified standard deviation of $\frac{T}{2\pi}$ seconds (cutoff frequency of $\frac{1}{T}$ Hz). $T$ is referred to as the invariance scale. The scattering operator $\mathcal{S}$ performs filtering of the scalogram:
\begin{equation}
    \mathcal{S}x[n, \lambda] = \mathcal{U}x[n, \lambda] * \varphi[n], \ \lambda \in \Lambda,
\end{equation}

For computational efficiency, the critically down-sampled scattering coefficients are instead calculated as
\begin{equation}
\label{eqn:scattering:downsample}
    \mathcal{S}_{\downarrow r}x[n, \lambda] = \left(\mathcal{U}_{\downarrow d}x[n, \lambda] * \varphi_{\downarrow d}[n]\right)_{\downarrow r}, \ \lambda \in \Lambda,
\end{equation}
with $r = \left\lfloor\frac{f_s}{ 4 T d V}\right\rfloor$ and $V$ indicating the oversampling factor, since $T$ further band-limits the content when chosen such that it is smaller than the time support of $\uppsi_{\lambda_\text{max}}$.

The discretisation of $\varphi(t)$ is identical to that of $\uppsi_\lambda(t)$. The down-sampling factor $r$ down-samples to within 2 standard deviations of $\Phi(\omega)$, as is done with $\uppsi$. The resulting sampling frequency of the scattering coefficients is approximately $\frac{4}{T}$, which may not be exact due to the compounding down-sampling steps of $r$ and $d$.

To implement the filter bank as defined by $\Lambda$, we use the MATLAB Parallel Processing toolbox, with all convolutions carried out as \ac{fft} convolutions on the \ac{gpu}. We pre-compute the \acp{fft} of the filter bank and the low-pass filter for fast execution, thereby only requiring a \ac{fft} on the input signal. This limits the input signal to a specified length, set to the segment size of $L$. Shorter signals are zero-padded and stripped after convolution.

\subsection{Scattering Transform}
A scattering transform is performed by $F$ scattering filter banks defined by $\{\Lambda_1, ..., \Lambda_F\}$. A scattering transform to the $i$'th level is denoted by the operator $\mathcal{S}_i$. We suppress notation indicating time dependence for clarity. The scattering transform performed on a signal $x$ is given as follows:
\begin{gather} 
    \mathcal{S}_0 x (t) = x * \varphi \\ 
    \mathcal{S}_1 x(t, \lambda_1) = \left|x * \uppsi_{\lambda_1}\right| * \varphi, \ \lambda_1 \in \Lambda_1 \\
    \mathcal{S}_2 x(t, \lambda_1, \lambda_2) = \left|\left|x * \uppsi_{\lambda_1}\right| * \uppsi_{\lambda_2}\right| * \varphi, \ (\lambda_1, \lambda_2) \in \Lambda_1 \times \Lambda_2 \\
    \mathcal{S}_i x(t, \lambda_1, ...,  \lambda_i) = \left|\left|...\left|x * \uppsi_{\lambda_1}\right|*...\right| * \uppsi_{\lambda_i}\right| * \varphi, \ (\lambda_1, ..., \  \lambda_i) \in \Lambda_1 \times ... \times \Lambda_i  \label{eqn:scat2}
\end{gather}

Each scattering filter bank defined by $\Lambda_i$ is constructed by the user. In this paper, we neglect the 0'th level scattering transform $\mathcal{S}_0$ since signals of interest are contained in the 20-120 Hz band, thereby not requiring any low-frequency information. We additionally only utilise up to the second level scattering $\mathcal{S}_2$ for feature extraction. The down-sampling steps detailed by equation (\ref{eqn:scattering:downsample}) are only performed by $\mathcal{S}_1$.

The scattering computations used in this study are given by
\begin{gather}
\label{eqn:scattering:s1}
    \mathcal{S}_{1}x [n, \lambda_1] =  \left(\mathcal{U}_{\downarrow d}x * \varphi_{\downarrow d}\right)_{\downarrow r}, \ \lambda_1 \in \Lambda_1; \\
    \label{eqn:scattering:s2}
    \mathcal{S}_{2}x [n, \lambda_1, \lambda_2] =  \left(\left| \mathcal{U}_{\downarrow d}x * \left(\uppsi_{\lambda_2}\right)_{\downarrow d}\right| * \varphi_{\downarrow d}\right)_{\downarrow r}, \ (\lambda_1, \lambda_2) \in \Lambda_1 \times \Lambda_2.
\end{gather}
Refer to Figure \ref{fig:scattering} for a visualisation of the successive filtering steps to perform a scattering computation.

\begin{figure}[h]
    \centering
    \includegraphics[width=0.9\textwidth]{scattering.png}
    \caption{An illustration of the successive filtering operations performed by a scattering transform. Nodes in colour represent the extracted scattering coefficients, which is down-sampled according to an invariance scale specified by the user.}
    \label{fig:scattering}
\end{figure}

For computational efficiency, it is not required to to calculate all paths for $\lambda_2$, since the lower frequency filters defined by $\Lambda_1$ have small bandwidths, implying their demodulated coefficients may not have energy contained in all the paths defined by $\Lambda_2$. In this study, we only compute the $\mathcal{S}_2$ scattering coefficients when 
\begin{equation}
\label{eqn:restrictlambda2}
    \lambda_2 - \frac{\lambda_2}{Q_2} <  \frac{\lambda_1}{Q_1},
\end{equation}
where $Q_1$ and $Q_2$ are the number of filters per octave for each filter bank defined by $\Lambda_1$ and $\Lambda_2$. A similar optimisation is also performed by \citet{ws_audio}. The filter-banks used in this paper is shown in Figure \ref{fig:fb}.

The second filter banks in equations (\ref{eqn:scattering:s1}, \ref{eqn:scattering:s2}) is sampled at a rate of $f_{ss} = \frac{f_s}{d}$. This implementation differs from the fast computation suggested by \citet{ws_audio}. Typically, lower frequency filters are arranged so that the maximum time support does not exceed the time support of $\varphi$. Therefore, lower frequency filters are linearly spaced, whereas higher frequency filters are exponentially spaced when their time support become smaller than $T$. In the implementation for this paper, all filters are exponentially spaced, regardless of their time support exceeding $\varphi$. The resulting information leakage is largely irrelevant in the classification step for this paper, as later detailed in Section \ref{sec:cls}. For models dependent on time information, such as \acp{hmm} or \ac{lstm} networks, information leakage may affect results and the restriction of the time support of $\uppsi$ specified by \citet{ws_audio} should rather be used.

% In summary, we construct the filter banks using
% \begin{gather}
% \label{eqn:configfb1}
% \Lambda_1 = \left\{\lambda_n \ \middle| \lambda_{n+1} = 2^\frac{1}{Q_1}\lambda_n,\ n \in \mathbb{N}_0,\ \lambda_0 = 2 \pi  f_\text{min},\ \lambda_\text{max} \leq 2 \pi f_\text{max} \right\},\\
% \label{eqn:configfb2}
% \Lambda_2 = \left\{ \lambda_n \ \middle| \lambda_{n+1} = 2^\frac{1}{Q_2}\lambda_n,\ n \in \mathbb{N}_0,\ \lambda_0 = \frac{2\pi}{T},\ \lambda_\text{max} \leq \frac{\sup (\Lambda_1)}{Q_1} \right\}.    
% \end{gather}

\begin{figure}[h]
    \centering
    \includegraphics[width=1\textwidth]{fb.jpg}
    \caption{The filter banks used in this study, with $Q_1 = 16$ and $Q_2 = 4$. The filter bank used to calculate $\mathcal{U}_1$ is shown on the left and the $\mathcal{U}_2$ filter bank on the right. Note that the maximum magnitude of 2 accounts for the analytic nature of the wavelet, thereby capturing the true frequency amplitudes in a signal.}
    \label{fig:fb}
\end{figure}

\clearpage
\section{Improved Spectral Entropy Detector}
The \ac{se} detector described by \citet{mypaper} is improved in this section. Rademan shows empirically that the \ac{cwt} yields improved results compared to the \ac{stft} for practical \acp{snr}. As such, we utilise the first level scattering coefficients $\mathcal{S}_{1}$, since it is equivalent to a down sampled version of the \ac{cwt}. We improve on the detector posed by Rademan by introducing true probabilistic output and spectral noise whitening.

\subsection{Spectral Entropy}
\label{sec:spectral_entropy}
Given a discrete-time signal $x[n]$, the first level scattering coefficients are expressed as
\begin{equation*}
	X[k, m] = \mathcal{S}_1x[m, \lambda_k],\ \lambda_k \in \Lambda_1
\end{equation*}
where $k$ is the frequency index and $m$ is sample index. This notation is adopted to emphasise that $X$ is a \ac{tf} decomposition similar to a \ac{stft}.

The power scalogram $S[k, m]$, which is proportional to the equivalent sinusoidal power per frequency band, may be derived from as $S[k,m] = X^2[k, m]$. $S$ is used to construct a time-varying entropy measure $H[m]$, based on Shannon's entropy:
\begin{gather}
\label{eqn:specdist}
	P[k,m] = \frac{S[k,m]}{\sum_{j=0}^{|\Lambda_1|-1} S[j,m]}; \\
\label{eqn:entropy}
	H[m] = - \sum_{k=0}^{|\Lambda_1|-1} P[k,m] \log (P[k,m]).
\end{gather}

% This detector is highly effective under the following assumptions:
% \begin{enumerate}
%     \item  Signals of interest produce scattering coefficients which are sparse in the frequency domain, much larger than noise coefficients. This holds true for periodic and band-limited transient signals. This detector therefore measures the extent of sparsity in the frequency domain over the time invariance scale specified by $T$.
%     \item The noise \ac{psd} is smooth and relatively consistent over the considered time period.
% \end{enumerate}

% Given the aforementioned signal and noise assumptions are true, distinct high and low values in $H[m]$ are created, corresponding to noise and signal respectively.

\subsection{Improving SE Robustness}

Pre-whitening the power spectrum is an effective method of increasing the robustness between signal and noise entropy measures, and serves to stabilize the noise entropy level by enforcing the assumption of stable noise \ac{psd}. To accomplish the whitening, a robust noise \ac{psd} estimator which can function with and without signal presence is required.

A rolling median filter operator $\text{RM}_v^B$ acting on a signal $y[v, w]$ with respect to the index $v$, is defined as
\begin{equation}
    \text{RM}_v^B y[v, w] = \underset{z}{\text{median}}(y[z,w]), \ z \in \left\{v - \frac{B-1}{2}, ..., v, ..., v + \frac{B-1}{2}\right\},
\end{equation}

\noindent where $B$ is an odd number describing the length of the filtering window. Similarly, we define a rolling average operator $\text{RA}_v^B$ as
\begin{equation}
    \text{RA}_v^B y[v, w] = \frac{1}{B}\sum_z y[z,w], \ z \in \left\{v - \frac{B-1}{2}, ..., v, ..., v + \frac{B-1}{2}\right\}.
\end{equation}

Under the signal and noise assumptions specified in subsection \ref{sec:spectral_entropy}, the coefficients of potential signals are first suppressed by median filtering over the frequency indices $k$, followed by another rolling median estimate over time indices $m$ to suppress any transient signals. Finally, a rolling average calculates the time-varying noise PSD. We utilise the amplitude spectra in this estimation, as we find this works better in practise as squaring increases the noise variance and amplifies outliers.

\begin{equation}
\label{eqn:psdestimate}
    \bar{S}[k, m] = \text{RA}_m^{F} \left( \text{RM}_m^{D} \left(\text{RM}_k^{B} \sqrt{S[k,m]}\right) \right)^2.
\end{equation}
In this paper, we set $B=9$ frequency bins, $D$ to a value corresponding to 5 seconds, and $F$ to 60 seconds.

% Refer to Figure \ref{fig:noise_estimate_ex} for a visualisation of the robust noise \ac{psd} estimator from equation (\ref{eqn:psdestimate}).

% \begin{figure}[h]
%     \centering
%     \includegraphics[width=0.9\textwidth]{noise_estimation.eps}
%     \caption{An illustration of the noise estimation technique. This example contains varying noise conditions with blue whale calls are shown. The original $\mathcal{S}_1$ coefficients are shown on the left, with the noise estimate on the right. Note how the noise estimation method adequately suppresses tonal and transient signals.}
%     \label{fig:noise_estimate_ex}
% \end{figure}

A whitened power decomposition is calculated:
\begin{equation}
\label{eqn:whitened_scalogram}
    S_w[k,m] = \frac{S[k,m]}{\bar{S}[k, m]}.
\end{equation}
$S_w$ is used instead of $S$ to produce the whitened spectral distribution $P_w$ and \ac{se} $H_w$ with equations (\ref{eqn:specdist}) and (\ref{eqn:entropy}) respectively. As in the related study by \citet{mypaper}, a rolling median filter is used as an effective method to stabilize the \ac{se} measure:
\begin{equation}
    H_w'[m] = \text{RM}_m^{R} (H_w[m]).
\end{equation}

% \subsection{Noise Conditions}
% The noise conditions estimated by equation (\ref{eqn:psdestimate}) are evaluated in this section. The importance of adaptive algorithms is demonstrated by referring to the volatility of the noise in False Bay.

% All power spectrograms in this study are calculated using a Blackman window with 512 samples (256 ms), with 448 samples (224 ms) of overlap. The noise estimate $\bar{S}[k]$ is calculated for each segment with $L = 15$ and $h[k]$ as a Gaussian smoothing filter with a standard deviation of 5 samples. The detector is configured to only consider frequencies above 30 Hz, since this is where all the signals of interest may lie.

% Figure \ref{fig:psdovertime} illustrates how the noise \ac{psd} varies over time. Large variations are present, with the total noise power showing some periodic behaviour over time corresponding to tide, weather and environment changes. Total noise power is calculated by summing over the \ac{psd} estimated for frequencies above 30 Hz.

% \begin{figure}[h]
%     \centering
%     \includegraphics[width=\textwidth]{psd_over_time.pdf}
%     \caption{The noise conditions over time for the test set. All noise magnitude measurements are normalised according the maximum \ac{psd} observed in the test set. The range of the observed noise vary from 26-60 dB for each of the frequency bins.}
%     \label{fig:psdovertime}
% \end{figure}

\subsection{GMM Automatic Threshold Selection}
\Ac{pam} system recordings which span multiple days are bound to be subjected to many noise conditions when tides, weather and the presence of wildlife change. As such, automatic threshold selection for small audio segments which take into account the specific conditions of the segment is paramount, since a fixed threshold will not necessarily yield good results over the entire database.

In the previous study by \citet{mypaper}, the k-means algorithm \citep{origkmeans} is used to find an adaptive estimate for the low and high levels of an audio segment and generate a pseudo-probabilistic output of signal presence. This pseudo-probability is a first step in normalizing the entropy, but fails to account for the variance of the low and high levels respectively, since the k-means algorithm assumes spherical (equal variance) clusters without accounting for the prior probability of a cluster (i.e., the proportions of signal and noise within the audio segment).

Although this allows for normalized measure between 0 and 1, there is still some uncertainty on the threshold selection, and it requires a hyper-parameter to  ``squeeze" the measures around the equi-probability point (0.5) with a specific steepness.

To improve on threshold selection, this paper models the whitened and median filtered \ac{se} measure $H_w'$ with the probability distribution $p(H_w')$ as a mixture of Gaussians. A \ac{gmm} is fitted using the \ac{em} algorithm. The \ac{gmm} model accounts for each \ac{se} level's variance and the prior probability of each level occurring, assuming independence across all time indices $m$:
\begin{gather}
    H_w' | \mathcal{C} = C_n \sim \mathcal{N}(\mu_n, \sigma_n^2) \\   
    H_w' | \mathcal{C} = C_s \sim \mathcal{N}(\mu_s, \sigma_s^2)
\end{gather}

The observable probability density of $H_w'$ is found through marginalization
\begin{equation}
    p(H_w') = p(C_n)\mathcal{N}(H_w' | \mu_n, \sigma_n^2) + p(C_s)\mathcal{N}(H_w' | \mu_s, \sigma_s^2),
\end{equation}
 where $(\mu_n,\mu_s)$ and $(\sigma_n, \sigma_s)$ are the noise and signal entropy means and variances respectively. $\mathcal{C} \in \{C_n ,C_s\}$ represents the noise and signal levels of $H_w'$. The class prior probabilities $p(\mathcal{C})$ are given by the mixture weights of the \ac{gmm}. The notation $H_w'$ is used to express a random variable, over which all $H_w'[m]$ are independent observations indexed by $m$. The restriction $\mu_n > \mu_s$ is utilised, since the noise entropy measure is always greater than the signal entropy measure under the signal/noise assumptions.

To find the probability of an entropy measure belonging to the signal class $C_s$, Bayes' rule is employed:

\begin{equation}
\label{eqn:entropyposterior}
    p(C_s|H_w') = \frac{p(H_w' | C_s) p(C_s)}{p(H_w')}.
\end{equation}

% Due to $p(C_s|H_w')$ being a true probability measure which accounts for class mean, variance and prior probability, a signal indicator function $\mathcal{I}$ for each entropy measure $H_w'[m]$ is trivially constructed:

% \begin{equation}
% \label{eqn:indicator}
%     \mathcal{I}(H_w'[m]) = \begin{cases}
%     1, & p(C_s|H_w'[m]) \ge p_t \\
%     0, & \text{otherwise}
%     \end{cases}
% \end{equation}

% The threshold value of $p_t$ is indirectly a scaled automatic threshold selection, since the \ac{gmm} parameters automate the shifting and scaling process for each considered audio segment.

\subsection{Problems with the Posterior}
Equation (\ref{eqn:entropyposterior}) has complex behaviour with respect to its parameters determined by the \ac{gmm}. It is known a priori that a higher entropy corresponds to a larger probability of noise, thus requiring equation (\ref{eqn:entropyposterior}) to be a monotonically decreasing function over the entire observed entropy range of audio segment of interest. To achieve this, \citet{mypaper} uses a sigmoid function which is shaped by the k-means algorithm and a hyper-parameter. Most audio segments from the tested database do not satisfy the condition of monotonocity of the posterior.

% Refer to Figure \ref{fig:post_problems} for an example of a \ac{gmm} model that does not adhere to these conditions.

% \begin{figure}[h]
%     \centering
%     \includegraphics[width=1\textwidth]{post_problems.eps}
%     \caption{Non-monotonic behaviour of the posterior signal probability as a function of entropy. This example utilises $\mu_n = 4$, $\mu_s = 3$, $\sigma_n = 0.3$, $\sigma_s = 0.7$, $p(\mathcal{C}_n) = 0.6$. Probability densities are shown on the left and the posterior function is shown on the right. Clamping is illustrated with the dashed line on the right.}
%     \label{fig:post_problems}
% \end{figure}

% \begin{figure}[h]
%      \centering
%      \begin{subfigure}{0.45\textwidth}
%          \centering
%          \includegravarphics[width=\textwidth]{det_example.pdf}
%          \caption{Detection using the modified posterior. }
%          \label{fig:det_ex}
%      \end{subfigure}
%      %
%      \begin{subfigure}{0.45\textwidth}
%          \centering
%          \includegravarphics[width=\textwidth]{post_example.pdf}
%          \caption{The unmodified posterior function.}
%          \label{fig:post_ex}
%     \end{subfigure}
%     \caption{An example illustrating the proposed detector output with whitening and posterior modification and it's corresponding non-monotic posterior function when left unmodified. The spectrogram shown (\ref{fig:det_ex}) is suspected to portray fish, which ``hum" at a near-constant frequency and power at irregular intervals for long periods of time. The humming is accurately identified as a signal with high probability by the detector, despite the poor average estimated \ac{snr} of 7 dB (calculated above 100 Hz to neglect anchor rumble in the estimation). The detector also detects anchor rumble since it presents as narrow-band noise. Note that only half of the audio segment is shown in this example. The unmodified posterior function (\ref{fig:post_ex}) in this example clearly illustrates the non-monotonic behaviour, even within the range of the observed entropy levels (shown as the dotted lines). The signal (blue) and noise (orange) Gaussian range estimates in \ref{fig:post_ex} are shown to visualize the width of the distributions. The width of the signal class crosses over the noise class, which presents the need to modify the posterior function with clamping at the turning point.}
%      \label{fig:post^problems} 
% \end{figure}

This paper proposes that the posterior is clamped after reaching the monotonicity limits. Refer to appendix \ref{app:mono} for the derivation of these limits. With this modification made, it is guaranteed that the posterior is a piece-wise function which is either decreasing or remains constant.

\begin{equation}
p_{\text{clamp}}(C_s | H_w') = 
    \begin{cases}
        p_\text{max},&  \sigma_s^2 > \sigma_n^2 \ \wedge \  x < \theta \\
        p_\text{min},& \sigma_s^2 < \sigma_n^2 \ \wedge \ x > \theta\\
        p(C_s | H_w'),&  \text{otherwise}
    \end{cases},
\end{equation}
where
\begin{equation}
    \theta = \frac{\sigma_s^2\mu_n - \sigma_n^2\mu_s}{\sigma_s^2 - \sigma_n^2},
\end{equation}
with $p_\text{max}$ and $p_\text{min}$ as the maximum and minimum values of the posterior function that is observed within the range of $H_w'$.
% \begin{gather}
%     p_\text{max} = \underset{m}{\max} \ p(C_s | H_w'[m]); \\
%     p_\text{min} = \underset{m}{\min} \ p(C_s | H_w'[m]).
% \end{gather}
$p_\text{max}$ and $p_\text{min}$ are numerically approximated\footnote{The maximum and minimum entropy values in the posterior distribution is calculated by sampling 1000 regularly spaced points in the observed entropy range and finding the maximum and minimum of the sampling points. For a more efficient approach, Newton's method may be used.}. For an example of the output of the entropy detector, refer to Figure \ref{fig:det_example_1}.

\begin{figure}[]
    \centering
    \includegraphics[width=1\textwidth]{detector_example_2.eps}
    \caption{An example of the output of the detector taken from a densely populated vocalisation region of the dataset. The annotation bounding boxes are also indicated (top). In this example, the detector successfully detects all vocalisations. The whitened SE measure is shown (middle) with the posterior probability of signal (bottom).}
    \label{fig:det_example_1}
\end{figure}

% \begin{figure}[]
%     \centering
%     \includegraphics[width=1\textwidth]{detector_example_1.eps}
%     \caption{An example of the output of the detector similar to that of Figure \ref{fig:det_example_1}. Note how the detector also detects two possible low-SNR blue whale vocalisations (not in the annotated dataset) around $t = 95$ and $t = 120$ seconds. The detector cannot distinguish between vocalisation and noise segments fulfilling the detection criteria of $\mathcal{S}_1$ coefficient sparsity. This example section does not contain any annotations, indicating missed annotations by the analyst.}
%     \label{fig:det_example_2}
% \end{figure}

% \clearpage
% \section{Noise}
% \label{sec:noise}
% This section aims to illustrate the variability of noise conditions present in the dataset. We utilise 1-st level \ac{ws} to calculate the noise \ac{psd} over the 15-120 Hz frequency range. Given the noise estimate $\bar{S}$ specified in equation (\ref{eqn:psdestimate}), we convert the L1 normalised scattering coefficients to L2 normalised, thereby preserving the magnitude of energy contained in the filter:

% \begin{equation}
% \label{eqn:psd}
%     \text{PSD}_\text{dB}[k, m] = 20 \log_{10}\frac{\bar{S}[k, m]}{\lambda_k},\ \lambda_k \in \Lambda_1.
% \end{equation}

% For each segment, we calculate the mean and standard deviation to illustrate the noise profile and its withinW-segment variation:
% \begin{gather}
%     \mu_\text{dB}[k] = \frac{1}{M} \sum_m \text{PSD}_\text{dB}[k, m]; \\
%     \sigma_\text{dB}[k] = \frac{1}{M} \sum_m \left(\text{PSD}_\text{dB}[k, m] - \mu_\text{dB}[k]\right)^2.
% \end{gather}

% \begin{figure}[h]
%      \centering
%      \begin{subfigure}[]{0.45\textwidth}
%          \centering
%          \includegraphics[width=\textwidth]{noise_over_time.eps}
%          \caption{Normalised noise power over time.}
%          \label{fig:noise_time}
%      \end{subfigure}
%      \hfill
%      \begin{subfigure}[]{0.45\textwidth}
%          \centering
%          \includegraphics[width=\textwidth]{noise_std.eps}
%          \caption{Standard deviation of the noise estimate within each segment.}
%          \label{fig:noise_std}
%      \end{subfigure}
%      \hfill
%      \begin{subfigure}[]{0.38\textwidth}
%          \centering
%          \includegraphics[width=\textwidth]{noise_average_profile.eps}
%          \caption{The average noise profile and its maximum and minimum bounds within the dataset.}
%          \label{fig:noise_profile}
%      \end{subfigure}
%      \caption{The noise estimates of each segment for the entire dataset. Note that the ridges in (a) are a result of the non-continuous nature of the recordings of this dataset as detailed in section \ref{sec:data}. Some segments contain large noise variations, as seen in (b). The average and range of the noise profile for the entire dataset is seen in (c).}
%      \label{fig:noise}
% \end{figure}

% Refer to Figure \ref{fig:noise} for an illustration of the noise profile over the recorded segments. A large variation in noise power is observed, with low frequency noise power changing by up to 60 dB over the year-long time period. As such, any deployed detector must be adaptive to these conditions, since a non-adaptive threshold will fail for many segments. The method proposed in this paper indirectly incorporates an adaptive threshold using a \ac{gmm} fit to the \ac{se} measure.

% Segments with large noise variation in low frequencies may also indicate blue/fin whale choruses, which are the layered sounds of many whales vocalising at once \citep{casey2017}. Annotations in these sections are likely not very accurate, as the entire audio segment contains vocalisations.

\subsection{Method}

We measure the performance of the detectors via \ac{roc} curves, by testing each annotated sample in the power scalogram $S$. No annotation cleaning is performed and all audio files are included. Annotations which have bounding boxes that do not intersect with the 15-120 Hz band are neglected, since the detectors do not operate outside this bandwidth.

Three \ac{se} detectors are evaluated: \ac{gmm} fitted on the non-whitened $\mathcal{S}_1$ coefficients; \ac{gmm} on the whitened $\mathcal{S}_1$ coefficients; and k-means fitted on the non-whitened $\mathcal{S}_1$ coefficients as performed by \citet{mypaper}. Comparing these detectors will validate that modelling the entropy with a \ac{gmm} and applying whitening yields incremental improvements from the original k-means implementation. As a baseline, we also compare the results to a \ac{bled} which sums over the whitened $\mathcal{S}_1$ coefficients. This form of \ac{bled} is more akin to the detector implementations present in \ac{pam} software, as it indirectly adapts to noise conditions during the whitening process. The whitened \ac{bled} additionally has its energy median filtered in the same manner as the \ac{se} detectors, such that a fair comparison is obtained with respect to whitening and filtering.

To perform unmodified energy detection with wavelet scattering, the average energy $E$ of each time slice calculated using
\begin{equation}
\label{eqn:bled_power}
    E[m] = \frac{1}{\left| \Lambda_1 \right|}\sum_{\lambda_1 \in \Lambda_1} \left(\frac{\mathcal{S}_1 x [m, \lambda_1]}{\sqrt{\lambda_1}}\right)^2,
\end{equation}
which converts the wavelet L1 normalisation from Equation (\ref{eqn:waveletscaling}) to L2.

Two methods of evaluation are employed. We perform a sample-by-sample evaluation to produce a \ac{roc} curve. This method will measure the proportion of samples which has been detected and therefore penalise sporadic detections. However, this is not an optimal method of measuring the detector. The dataset does not contain accurate annotation bounding boxes, thereby unfairly penalising the detectors based on annotation endpoint accuracy. Additionally, we count the proportion of annotated calls that have at least 1 detected time sample, and plot it against the proportion of noise/un-annotated samples which have been detected. The true efficacy of the detection is therefore determined, however, sporadic detections are not penalised.

\subsection{Results}
\label{sec:detector_results}

Figure \ref{fig:det_roc} shows the \ac{roc} curve for detection on a sample-by-sample basis. Incremental performance improvement is demonstrated by adding a \ac{gmm} and enhancing the detector with whitening. In Figure \ref{fig:det_calls}, we show that the proposed detector is able to detect the majority calls while detecting fewer noise.  As illustrated in many previous studies, \ac{se} greatly outperforms a traditional \ac{bled} detector \citep{se_erbe_king}. Whitening is also shown to greatly improve the capabilities of the \ac{bled} detector.

% \begin{figure}[h]
%     \centering
%     \includegraphics[width=0.6\textwidth]{roc.eps}
%     \caption{The performance of the detectors when compared on a sample-by-sample bases. \ac{se} detectors are indicated as a solid line, while \ac{bled} is dashed. The whitened and median filtered \ac{bled} performs similarly to the k-means \ac{se} detector (unwhitened), but is ouperformed by the whitened \ac{gmm} \ac{se} detector. Adding a \ac{gmm} without whitening indicates incremental improvement over using k-means. }
%     \label{fig:det_roc}
% \end{figure}

% % \begin{figure}[h]
% %     \centering
% %     \includegraphics[width=0.6\textwidth]{threshold.eps}
% %     \caption{The proportion of detected blue and fin whale calls and noise samples as a function of the selected probability threshold.}
% %     \label{fig:det_thresh}
% % \end{figure}

% \begin{figure}[h]
%     \centering
%     \includegraphics[width=0.6\textwidth]{calls.eps}
%     \caption{The performance curve of the detectors expressed as the percentage of detected noise samples versus the percentage of detected calls. \ac{se} detectors are indicated as a solid line, while \ac{bled} is dashed. The whitened \ac{se} with \ac{gmm} detector significantly outperforms all other detectors.}
%     \label{fig:det_calls}
% \end{figure}

\begin{figure}[h]
     \centering
     \begin{subfigure}[t]{0.45\textwidth}
    \centering
    \includegraphics[width=1\textwidth]{roc.eps}    
     \caption{} \label{fig:det_roc}
     \end{subfigure}
     % \hfill
     \begin{subfigure}[t]{0.45\textwidth}
    \centering
    \includegraphics[width=1\textwidth]{calls.eps}    
    \caption{} \label{fig:det_calls}
     \end{subfigure}
     \caption{Results of the evaluated detectors. \ac{se} detectors are indicated as a solid line, while \ac{bled} is dashed. (a) The performance of the detectors when compared on a sample-by-sample bases. (b) The performance curve of the detectors expressed as the percentage of detected noise samples versus the percentage of detected calls. The whitened \ac{se} with \ac{gmm} detector significantly outperforms all other detectors.}
\end{figure}

\begin{table}[h] 
    \centering
\caption{The number of missed calls as a result of a non-converging \ac{gmm} fit.}
\label{tab:convergence}
\begin{tabular}{lc} 
\hline
\textbf{Vocalisation}       & \textbf{Missed vocalisations}  \\ \hline
SE GMM                      & 196 (5.8\%)                      \\
Whitened SE GMM             & 43 (1.2\%)                    \\ \hline
\end{tabular}
\end{table}

\subsection{Discussion}

Results for \ac{stft}-based \ac{bled} has been reported for the Casey Islands 2017 dataset in \citep{casey2017} using the PAMGuard Ishmael energy sum detector. These results are not directly comparable to this paper, since the detectors have additional parameters controlling false positives, such as time length and distance between detections and adaptive thresholding. These detections are measured on a per run-length basis, i.e., over the total run of the detector exceeding a threshold, instead of a per-sample basis. Therefore, this form of measurement may not accurately represent endpoint detection capabilities.

As seen in Figures \ref{fig:det_roc} and \ref{fig:det_calls}, the proposed \ac{se} detector outperforms the previous method \citep{mypaper}. Typical \ac{bled} with a fixed threshold performs very poorly due to varying noise conditions. \ac{bled} has repeatedly been shown to not be a reliable \ac{poi} detector \citep{jacques1, se_erbe_king}.

Interestingly, adding whitening to \ac{bled} results in similar performance to the method proposed by \citet{mypaper}. Spectral whitening and \ac{gmm} fitting shows incremental improvements to the k-means based \ac{se} detector. The final proposed detector outperforms the other detectors both in terms of endpoint accuracy (figure \ref{fig:det_roc}) and the number of calls versus noise detected (figure \ref{fig:det_calls}). This paper later discusses the implication of annotation inaccuracies in section \ref{sec:classification_discussion}. As such, the indicated results may not be entirely accurate, and could be better than indicated. From the false positives estimated in Table \ref{tab:numannotations}, all of the detectors likely detect 5-10\% more calls for realistic thresholds than indicated.

\clearpage
\section{Classification}
\label{sec:cls}
We combine the output of the \ac{se} detector with \ac{ws} features and utilise \ac{lda} to classify the detected calls. Our method shares many similarities with studies that also employ \ac{ws} without a \ac{nn}, but differs in allowing for variable window sizes produced by a detector \citep{ws_audio, ws_ecg}.

\subsection{Feature Extraction}
Detections are performed with a posterior signal probability threshold of 0.3, thereby corresponding to 91\% of vocalisations detected, with a significant number of unlabeled noise detections. Refer to Table \ref{tab:numdetections} for the number of detected calls.

After detection with with a whitened \ac{se} \ac{gmm} detector, we threshold the posterior probability given by Equation (\ref{eqn:entropyposterior}). We ensure that an entire vocalisation is captured by applying a rolling maximum filter over Equation (\ref{eqn:entropyposterior}), thereby extending the boundaries of the detections with a length of $G$. We set the filter length to 6 seconds, thus extending detections by 3 seconds on either side. Note that boundary extension may add additional calls to the detected vocalisations compared to the results from section \ref{sec:spectral_entropy}.

Labels are obtained by assigning each extended detection to an annotation that which occurs in the same timeframe, with the noise class being assigned if there are no corresponding annotations. Annotations have their endpoints shrunk by 5\% to account for endpoint inaccuracies. Detections having multiple annotation assignments which do not belong to the same class are discarded in this study. Noise detections may erroneously overlap with annotations, which could negatively impact the classification results. This is not addressed in this paper.

We dispose of the time dimension across scattering features for each detection by taking the maximum over the consecutive detected indices given by the set $\mathcal{M}$ and normalising power:
\begin{gather}
    \label{eqn:norm_rms}
    \bar{P} = \frac{1}{\left| \mathcal{M} \right|}\sum_{m \in \mathcal{M}} \sum_{\lambda_1 \in \Lambda_1} \left(\frac{\mathcal{S}_1[m, \lambda_1]}{\sqrt{\lambda_1}}\right)^2; \\
    \label{eqn:scatf1}
    f_1[\lambda_1] = \frac{1}{\bar{P}}\underset{{m \in \mathcal{M}}}{\text{max}}\mathcal{S}_1[m, \lambda_1]; \\
    \label{eqn:scatf2}
    f_2[\lambda_1, \lambda_2] = \frac{1}{\bar{P}}\underset{{m \in \mathcal{M}}}{\text{max}} \mathcal{S}_2[m, \lambda_1, \lambda_2].    
\end{gather}
Equation (\ref{eqn:norm_rms}) is the average power over detection indices which is calculated using equation (\ref{eqn:bled_power}).

\Ac{cls} features are used, as detailed by \citet{ws_audio}. First-level \ac{cls} features are directly comparable to \ac{mfcc}, with the addition of the the second level scattering features. The transform is applied to equations (\ref{eqn:scatf1}) and (\ref{eqn:scatf2}) as follows:
\begin{gather}
    F_1[\lambda_1] = \text{DCT}_{\lambda_1}\left(\log f_1[\lambda_1]\right); \\
    F_2[\lambda_1, \lambda_2] = \text{DCT}_{\lambda_2}\left(\log f_2[\lambda_1, \lambda_2]\right),    
\end{gather}
where $\text{DCT}_{x}$ denotes the discreet cosine transform over the variable $x$. The feature vector $v$ is constructed using using $F_1$ and $F_2$ over all significant scattering paths. This dataset is normalised to zero-mean and unit variance.

% The feature vector $v$ is constructed using
% \begin{equation}
%     v = \begin{pmatrix}
%         F_1[\lambda_{11}] \\
%         F_1[\lambda_{12}] \\
%         \vdots \\
%         F_2[\lambda_{11}, \lambda_{21}] \\
%         \vdots \\
%         F_2[\lambda_{1A}, \lambda_{2B}] \\
%         L
%     \end{pmatrix}, \ \lambda_{1a} \in \Lambda_1, \ \lambda_{1b} \in \Lambda_2.
% \end{equation}
% Note that for each $(\lambda_1, \lambda_2)$, the number of coefficients of $\lambda_{2}$ is a function of $\lambda_{1}$ as per the restriction given by equation (\ref{eqn:restrictlambda2}). Finally, all the features are normalised over the entire dataset by subtracting the means of each feature and dividing by each feature's standard deviation.

\subsection{Linear Disciminant Analysis}
We train a multiple ``one versus the rest" \ac{lda} classifiers using a balanced covariance estimate. There are 2 classes (vocalisation and noise) indexed by $i \in \{1, 2\}$, each with $K_i$ training observations. Each model considers only a single vocalisation class and regards all the other classes as noise.

Due to the similarity between calls of the \textit{Bm\_Ant} classes, models for these classes should not be penalised for misclassifying other \textit{Bm\_Ant} call types. As such, \textit{Bm\_Ant} models have all similar classes removed from their training and test sets. Many fin whale calls do not reach the classification stage of this paper due to their presence in the chorus sections of the audio and overlap. As such, we combine the fin whale vocalisation classes (\textit{Bp\_20Hz}, \textit{Bp\_20Plus}) into a single class (\textit{Bp}) due to their similarity.

We utilise regularised \ac{lda} with prior class probabilities $p(\mathcal{C})$. \ac{lda} assumes that each class is normally distributed, with a single class covariance matrix $\Sigma_\gamma = (1-\gamma)\hat{\Sigma} + \gamma \text{diag}\left(\hat{\Sigma}\right)$, where $\hat{\Sigma}$ is the sample covariance estimate shared between the classes and $\gamma$ is the covariance regularisation hyperparameter. Given an observation $x$ as a column vector, the log posterior probability of $x$ belonging to class $\mathcal{C}_i$ is
\begin{equation}
    \log p\left(\mathcal{C} = \mathcal{C}_i \middle| x\right) = -\frac{1}{2}\Sigma_\gamma^{-1}\hat{\mu}_i x -\frac{1}{2} \hat{\mu}_i^T\Sigma_\gamma^{-1}\hat{\mu}_i + \log p(\mathcal{C}_i) + \beta,
\end{equation}
where $\beta$ is a normalising constant that may be neglected to assign scores to each class.

To combat class imbalance, we use the following estimates:
\begin{gather}
    \label{eqn:lda_median}
    \hat{\mu}_i = \frac{1}{K_i}\sum_{n \in \mathcal{N}_i} x_n ;\\
    \hat{\Sigma}_i = \frac{1}{K_i}\sum_{n \in \mathcal{N}_i} \left(x - \hat{\mu}_i\right)^T\left(x - \hat{\mu}_i\right);\\
    \label{eqn:lda_sigma}
    \hat{\Sigma} = \frac{1}{2}\left(\hat{\Sigma}_1 + \hat{\Sigma}_2 \right),
\end{gather}
where $\mathcal{N}_i$ is the set containing the indices of $x_n$ belonging to class $\mathcal{C}_i$. Equation (\ref{eqn:lda_sigma}) approximates the covariance by assuming equal class weights, although this is typically weighted by the observation proportion in other applications. Equal weighting in shared covariance estimation ensures that all classes are able to significantly contribute to the covariance estimate, which combats the high proportion of noise samples in this study.

The prior probabilities $p(\mathcal{C})$ are typically empirically estimated from the training data. In this study, we employ the following model for prior probabilities to combat the high number of noise observations:
\begin{equation}
    p(\mathcal{C}) = \begin{cases}
        \alpha, \ \mathcal{C} = \mathcal{C}_\text{noise}\\
        1-\alpha,\ \mathcal{C} = \mathcal{C}_\text{vocalisation}
    \end{cases}.
\end{equation}
The hyperparameter $\alpha \in (0,1)$ sets the sensitivity of the classifier to noise detections.

\subsection{Results}
To evaluate our model, we split the training data into training and test sets proportional to the number of items belonging to that class. To demonstrate the practicality of our proposed technique, we perform experiments for different train-test splits with chi-square feature selection. An 80\%-20\% train-test split without feature selection is also used as a baseline. Results are shown in Table \ref{tab:clsres}. Feature selection is performed with a chi-square test of independence (noise versus signal classes). We select features with a statistical significance greater than 0.05, by binning each feature into 10 uniform bins. Feature selection is performed using only the training data.
% \begin{enumerate}
%     \item Training data is composed of 80\% of the noise and signal classes. This experiment illustrates the effectiveness of the the classifier with a full training set. Refer to Table \ref{tab:res80} for the results.
%     \item In practise, large annotated \ac{pam} datasets are not readily available. The second experiment utilises 20\% of the noise and signal classes for training the models. This experiment illustrates the effectiveness of the proposed technique for small datasets, allowing it to be more useful for practical applications.
%     \item Each class, especially the \textit{Bm\_Ant} classes, are bandlimited and may only require specific $\mathcal{S}_2$ features for classification. As such, it may be useful to perform feature selection for each class. For this experiment, we perform a chi-square test of independence and select only features with a statistical significance greater than 0.05, by binning each feature into 10 uniform bins. Feature reduction is only performed using the training data. As a baseline, we again use an 80\%-split for the training data.
%     \item To compare feature selection with the full feature set with little data, we employ another 20\% split with chi-square feature selection.
%     \item Finally, to show that adequate results may still be obtained with very little data, we use a 5\% split combined with chi-square feature selection. 
%     \item We evaluate the classes as is done in the previous experiment, with the 5\% of the lowest SNR calls are removed. SNR is determined by evaluating the power from the detector-provided window and comparing it to the noise power from the noise estimate. This shows that low-SNR and erroneous/partial detections have a significant impact on classification accuracy. Most fin whale calls are discarded due to low SNR, making it impractical to evaluate under the 5/95\% train-test split. As such, this experiment only evaluates blue whale calls.
% \end{enumerate}

Classification accuracy is measured using the observations not included in the test set. Unless specified otherwise, training is repeated 20 times and the results presented are the average accuracy and its standard deviation. We additionally perform a coarse grid search to optimise the hyperparameters $\alpha$ and $\gamma$ and present the results for the best hyperparameters. Hyperparameters are optimised by maximising the cost function $A_\text{Vocalisation} + 2A_\text{Noise}$, with $A$ the accuracy of the test set, thereby enforcing a stricter policy on false detections.

% \begin{table}[h!]
% \caption{Number of labeled detections for all vocalisations of interest. The detections are produced by the whitened \ac{gmm} \ac{se} detector. A significant number of detections are discarded in the training data due to overlapping.}
% \label{tab:numdetections}
% \begin{ruledtabular}
% \renewcommand{\arraystretch}{0.8}
% \begin{tabular}{lc} 
% \textbf{Vocalisation}      & \textbf{Number of detections}  \\ \hline
% Bm\_Ant\_A & 1514              \\
% Bm\_Ant\_B & 474              \\
% Bm\_Ant\_Z & 91               \\
% Bm\_D      & 328               \\
% Bp & 50                \\
% Noise      & 25638          \\
% Overlapping & 150 \footnotemark[1]      \\ 
% \end{tabular}
% \end{ruledtabular}
% \footnotetext[1]{Overlapping detections contain at least 2 overlapping annotations of different classes. This implies that more than 300 vocalisations are discarded as a result of overlap, thereby drastically reducing training data for the \textit{Bm\_D} and \textit{Bp\_20Plus} classes, which have a higher tendency for overlap.}
% \end{table}

% \begin{table}[]
% \caption{Average model classification accuracies for 80\% training samples.}
% \label{tab:res80}
% \begin{ruledtabular}  
% \begin{tabular}{lccccc}
% \textbf{Vocalisation} & \textbf{Vocalisation Accuracy (\%)} & \textbf{Noise Accuracy (\%)} & $\alpha$ & $\gamma$  \\ \hline  
% Bm\_Ant\_A & 82.2 ± 1.5  & 88.6 ± 1.2 & 0.66 & 0.71 \\
% Bm\_Ant\_B & 83.3 ± 6.2  & 92.3 ± 1.1 & 0.54 & 0.43 \\
% Bm\_Ant\_Z & 92.5 ± 10.6 & 98.2 ± 0.1 & 0.05 & 0.43 \\
% Bm\_D      & 74.1 ± 6.6  & 90.3 ± 2.0 & 0.66 & 0.86 \\
% Bp         & 96.0 ± 14.6 & 99.0 ± 0.1 & 0.17 & 0.57\\                   
% \end{tabular}
% \end{ruledtabular}  
% \end{table}

% \begin{table}[]
% \caption{Average model classification accuracies for 20\% training samples.}
% \label{tab:res20}
% \begin{ruledtabular}  
% \begin{tabular}{lccccc}
% \textbf{Vocalisation} & \textbf{Vocalisation Accuracy (\%)} & \textbf{Noise Accuracy (\%)} & $\alpha$ & $\gamma$  \\ \hline  
% Bm\_Ant\_A & 77.0 ± 7.1  & 88.2 ± 2.8 & 0.54 & 0.71 \\
% Bm\_Ant\_B & 67.6 ± 10.6 & 94.1 ± 0.6 & 0.41 & 0.86 \\
% Bm\_Ant\_Z & 75.8 ± 10.5 & 99.2 ± 0.0 & 0.05 & 0.71 \\
% Bm\_D      & 67.8 ± 5.0  & 90.3 ± 0.7 & 0.17 & 0.86 \\
% Bp         & 89.9 ± 4.0  & 93.3 ± 0.0 & 0.29 & 1.00 \\                    
% \end{tabular}
% \end{ruledtabular}  
% \end{table}

\begin{table}[]
\caption{Average model classification accuracies for varying train-test splits with chi-squared feature selection. An 80\% training split without chi-square feature selection is included as a baseline. 20 trials are performed, whereas 5\% splits use 100 trials. The median accuracy is displayed.}
\label{tab:clsres}
% \begin{ruledtabular}  
\renewcommand{\arraystretch}{0.8}
\begin{tabular}{lcccc}
\textbf{Vocalisation} & \textbf{Vocalisation Accuracy (\%)} & \textbf{Noise Accuracy (\%)} & $\alpha$ & $\gamma$  \\ \hline  
\multicolumn{5}{c}{80\% Train (No Feature Selection)} \\ \hline 
Bm\_Ant\_A & 82.2 ± 1.5  & 88.6 ± 1.2 & 0.66 & 0.71 \\
Bm\_Ant\_B & 83.3 ± 6.2  & 92.3 ± 1.1 & 0.54 & 0.43 \\
Bm\_Ant\_Z & 92.5 ± 10.6 & 98.2 ± 0.1 & 0.05 & 0.43 \\
Bm\_D      & 74.1 ± 6.6  & 90.3 ± 2.0 & 0.66 & 0.86 \\
Bp         & 96.0 ± 14.6 & 99.0 ± 0.1 & 0.17 & 0.57\\  \hline

\multicolumn{5}{c}{80\% Train (With Feature Selection)} \\ \hline 
Bm\_Ant\_A & 80.8 ± 1.5  & 89.9 ± 1.6 & 0.66 & 0.29 \\
Bm\_Ant\_B & 84.5 ± 6.1  & 91.6 ± 3.0 & 0.54 & 0.43 \\
Bm\_Ant\_Z & 96.4 ± 10.5 & 98.0 ± 0.5 & 0.41 & 0.43 \\
Bm\_D      & 73.6 ± 5.3  & 90.7 ± 4.0 & 0.66 & 0.71 \\
Bp         & 97.5 ± 9.2  & 98.0 ± 0.3 & 0.54 & 0.57 \\ \hline     
\multicolumn{5}{c}{20\% Train (With Feature Selection)} \\ \hline 
Bm\_Ant\_A & 80.3 ± 4.9  & 89.0 ± 8.8 & 0.66 & 0.43 \\
Bm\_Ant\_B & 79.1 ± 15.2 & 92.1 ± 4.3 & 0.54 & 0.57 \\
Bm\_Ant\_Z & 91.8 ± 12.4 & 97.9 ± 0.4 & 0.17 & 0.29 \\
Bm\_D      & 71.4 ± 12.9 & 90.6 ± 4.7 & 0.66 & 0.86 \\
Bp         & 92.5 ± 13.8 & 97.9 ± 0.2 & 0.41 & 0.71\\ \hline  
\multicolumn{5}{c}{5\% Train (With Feature Selection)} \\ \hline 
Bm\_Ant\_A & 73.5 ± 6.1  & 89.6 ± 8.5  & 0.66 & 0.57 \\
Bm\_Ant\_B & 64.7 ± 19.5 & 93.2 ± 12.3 & 0.54 & 0.71 \\
Bm\_Ant\_Z & 73.5 ± 18.7 & 95.7 ± 0.5  & 0.29 & 1.00 \\
Bm\_D      & 64.4 ± 18.5 & 90.7 ± 8.8  & 0.41 & 0.86 \\
Bp         & 81.2 ± 19.9 & 95.2 ± 12.7 & 0.54 & 0.86 \\  \hline

\multicolumn{5}{c}{5\% Train (With Feature Selection), Lowest 5\% SNR Calls Removed} \\ \hline 
Bm\_D  & 74.2 ± 15.7 & 89.1 ± 6.1  & 0.05 & 0.86 \\

\end{tabular}
% \end{ruledtabular}  
\end{table}

% \begin{table}[]
% \caption{Average model classification accuracies for 20\% training samples with chi-square feature selection.}
% \label{tab:res20chi2}
% \begin{ruledtabular} 
% \begin{tabular}{lccccc}
% \textbf{Vocalisation} & \textbf{Vocalisation Accuracy (\%)} & \textbf{Noise Accuracy (\%)} & $\alpha$ & $\gamma$  \\ \hline  
% Bm\_Ant\_A & 80.3 ± 4.9  & 89.0 ± 8.8 & 0.66 & 0.43 \\
% Bm\_Ant\_B & 79.1 ± 15.2 & 92.1 ± 4.3 & 0.54 & 0.57 \\
% Bm\_Ant\_Z & 91.8 ± 12.4 & 97.9 ± 0.4 & 0.17 & 0.29 \\
% Bm\_D      & 71.4 ± 12.9 & 90.6 ± 4.7 & 0.66 & 0.86 \\
% Bp         & 92.5 ± 13.8 & 97.9 ± 0.2 & 0.41 & 0.71\\                    
% \end{tabular}
% \end{ruledtabular}  
% \end{table}

% \begin{table}[]
% \caption{Average model classification accuracies for 5\% training samples with chi-square feature selection. This experiment uses 100 trials due to the high variance of the accuracy measures.}
% \label{tab:res5chi2}
% \begin{ruledtabular}  
% \begin{tabular}{lccccc}
% \textbf{Vocalisation} & \textbf{Vocalisation Accuracy (\%)} & \textbf{Noise Accuracy (\%)} & $\alpha$ & $\gamma$  \\ \hline  
% Bm\_Ant\_A & 73.5 ± 6.1  & 89.6 ± 8.5  & 0.66 & 0.57 \\
% Bm\_Ant\_B & 64.7 ± 19.5 & 93.2 ± 12.3 & 0.54 & 0.71 \\
% Bm\_Ant\_Z & 73.5 ± 18.7 & 95.7 ± 0.5  & 0.29 & 1.00 \\
% Bm\_D      & 64.4 ± 18.5 & 90.7 ± 8.8  & 0.41 & 0.86 \\
% Bp         & 81.2 ± 19.9 & 95.2 ± 12.7 & 0.54 & 0.86 \\                    
% \end{tabular}
% \end{ruledtabular}  
% \end{table}

% \begin{table}[]
% \caption{Average model classification accuracies for 5\% training samples with chi-square feature selection when the 5\% of the lowest SNR calls are removed from the detected windows provided to the classifier. This experiment uses 100 trials due to the high variance of the accuracy measures.}
% \label{tab:res5chi2discard}
% \begin{ruledtabular}  
% \begin{tabular}{lccccc}
% \textbf{Vocalisation} & \textbf{Vocalisation Accuracy (\%)} & \textbf{Noise Accuracy (\%)} & $\alpha$ & $\gamma$  \\ \hline  
% Bm\_Ant\_A & 73.4 ± 11.0 & 89.5 ± 9.4  & 0.66 & 0.71 \\
% Bm\_Ant\_B & 66.7 ± 23.4 & 93.0 ± 10.5 & 0.54 & 0.86 \\
% Bm\_Ant\_Z & 75.4 ± 18.3 & 95.2 ± 0.3  & 0.05 & 1.00 \\
% Bm\_D  & 74.2 ± 15.7 & 89.1 ± 6.1  & 0.05 & 0.86 \\                      
% \end{tabular}
% \end{ruledtabular}  
% \end{table}

\subsection{Discussion}
\label{sec:classification_discussion}
High noise rejection accuracies can be observed for both full and small training sets. As discussed in section \ref{sec:data}, the dataset contains a significant portion of unlabeled and mislabeled vocalisations. There are many vocalisations labelled with very low \ac{snr}, thereby not allowing a certain positive identification. Additionally, many low frequency blue whale calls (\textit{Bm\_Ant}) are unlabeled, and therefore incorrectly labeled as noise. As such, the models often detect calls which are not annotated, thereby penalising the noise rejection capabilities. Erroneous annotation assignments as a result of boundary extension and inaccurate annotation bounding boxes are also suspected to negatively contribute the results. The high variance compared to the reported median accuracy reflects on the quality of dataset annotations.

Using Table \ref{tab:numannotations}, we estimate that the true accuracies of the models' vocalisation classification capabilities may be between 1-17\% higher than indicated\footnote{This range is determined by considering that if our models correctly classified all erroneous annotations as noise, then their true accuracy is scaled by between $\frac{1}{1-\text{LB}}$ and $\frac{1}{1-\text{UB}}$, where LB and UB are the lower and upper bounds of erroneous annotation proportions. This calculation stems from the fact that the number of annotations are reduced, weheras the correct number vocalisation classifications remain the same. This range will change significantly based on whether the model classifies erroneous annotations as a vocalisation, which cannot be determined unless similar methods to \citet{casey2019} are employed . The estimation does not account for the \ac{se} detector's capability to rule out false positives.}, depending on the class. As discussed in section \ref{sec:detector_results}, adjusting for the noise accuracy of the models is out of the scope of this paper, since it must also account for the \ac{se} detector's capability of removing false postives. Our estimated proportion of false positives correspond well with the discrepency in model accuracies per class: a higher proportion of estimated false positive annotations yields a clear lower classification performance compared to other models.

The use of feature selection greatly increases performance for smaller datasets, as the accuracies for a 20\% training size with feature selection is nearly identical to that of the 80\% training size without feature selection. Feature selection is therefore crucial when attempting to use smaller datasets.

A lower accuracy for \textit{Bm\_D} is to be expected, since frequency-modulated downsweeps require temporal information and adjacent frequency information to be extracted. Furthermore, noise impulse sources may also resemble downsweeps when temporal information is discarded, as is performed in this study. There are especially many \textit{Bm\_D} annotations in the dataset with a very low \ac{snr}, which are most likely noise patterns that happen to resemble a downsweep. This is suspected to have a large impact on \textit{Bm\_D} noise rejection accuracy. This study may also be extended to use higher order wavelet scattering features, such as deeper levels and joint-\ac{tf} scattering, which results in classification that is more reminiscent of \ac{cnn} operating on spectrogram \citep{ws_joint_tf}. Joint-\ac{tf} scattering will greatly improve the \textit{Bm\_D} noise rejection.

The use of the annotations as the ``ground truth" is questionable and is well documented and discussed by Miller et. al. \citep{casey2019}. Miller utilises the same dataset that is used in this study to train a \textit{Bm\_D} classifier, and notes that annotations are significantly inaccurate across the dataset. They use an additional dataset for model evaluation, gathered in 2019 at the same location. Miller notes that more than 15\% of the analyst annotations in the evaluation dataset are false positives, and that there are nearly 30\% more calls than initially indicated by the analyst. After adjustment procedures, their measured model accuracy increased from 75\% (ground truth analyst notation) to 92\% (adjusted). This adjusted accuracy also corresponds well to the estimated adjustment required for our \textit{Bm\_D} model (8-17\%), assuming our model did not detect any erroneous annotations and that the entropy detector did not significantly alter the proportion of false positives. This effect is clearly seen when discarding low-SNR samples from training and testing, where the \textit{Bm\_D} accuracy increases by 10\% compared to when all samples are included.

The fin whale call (\textit{Bp}) indicates particularly effective classification, despite utilising only 5 training samples in the most extreme case. However, the accuracy varies dramatically over the randomised training trails, due to the small sample size, making the model sensitive to the choice of training data. Erroneous and low-\ac{snr} annotations are suspected to contribute to the large variation when the training samples are small ($<100$).

For the \textit{Bm\_Ant\_A}, \textit{Bm\_Ant\_B} and \textit{Bm\_D} classes, our noise rejection may not be sufficient, which currently results in $>2000$ false positives per class, a portion of which may be annotations missed by the analyst. Manual inspection of some audio segments indicate that there are large portions of unannotated calls from these classes. Further investigation is required to determine the practical acceptability of optimised noise rejection capabilities. The \textit{Bm\_Ant\_A} and \textit{Bm\_Ant\_B} models detect many unannotated vocalisations of all \textit{Bm\_Ant} types due to their similarity, thus explaining their lower noise accuracy compared to the \textit{Bm\_Ant\_Z} and \textit{Bp} models, which tend to have higher \ac{snr} compared to the other classes. Additionally, segments in which blue whale choruses are present, which are typically left unannotated, may negatively impact the \textit{Bm\_Ant} models' results.

Since we use independent models for classification, some calls will be detected by multiple models, especially for similar calls such as the \textit{Bm\_Ant} classes. It is not possible to implement multiclass classification in this paper due to large proportion of analyst misclassifications in the \textit{Bm\_Ant} classes. However, the posterior class probability may be used determine the correct class when multiple models have made a non-noise prediction in further studies.

Our disposal of the time dimension by using the maximum naturally lends our classification model being replaced with a \ac{hmm} to incorporate time-dependent relationships similar to that of \citet{mfcc_hmm_brydes} and \citet{dmd_hmm}. However, instead of classifying on a frame-by-frame basis such as in \citet{otherScattering, ws_audio, ws_ecg}, we utilise the scattering coefficients to first detect a variable-length segment of audio which may contain whale vocalisations, thereby reducing the amount of data in the noise class. This provides a distinct advantage over other methods, in that it already incorporates the extracted feature coefficients in detection prior to classification. As such, large portions of noise are discarded, the amount of which is tuned by the probability threshold of the proposed \ac{se} detector. This contrasts to \citet{casey2019}, which contains over 3 million noise class samples for 5000 vocalisation samples using the sliding window approach. This corresponds to a ratio of 1:600 for vocalisation to noise classes, whereas our method, which includes a detector, reduces this ratio to approximately 1:7 (all classes), at the cost of rejecting very low \ac{snr} vocalisations.

Despite the problems posed by the dataset used for evaluation, this study is the first to show the effectiveness of non-neural network models combined with scattering features. A detector is also utilised to remove noise segments, which assists in reducing class imbalance and reduces training time. The proposed method is particularly effective for a very small and unbalanced dataset, while allowing for a variety of calls to be detected.

\clearpage
\section{Conclusion}

In this study, we introduce \ac{ws} into the field of \ac{pam} for whale detection and classification and suggest further improvements of the traditional \ac{se} detector. We combine the detector with \ac{ws} and demonstrate the practicality of the proposed technique on a real, large dataset for blue and fin whale calls. We evaluate classification accuracy for many types of calls, demonstrating the applicability of the proposed technique on many whale species.

The classical \ac{se} detector is not equipped for adaptivity to a changing noise profile, which this study improves upon by applying a noise whitening procedure and fitting the \ac{se} output with a \ac{gmm}. The detector provides true probabilistic output, which accounts for varying noise and signal conditions, with the added benefit of interpretability for the user of this detector. The true-probabilistic extension and noise-whitening of the classical \ac{se} detector in this paper may be employed as a method of modifying and improving other existing detectors in terms of noise/signal adapativity and detector output interpretability.

The proposed system uses first level \ac{ws} coefficients, thereby allowing an efficient detection and feature extraction using a single \ac{tf} decomposition. Using a sensitive detector prior to classification allows for the elimination of large portions of noise, which assists in reducing the noise/signal class imbalance.

We use the \ac{se} detector's output is used to find \acp{poi} -- identifying windows of an arbitrary length to classify with \ac{cls} features with \ac{lda}. For this study, we ignore the time-dependent nature of the features in the windows by taking the maximum of the features over the detected time-frame, and reduce the number of features using chi-square feature selection. Feature selection proves to be effective in improving the performance for smaller datasets.

Further work includes incorporating time information into the classification stage, such as utilising models like \acp{hmm}. Frequency modulated calls, like the blue whale downsweep, is not well-suited to the proposed technique which may be addressed by modifying the proposed method with joint-\ac{tf} features, thereby more closely resembling the front-end of a 2D \ac{cnn} operating on a spectrogram. This modification will allow for relationships between frequency bins to also be taken into account, whereas the proposed \ac{ws} feature extraction method does not. More advanced and/or optimised feature selection methods may also be investigated to improve classification performance for smaller datasets.

The distinct advantage of the \ac{lda} classifier is the ability to use very small datasets consisting of 10s of samples per class to find other similar calls in an otherwise large audio database. Although not providing \ac{sota} classification accuracy, the proposed technique enables conservation researchers to search for similar sounds without requiring many training examples, allowing researchers to correctly classify more than 65\% of similar vocalisations detected by the \ac{se} detector when tuned for high noise rejection ($> 90\%$) with few training examples. The presented classification system will still yield some false positive detections, but will significantly reduce the effort required during annotation.

This paper provides the groundwork required to implement \ac{ws}-based classification without the need of \acp{nn} and large datasets, and provides a method to combine classical signal detectors and classifiers in an efficient way, thus bridging the gap between \ac{poi} detectors and classification models.

\clearpage

\begin{subappendices}

    \section{Conditions of Monotonicity of the GMM Posterior}
    \label{app:mono}
    Recall the normal distribution and its derivative:
    \begin{equation}
        \frac{d}{dx}\mathcal{N}(x|\mu,\sigma^2) = 2\frac{\mu -x}{\sigma^2} \mathcal{N}(x|\mu,\sigma^2).
    \end{equation}

    We want to find the regions in which the posterior is decreasing:
    \begin{equation}
        \label{eqn:toprove}
        \frac{d}{dx} p(C_s | x) < 0,
    \end{equation}
    where $x = H_w'$. $H_w' \in [0, \log(N)]$ follows trivially from equation (\ref{eqn:entropy}) and from the uniform distribution producing maximum entropy. $N$ is the number of points in the spectral distribution.

    The following restrictions are given:
    \begin{enumerate}
        \item Finite support on the interval: $ x_\text{max} > x > x_\text{min}.$
        \item The \ac{em} algorithm will converge to a solution where $x_\text{max} > \mu_n > \mu_s > 0.$
    \end{enumerate}

    Let $\varphi_i = p(C_i)\mathcal{N}(x|\mu_i,\sigma_i^2)$, implying $\frac{d}{dx}\varphi_i = \varphi_i' = k_i \mathcal{N}(x|\mu_i,\sigma_i^2)$, where $k_i = 2\frac{\mu_i - x}{\sigma_i^2}$.

    We calculate the derivative required by equation (\ref{eqn:toprove}):
    \begin{equation}
        \frac{d}{dx} p(C_s | x) = \frac{\varphi_n \varphi_s (k_s - k_n)}{(\varphi_n + \varphi_s)^2}.
    \end{equation}

    Since $\varphi_n$ and  $\varphi_n$ are always positive and nonzero, equation (\ref{eqn:toprove}) reduces to $k_s - k_n < 0$:
    \begin{equation}
        (\sigma_s^2 - \sigma_n^2) x - \sigma_s^2\mu_n +\sigma_n^2\mu_s < 0,
    \end{equation}
    which is linear with respect $x$, showing that the curvature only changes direction once when assuming all other variables are constants.

    Case 1: $\sigma_s^2 - \sigma_n^2 > 0$: $ x < \frac{\sigma_s^2\mu_n - \sigma_n^2\mu_s}{\sigma_s^2 - \sigma_n^2}.$

    Case 2: $\sigma_s^2 - \sigma_n^2 < 0$: $x > \frac{\sigma_s^2\mu_n - \sigma_n^2\mu_s}{\sigma_s^2 - \sigma_n^2}.$

    Case 3: $\sigma_s^2 = \sigma_n^2$: $\mu_s < \mu_n$, which is always true given the entropy assumptions.

    Case 3 is always true under the restrictions stated for this problem, implying that case 3 will always result in a monotonically decreasing function. To evaluate whether a given \ac{gmm} model with $x \in [x_{min}, x_{max}]$ satisfies monotonicity conditions, we substitute $x$ for $x_{max}$ and $x_{min}$ for cases 1 and 2 respectively.

    % \clearpage

    % \section{Summary of parameter values}

    % \begin{table}[h]
        % \caption{Summary of parameter values used in this study. Chosen values are indicated in boldface text, whereas non-bold parameters result from the chosen parameters.}
        % \label{tab:parameters}
        % \begin{ruledtabular}    
            % \renewcommand{\arraystretch}{0.8}% Tighter
            % \begin{tabular}{>{\centering\arraybackslash}l>{\centering\arraybackslash}l>{\centering\arraybackslash}l}  \small
                % \textbf{Parameter}      & \textbf{Description} & \textbf{Value} \\ \hline
                % $T$ & \ac{ws} invariance scale & \textbf{1.5 seconds} \\
                % $Q_1$ & Wavelets per octave in $\mathcal{S}_1$ & \textbf{16} \\
                % $Q_2$ & Wavelets per octave in $\mathcal{S}_2$ & \textbf{4} \\
                % $f_\text{min}$ & Minimum filterbank frequency in $\mathcal{S}_1$ & \textbf{15 Hz}\\
                % $f_\text{max}$ & Maximum filterbank frequency in $\mathcal{S}_1$ & \textbf{120 Hz}\\
                % $|\Lambda_1|$ & Resulting number of filters in $\mathcal{S}_1$ & 48 \\
                % $|\Lambda_2|$ & Resulting number of filters in $\mathcal{S}_2$ & 14 \\
                % $V$ & Wavelet scattering upsample factor & \textbf{2} \\
                % $r$ & Downsampling factor of $\mathcal{U}_1$ due to maximum bandwidth & 8 \\
                % $d$ & Downsampling factor of $\mathcal{S}_1$ & 5 \\
                % $f_{ss}$ & Resulting scattering transform sample frequency & 6.25 Hz \\
                % $B$ & Median filter length for noise coefficient suppression & \textbf{9} \\
                % $D$ & Median filter length for transient coefficient suppression  & \textbf{5 seconds} \\
                % $F$ & Average filter length for noise coefficient estimation  & \textbf{60 seconds} \\
                % $R$ & Median filter length for stabilizing entropy  & \textbf{1.5 seconds} \\
                % $G$ & Maximum filter length for boundary extension  & \textbf{6 seconds} \\
                % $p_t$ & Probability threshold for the whitened \ac{se} \ac{gmm} detector & \textbf{0.3} \\ 
                % $k$ & Dimensions in the classification feature vector $v$ & 521\\ 
                % \end{tabular}
                % \end{ruledtabular}
                % \end{table}

                % \clearpage
                % \section{Examples of misannotated vocalisations}
                % \label{app:ann}
                % \begin{figure}[h]
                    %      \centering
                    %      \begin{subfigure}[]{0.45\textwidth}
                        %          \centering
                        %          \includegraphics[width=\textwidth]{Bm_Ant-B_30.jpg}
                        %          \label{fig:BmA_ex}
                        %      \end{subfigure}
                        %      \hfill
                        %      \begin{subfigure}[]{0.45\textwidth}
                            %          \centering
                            %          \includegraphics[width=\textwidth]{Bm_Ant-B_37.jpg}
                            %          \label{fig:BmB_ex}
                            %      \end{subfigure}
                            %      \hfill
                            %      \begin{subfigure}[]{0.45\textwidth}
                                %          \centering
                                %          \includegraphics[width=\textwidth]{Bm_Ant-Z_27.jpg}
                                %          \label{fig:BmZ_ex}
                                %      \end{subfigure}
                                %      \hfill
                                %      \begin{subfigure}[]{0.45\textwidth}
                                    %          \centering
                                    %          \includegraphics[width=\textwidth]{Bm_D_33.jpg}
                                    %          \label{fig:BmD_ex}
                                    %      \end{subfigure}
                                    %         \caption{(a) \textit{Bm\_Ant\_A} labeled as \textit{Bm\_Ant\_B}, although the noise could be misinterpreted the ``tail" required for the B-type vocalisation. (b) Low SNR makes it difficult to distinguish between A-type and B-type calls. This example is labeled as \textit{Bm\_Ant\_B}. (c) A vocalisation more akin to the A-type may be easily mislabeled as \textit{Bm\_Ant\_Z} due to low-frequency noise. (d) The \textit{Bm\_D} class contains many low SNR annotations, some of which are likely patterns in the noise that have been misinterpreted as a vocalisation. Many of the misannotations in (a)-(d) become more apparent when using a constant Q transform such as the WST instead of a conventional \ac{stft}.}
                                    %         \label{fig:Bm_misannotations}
                                    % \end{figure}
\end{subappendices}

### 2024-05-05 dissertation/content/paper3/main.tex
added dissertation

\chapter{Separable Wavelet Scattering}

\articleinfo{M.W. Rademan, D.J.J Verself, J.A. Du Preez}{IEEE Signal Processing Letters}{15 May 2024}{doi}{Wavelet scattering is a widely used feature extraction method due to it efficacy in extracting invariant features, 
while retaining any lost high frequency information resulting from averaging to obtain the desired amount of invariance. 
By generalising a 1-dimensional scattering transform, we extend its definition to an arbitrary number of independent dimensions. 
We show that, in a modern deep learning setting, separable wavelet scattering performs similarly to its non-separable counterparts 
with the MNIST hand-written digit dataset. We additionally demonstrate state-of-the-art results for a subset of the MedMNIST3D datasets.}

\section{Introduction}
\Ac{ws} has seen widespread use in classification applications as a powerful feature extraction method. It is an especially effective method for small datasets, since the feature filters are reminiscent of a \ac{cnn} front-end, while requiring no learning. The resulting features are invariant across all averaged dimensions, and exhibit separable class subspaces, allowing linear classifiers to be used with high efficacy \citep{2dscattering}.

Various forms of \ac{ws} exist, with \ac{1d} scattering first proposed by Anden and Mallat \citep{1dscattering1, 1dscattering2}, which was later extended to 2 \citep{2dscattering} and 3 dimensions \citep{3dscattering, harmonicscattering}. Additional and more specialised forms for \ac{ws} include rotation-invariant scattering \citep{groupinvariantscattering} for the classification of textures and joint-\ac{tf} scattering \citep{jointtfscattering1, jointtfscattering2}. Joint-\ac{tf} scattering utilises a separable \ac{2d} filter that operates on the first level scattering scalogram, denoted by the operator $\mathcal{U}_1$, which is the only apparent usage of separable filters in the current scattering literature.

In a deep-learning setting, sensible initialisation of filters prior to learning can significantly improve performance and interpretability \citep{sincnet}. The greatest advantage of separable filters is their computation speed when computing convolutions directly, as is performed in \acp{cnn} \citep{separablecnn}. Additionally, separable filters reduce the number of parameters of a \ac{cnn} if the filters are configured to be learnable. Learnable filters can typically improve performance compared to their fixed-filters counterpart \citep{learnablewavelettransform}.

At first glance, a \ac{ws} decomposition seems to be prohibitively expensive, but fast algorithms are possible due to the limited bandwidth of filter output. Fast algorithms utilise downsampling to take advantage of the demodulation of band-limited signals that result from the convolution of analytic wavelets \citep{2dscattering, 3dscattering, 1dscattering1}. Path pruning is also used to ignore filter combinations which have negligible energy. These algorithms are the standard implementation on many platforms, such as the Kymatio python package \citep{kymatio} and MATLAB \citep{MATLAB}.

The Morlet wavelet is the most widely used wavelet filter for a scattering filter bank implementation. Since the Morlet does not have compact support, implementations vary when considering the bandwidth/time support of the Morlet. This affects how Morlet \ac{fir} filters are discretised and truncated, how scattering paths with negligible energy are pruned, and how filters overlap in the frequency domain.

In this work, we generalise \ac{1d} scattering to an arbitrary number of dimensions, which we refer to as the separable scattering transform. We propose flexible definitions for the Morlet bandwidth and filter overlap, which allows for an exact definition of near-optimal convolution computations with downsampling. Our implementation of separable scattering is \ac{gpu} accelerated, similar to the Kymatio implementations \citep{kymatio}.

We demonstrate using the MNIST handwritten digit dataset \citep{mnist} that separable wavelet filters still perform adequately in a \ac{nn} classification setting compared to the conventional 2D scattering filters. We demonstrate the arbitrarily scalable dimensionality of the separable scattering transform with three-dimensional medical datasets from MedMNIST3D \citep{medmnist}. Separable scattering coefficients show \ac{sota} results for some of the MedMNIST3D datasets when combined with a simple \ac{nn}, while performing comparably to \ac{cnn} benchmarks on datasets which do not indicate \ac{sota} results.

\section{Separable Morlet Filterbank}

% In this section, we consider the continuous Morlet and provide general definitions of bandwidth and time-support for multiple dimensions. We utilise the Gaussian window to construct $n$-dimensional Morlets of varying characteristics.

\subsection{Morlet}

We define a \ac{1d} zero-mean Gaussian as
\begin{equation}
    \theta_{\sigma_t}(t) = \frac{1}{\sqrt{2\pi\sigma_t^2}}e^{-\frac{1}{2}\left(\frac{t}{\sigma_t}\right)^2},
\end{equation}
which has a Fourier transform transform $\theta_{\sigma_t}(t) \xleftrightarrow{\mathcal{F}} \hat{\theta}_{\sigma_\omega}(\omega)$, where $\sigma_\omega = \frac{1}{\sigma_t}$.

For an arbitrary bandwidth definition, we define the bandwidth-to-$\sigma$ ratio $\beta \in \mathbb{R}^+$, such that the one-sided bandwidth of a zero-mean Gaussian is $\beta \sigma_\omega$. For reference, Kymatio's implementation of 2D scattering indirectly defines $\beta \approx 2.5$ \citep{kymatio}.

% The $-3$ dB point of a Gaussian \ac{lpf} occurs at $\beta \approx 0.833$, but since the Gaussian \ac{lpf} does not have a very steep cutoff, values of $\beta \in [1, 3]$ may be appropriate to combat aliasing.

A \ac{1d} Morlet $\uppsi$ has energy concentrated around 1 rad/s, with most of its energy contained in the interval $\omega \in [1 - \frac{1}{Q}, 1 + \frac{1}{Q}]$. $Q \in \mathbb{R}^+$ is defined as the number of wavelets per octave. The mother wavelet is given as
\begin{equation}
    \uppsi(t) = \theta_{\sigma_t}(t)\left(e^{jt} - \theta_0 \right),
\end{equation}
where $\theta_0 = \frac{\theta_{\sigma_t}(-1)}{\theta_{\sigma_t}(0)}$ to ensure zero mean: $\hat{\uppsi}(0) = 0$.

For reasons that will become apparent in section \ref{sec:filterbank}, we defined a 1D-wavelet dilated by a factor $\lambda$ as 
\begin{equation}
\label{eqn:dilwav}
    \uppsi_\lambda(t) = \begin{cases}
        \uppsi(\lambda t), \ \lambda \neq 0 \\
        \phi_j(t), \ \lambda = 0
    \end{cases},
\end{equation}
where $\phi$ represents the 1D \ac{lpf} utilised in scattering computations for the current dimension ($j$).

A $n$-dimensional separable wavelet may be constructed similarly, with
\begin{equation}
    \uppsi_{\vect{\lambda} }(\mathbf{u}) = \prod_{i=0}^{n-1} \uppsi_{\lambda_i}(u_i),
\end{equation}
where $\mathbf{u} = \left(u_1, ..., u_n\right)^T$ is the vector containing the dimensions of interest.

A filterbank is constructed by dilating the mother wavelet with a set of dilation factors $\vect{\lambda} =(\lambda_1, ..., \lambda_n )^T \in \Lambda_1 \times ... \times \Lambda_n$, with $\Lambda_i$ the set of dilation factors for the variable $u_i$. The dilation factor $\lambda_i$ is also the centre frequency of the dilated wavelet in rad/s.

% Each dimension indexed by $i$ has its own defined \ac{lpf}.

% with with a dilated one-sided bandwidth of $\lambda_i \beta \sigma_{\omega_i}$.

\subsection{1D Filterbank Construction}
\label{sec:filterbank}
 Suppose a 1D Morlet has a centre frequency $\lambda_0$. The following filter is placed at $\lambda_0 2^{\frac{1}{Q}}$. To define the amount of overlap between filters, it is useful express the corresponding filter's frequency \ac{std} $\sigma_\omega$ in terms of the distance between filters in the frequency domain. If the wavelet at $\lambda_0$ requires a decay equivalent to $\alpha \sigma_\omega$ \ac{std}s at the next wavelet at $\lambda_0 2^{\frac{1}{Q}}$. It follows that $\sigma_\omega = \frac{1}{\sigma_t} = \frac{1}{\alpha}\left( 2^\frac{1}{Q} - 1 \right)$.

We refer to $\alpha$ as the clearance factor. More overlap will result in more redundancy in the frequency representation, but also shorter filter impulse responses.

% It is possible to set the clearance factor equal to the defined Gaussian bandwidth: $\alpha = \beta$. However, different values of $\alpha$ and $\beta$ allow for the separate definition of filter clearance and bandwidth, allowing the application to tune downsampling and filter time support separately.

The provided definitions of filter clearance ($\alpha$) and bandwidth ($\beta$) may also be extended to non-separable wavelets, such as the rotational construction employed in \citep{2dscattering}. However, a conversion is required to find the bandwidth in each axis for a specific wavelet for rotationally constructed filterbanks, in order to employ the downsampling strategies proposed in this paper.

% A \ac{1d} wavelet filterbank is constructed with a set of filters $\mathbb{F} := \left\{\uppsi_\lambda \middle| \lambda \in \Lambda\right\}$ and a \ac{lpf} $\phi$ with the following specified parameters:
% \begin{enumerate}
%     \item $Q$ - wavelets per octave.
%     \item  $\omega_0$ - starting frequency in rad/s (optional).
%     \item  $d$ - invariance scale in samples, equivalent to the decimation factor applied to scattering coefficients.
% \end{enumerate}

All filterbank descriptions in this paper utilise normalised sampling frequency, i.e., the sample frequency is $f_s = 1$.

For some audio applications, for example speech recognition, low-frequency content is negligible, motivating filterbank construction starting at a specified frequency $\omega_0$. Otherwise, the entire frequency domain may be covered by setting $\omega_0$ to start at a position specified by $\phi$ and $\alpha$.

The \ac{lpf} $\phi$ is constant across all layers in a \ac{1d} scattering transform, and is chosen as $\phi(t) = \theta_{\sigma_{\phi, t}}(t)$, where $\sigma_{\phi, t} = \dfrac{d}{ \pi \beta}$.

The scattering transform requires that the time support of all filters do not exceed the time support of $\phi$, ensuring no filters contain time information exceeding the invariance scale $d$. $d$ is equivalently defined as the total downsample factor of the scattering tranform, and is unique for each dimension. Equivalently, the frequency \ac{std} of the filters may not exceed the \ac{std} of $\phi(\omega)$. For brevity, we denote the frequency \ac{std}s as $\sigma_{\lambda, \omega}$ and $\sigma_{\phi, \omega}$ for the Morlets and the \ac{lpf} respectively.

 Since the maximum time support (minimum bandwidth) is a function of $d$, all dilated wavelets have their Gaussian envelopes restricted to a maximum time support of $\sigma_{\phi, t}$. This requires linearly spaced filters until the dilated bandwidth is larger than the \ac{lpf} bandwidth. To construct a set of positive dilation factors $\lambda \in \Lambda^+$, refer to algorithm \ref{alg:fb}.

\begin{algorithm}    
\caption{$\Lambda^+$ construction.}\label{alg:fb}
\begin{algorithmic}
    \State $\sigma_{\phi, \omega} \gets \dfrac{\pi\beta}{d}$
    \If{frequency limited}
        \Ensure{$\omega_0 \ge \dfrac{\pi\alpha}{d}$}
        \State $\lambda \gets \omega_0$
    \Else{}
        \State $\lambda \gets \dfrac{\pi\alpha}{d}$
    \EndIf
    \State $\Lambda^+ \gets \varnothing$
    \State $\sigma_\omega \gets \dfrac{1}{\alpha}\left( 2^\frac{1}{Q} - 1 \right)$
    \State $\sigma_{\lambda, \omega} \gets \lambda\sigma_\omega$

    \While{$\sigma_{\phi, \omega} > \sigma_{\lambda, \omega}$ and $\lambda < \pi$} 
        \State $\Lambda^+ \gets \Lambda^+ \cup \{\lambda\}$
        \State $\lambda \gets \lambda + \alpha \sigma_{\phi, \omega}$ 
        \State $\sigma_{\lambda, \omega} \gets \lambda\sigma_\omega$
    \EndWhile

    \While{$\lambda < \pi$}
        \State $\Lambda^+ \gets \Lambda^+ \cup \{\lambda\}$
        \State $\lambda \gets \lambda 2^{\frac{1}{Q}}$
        \State $\sigma_{\lambda, \omega} \gets \lambda\sigma_\omega$
    \EndWhile
\end{algorithmic}
\end{algorithm}

Only positive $\lambda$'s have been defined thus far, which provides inadequate coverage of the frequency domain in multiple dimensions. For real input signals, it is only necessary to cover half of one of the dimensions (only positive $\lambda$'s), whereas full coverage (both negative and positive $\lambda$'s) is required for additional dimensions. On-axis coverage is also required, in which each $\uppsi_\lambda$ must be multiplied with a Gaussian (zero-frequency wavelet), which the dilated wavelet definition in equation (\ref{eqn:dilwav}) defines as $\lambda=0$. A similar construction procedure is followed in \citep{jointtfscattering2}.

Given $m$ 1D filterbanks, with $m \ge 2$, each having Morlet filters with a positive set of lambdas $\Lambda_i^+$ and invariance scales $\vect{d} \in \mathbb{N}^m$, with $i = 1, ..., m$ indexing the dimension, we construct the $m$-dimensional filterbank with 
\begin{equation}
    \mathbb{F} = \left\{ \uppsi_{\vect{\lambda} }(\mathbf{u}) \ \middle| \ \vect{\lambda} \in \mathbb{L} \setminus \{\vect{0}\} \right\},
\end{equation}
where $\vect{u}$ is the $m$-dimensional spatial and/or time variable in which the each 1D filterbank is defined.  By definition, 
\begin{gather}
    \mathbb{L} = (\Lambda_1^+ \!\cup\! \{0\})\!\times\! (\Lambda_2^+ \!\cup\! \Lambda_2^- \!\cup \!\{0\}) \!\times\! ... \!\times\! (\Lambda_m^+ \!\cup\! \Lambda_m^- \!\cup\! \{0\}); \\
    \Lambda_i^- = \left\{-\lambda \ \middle| \ \lambda \in \Lambda_i^+\right\},
\end{gather}
where $\cup$ indicates the set union operator and $\times$ the Cartesian product. For $m=1$, the provided definitions result in a conventional \ac{1d} scattering transform \citep{1dscattering1}.

The $m$-dimensional \ac{lpf} is defined as 
\begin{equation}
    \phi(\vect{u}) = \uppsi_{\vect{0}}(\vect{u}).
\end{equation}

% The dilated wavelet definition provided by (\ref{eqn:dilwav}) therefore provides a concise set construction, with which to construct the \ac{lpf} and on-axis wavelets simultaneously.

% All on-axis filters will be highly correlated with their negative counterpart due to the symmetry of the multidimensional Fourier transform of real signals. For example, the 2D filter specified by $\lambda = [0, x]^T$ will be highly correlated with the filter at $[0, -x]^T$. As the invariance scale $d$ increases, the on-axis filters will become more narrow in the zero coordinate dimension, thereby becoming increasingly correlated.

% \subsection{Discretisation}

% Morlets do not have compact support, as such, they must be truncated for computation. The invariance scale $d$ specifies the maximum allowable time support of all filters. Since input signals must be at least of length $d$, filters can be precomputed using the padded signal length, while also guaranteeing a suitable level of truncation. The padding required is on the order of $d$ samples.

% Note that, due to their separability, all $\uppsi_{\vect{\lambda}} \in \mathbb{F}$ can be stored as $d$ one-dimensional filters, and not as $N_1 \times ... \times N_d$ tensors, thereby saving on the storage required for pre-computed filters. The savings on storage space allow implementations to compute the filters at multiple input sample frequencies, allowing for optimal downsampling strategies without incurring a significant storage cost.

\section{Separable Scattering Transform} \label{sec:wst}
% In this section, we describe the efficient computation of $m$-dimensional separable scattering transforms. The described downsampling strategy is utilised to a limited extent in Kymatio scattering implementations \citep{kymatio} and the original proposed fast algorithms \citep{2dscattering}, which is generalised further in this section.

\subsection{Transform}

The scattering transform requires 2 steps to provide scattering coefficients. The scalogram operator $\mathcal{U}_j$ iteratively filters a discrete signal $x[\vect{n}]$ for a given set of filters $\mathbb{F}$, which is then averaged by the \ac{lpf}. $\vect{n}$ represents a multidimensional index variable.
\begin{gather}
    \mathcal{U}_j x [\vect{n}, \vect{\lambda}_1, ..., \vect{\lambda}_j] = \left|\left(\mathcal{U}_{j-1} \ ...\  \mathcal{U}_1 x\right) * \uppsi_{\vect{\lambda}_j} \right|, \ \forall \ \uppsi_{\vect{\lambda}_j} \in \mathbb{F}; \\
     \mathcal{U}_1 x [\vect{n}, \vect{\lambda}_1] =  \left| x * \uppsi_{\vect{\lambda}_1} \right| , \ \uppsi_{\vect{\lambda}_1} \in \mathbb{F}.
\end{gather}

The scattering operator $\mathcal{S}_j$ provides the output coefficients at the $j$'th order of the scattering transform:
\begin{equation}
    \mathcal{S}_j x[\vect{n}, \vect{\lambda}_1, ..., \vect{\lambda}_j] = \mathcal{U}_j x * \phi.
\end{equation}
The \ac{lpf} $\phi$ remains constant throughout the transform. Note that the modulus/magnitude operator $|\cdot|$ demodulates the output of the filters, effectively extracting the Hilbert envelope from a band-limited signal \citep{waveletsandsubbandcoding}.

The $j$'th scattering order adds an additional axis of paths indexed by $\vect{\lambda}_j$. However, not all paths need to be evaluated, since some paths have smaller bandwidths, thereby requiring fewer filters to extract the information lost by averaging. In particular, we only evaluate paths in which the centre frequencies of all elements of the vector $\vect{\lambda}_{j}$ are smaller than their corresponding bandwidth of the previous path's filter $\uppsi_{\vect{\lambda}_{j-1}}$. Path pruning is therefore dependent on $\alpha$ and $\beta$.

\subsection{Downsampling Strategy}
Since each filter specified by $\vect{\lambda}$ has its own bandwidth, we can employ downsampling across all paths non-uniformly. However, some care is required to ensure the compounded downsampling steps across all paths result in a uniform sampling frequency of the output scattering coefficients.

% The invariance scale $d$ is equivalent to the total downsampling required to compute scattering coefficients. $d$ is typically chosen as a power of $2$, which allows for multiple decimation stages to also subsample by powers of $2$. This decimation scheme is illustrated particularly well by dyadic wavelet transforms.

% However, it is not required that $d$ is restricted to a power of $2$, but rather that compounded decimation steps result in an effective decimation of $d$. As such, $d$ is most effective when it has as many factors as possible. In this study, we will require that $d$ is an even number.

Many applications are insensitive to small changes in $d$. As such, we propose a strategy to find an optimal $d$ given a target and tolerance value. For some applications, choosing $d$ such that the downsampling factor is a power of 2 is the simplest solution to achieve optimality.

Without prior knowledge of the filterbank configurations, given a target invariance scale of $\bar{d}$ samples and a tolerance $\epsilon$, we can optimise $d \in \{\floor{(1 - \epsilon)\bar{d}}, \ceil{(1 + \epsilon)\bar{d}}\}$ such that it results in largest number of supported downsampling configurations.

A downsample factor $d$ which decomposes into a set of $n$ prime factors $\{p_1, ..., p_n\}$ with a corresponding multiset $\mathbb{M} = \left\{ m_1, ..., m_n \right\}$, where $m_i$ is the multiplicity of the prime $p_i$. We can find an optimal $d$ by maximising the sum $\sum\limits_{m \in M} m$.

% However, the proposed scheme for finding an optimal $d$ allows for more choices, thereby no restricting the choice of $d$ much, while also allowing for efficient dowmsampling.

Morlet filters in a 1D filterbank may be downsampled by a factor $d_{\psi_1}$ prior to low-pass filtering, and then downsampling again by a factor $d_{\phi_1}$ after low-pass filtering. As such, the compounded effect of downsampling restricts $d = d_{\psi_1} \cdot d_{\phi_1}$.

In the second order of scattering, the process is repeated with an additional pre-low-pass downsampling factor of $d_{\psi_2}$. The second level application of $\phi$ then downsamples by a factor $d_{\phi_2}$. To maintain a consistent output sampling frequency, it restricts $d = d_{\psi_2} \cdot d_{\psi_1} \cdot d_{\phi_2}$.

Continuing the downsampling chain, the $i$'th level of downsampling requires $d = d_{\psi_i} \cdot ... \cdot d_{\psi_1} \cdot d_{\phi_i}$. The output of each operation of $\mathcal{U}_i$ and $\mathcal{S}_i$ must be downsampled as much as possible in order to make subsequent operations faster. To ensure that the application of all the downsampling steps are efficient, we require $d$ to have as many prime factors as possible, including factor multiplicity, so that a larger variety of downsampling combinations may be supported.

% \begin{algorithm}    
% \caption{Finding an optimal $d$.}\label{alg:optt}
% \begin{algorithmic}
%     \State $d_1 \gets \floor*{\frac{f_s \bar{T}(1 - \epsilon)}{2 \beta}}$\\
%     \State $d_2 \gets \ceil*{\frac{f_s \bar{T}(1 + \epsilon)}{2 \beta}}$\\
%     \State $k_\text{max} \gets 1$\\
%     \For{$d_j \in \{d_1, d_1 + 1, \ ... \ , d_2 - 1, d_2\}$}
%         \State $M \gets \text{PrimeMultiplicity}(d_j)$
%         \State  $k \gets \sum\limits_{m \in M} m$\\
%         \If{$k > k_\text{max}$}
%             \State $k_\text{max} \gets k$
%             \State $d \gets d_j$
%         \EndIf
%     \EndFor
% \end{algorithmic}
% \end{algorithm}

Consider a 1D wavelet filterbank and a single wavelet filter $\uppsi_1[n]$, applied to a discrete-time signal $x$. The operations required to compute the scattering coefficients is then notated for simplicity as
\begin{gather}
    u_1[n] = \left|  x * \uppsi_1   \right|; \\
    s_1[n] = y_1 * \phi.
\end{gather}

The bandwidth of $u_1$ is the bandwidth of an arbitrary first filter $\uppsi_1$. This follows from the Hilbert envelope computed by the analytic wavelet filter $\uppsi_\lambda$ and the modulus $|\cdot |$. The bandwidth of $s_1$ is the bandwidth of $\phi$.

Critical downsampling of a wavelet $\uppsi$ with a bandwidth of $\sigma_\omega$ is achieved by a factor of $d_{\uppsi} = \floor*{\frac{ \pi }{\beta\sigma_\omega}}$. Critical downsampling of $s_1$, is achieved via a factor of $d$, by definition.

We can efficiently compute $s_1$ using compounded downsampling steps:
\begin{equation}
    \left(s_1\right)_{\downarrow d} = \left( \left(u_1\right)_{\downarrow d_1} * (\phi)_{\downarrow d_1}\right)_{\downarrow d_2},
\end{equation}
such that $d = d_1 \cdot d_2, \ d_1, d_2 \in \mathbb{N}^+$, with $d_1 | d$ and $d_1 \le  d_{\uppsi_1}$. In order to find $d_1$, we decrement $d_{\uppsi_1}$ until it divides $d$ evenly. Each scalogram $u_j[n]$ is not necessarily downsampled optimally, but has a downsampling factor which guarentees a consistent scattering coefficient output sample frequency.

% This strategy may counter intuitively result in faster computations for filterbanks with more filters, since a higher $Q$ implies a smaller bandwidth of each filter.

A second order of scattering with a filter path of $(\uppsi_1, \uppsi_2)$ is performed on the downsampled $u_1$:
\begin{gather}
    u_2 = \left|  \left(u_1\right)_{\downarrow d_1} * (\uppsi_2)_{\downarrow d_1}  \right|; \\
    \left(s_2\right)_{\downarrow d} = \left( \left(u_2\right)_{\downarrow d_3} * (\phi)_{\downarrow d_3}\right)_{\downarrow d_4},
\end{gather}
such that $d = d_1 \cdot d_3 \cdot d_4, \ d_1, d_3, d_4 \in \mathbb{N}^+, d_3 \le d_{\uppsi_2}, d_3 | \frac{d}{d_1}$. The proposed downsampling scheme can be extended to an arbitrary number of scattering orders.

% Since the filterbank configuration is known, all required combinations of downsampled filters and their downsampling factors can be precomputed.

\begin{table*}[t!]
\centering
\caption{MedMNIST classification results of compared to baseline \ac{nn} approaches (\citep{medmnist})} \label{tab:medmnist}
\begin{tabular}{|l|rr|rr|rr|rr|rr|rr|}
\hline
\multirow{2}{*}{Methods}                & \multicolumn{2}{c}{Organ ($c=11$)}                  & \multicolumn{2}{c}{Nodule ($c=2$)}                 & \multicolumn{2}{c}{Fracture ($c=3$)}               & \multicolumn{2}{c}{Adrenal ($c=2$)}                & \multicolumn{2}{c}{Vessel ($c=2$)}                 & \multicolumn{2}{c|}{Synapse ($c=2$)}                \\
                         & \multicolumn{1}{l}{AUC} & \multicolumn{1}{l}{ACC} & \multicolumn{1}{l}{AUC} & \multicolumn{1}{l}{ACC} & \multicolumn{1}{l}{AUC} & \multicolumn{1}{l}{ACC} & \multicolumn{1}{l}{AUC} & \multicolumn{1}{l}{ACC} & \multicolumn{1}{l}{AUC} & \multicolumn{1}{l}{ACC} & \multicolumn{1}{l}{AUC} & \multicolumn{1}{l|}{ACC} \\ \hline
ResNet-1810+2.5D        & 0.977                   & 0.788                   & 0.838                   & 0.835                   & 0.587                   & 0.451                   & 0.718                   & 0.772                   & 0.748                   & 0.846                   & 0.634                   & 0.696                   \\
ResNet-1810+3D          & 0.996                   & 0.907                   & 0.863                   & 0.844                   & 0.712                   & 0.508                   & 0.827                   & 0.721                   & 0.874                   & 0.877                   & 0.82                    & 0.745                   \\
ResNet-1810+ACS41       & 0.994                   & 0.900                     & 0.873                   & 0.847                   & 0.714                   & 0.497                   & 0.839                   & 0.754                   & 0.930                    & 0.928                   & 0.705                   & 0.722                   \\
ResNet-5010+2.5D        & 0.974                   & 0.769                   & 0.835                   & 0.848                   & 0.552                   & 0.397                   & 0.732                   & 0.763                   & 0.751                   & 0.877                   & 0.669                   & 0.735                   \\
ResNet-5010+3D          & 0.994                   & 0.883                   & 0.875                   & 0.847                   & 0.725                   & 0.494                   & 0.828                   & 0.745                   & 0.907                   & 0.918                   & \textbf{0.851}          & \textbf{0.795}          \\
ResNet-5010+ACS41       & 0.994                   & 0.889                   & 0.886                   & 0.841                   & \textbf{0.750}           & \textbf{0.517}          & 0.828                   & 0.758                   & 0.912                   & 0.858                   & 0.719                   & 0.709                   \\
auto-sklearn11          & 0.977                   & 0.814                   & \textbf{0.914}          & \textbf{0.874}          & 0.628                   & 0.453                   & 0.828                   & \textbf{0.802}          & 0.910                   & \textbf{0.915}          & 0.631                   & 0.730                    \\
AutoKeras12             & 0.979                   & 0.804                   & 0.844                   & 0.834                   & 0.642                   & 0.458                   & 0.804                   & 0.705                   & 0.773                   & 0.894                   & 0.538                   & 0.724                   \\ \hline
Separable WS + NN (Ours)  & \textbf{0.998}          & \textbf{0.941}          & 0.858                   & 0.797                   & 0.614                   & 0.458                   & \textbf{0.875}          & 0.792                   & \textbf{0.962}          & 0.895                   & 0.715                   & 0.525    \\ \hline              
\end{tabular}
\end{table*}

\subsection{Convolutions}
Optimal \ac{fft} convolutions can be achieved by performing downsampling in the frequency domain instead of the time domain. It is straightforward to verify that $|x * \uppsi|_{\downarrow r} = \left|(x * \uppsi)_{\downarrow r}\right|$, since the modulus is an element-wise operation, for some downsampling factor $r$. Given the signal and filter Fourier transforms $\hat{x}$ and $\hat{\uppsi}$, we then have
\begin{equation}
\label{eqn:circconv}
    x \otimes \uppsi [n] \xleftrightarrow{\mathcal{FFT}}  \hat{x} \cdot \hat{\uppsi} [k],
\end{equation}
where the $\otimes$ operator represents a circular convolution and $k$ is the frequency index.

Given that $r|N$, we can express (\ref{eqn:circconv}) when downsampled as a periodised summation \citep{waveletsandsubbandcoding} in the frequency domain
\begin{equation}
    (x \otimes \uppsi [n])_{\downarrow r} \xleftrightarrow{\mathcal{FFT}}  \frac{1}{r} \sum_{i=0}^{r-1} \hat{x} \cdot \hat{\uppsi} [k + iN/r], %https://citeseerx.ist.psu.edu/document?repid=rep1&type=pdf&doi=5216ea733e562541b33a7f97dab0de072b2e8827
\end{equation}
which can be efficiently implemented via shape manipulation of tensors in computational packages like MATLAB or PyTorch \citep{pytorch}. To compute valid convolutions, we must pad $x$ and $\uppsi$ to have a total length of $N = N_x + d + c$, where $c \in \mathbb{N}^+$ is a constant that ensures that $d | N$.

% Note that only $d$ additional samples are required instead of $2d$, since $d$ is, by definition, the time support of $\phi$ in number of samples. Padding is applied equally to the left and right of $x$. For further reduction of boundary effects, reflection padding is utilised. Kymatio's 2D scattering implementation utilises a similar padding scheme.

% Thus far, the techniques discussed may be equivalently implemented on many scattering transforms. Separable scattering benefits from reduced computations resulting from filter separability. In particular, given a $m$-dimensional signal $x[\vect{n}]$ indexed by $\vect{n} = (n_1, ..., n_m)^T$, with each dimension having downsample factors expressed as a vector $\vect{r} = (r_1, ..., r_m)^T$, we can chain downsampling for each dimension:
% \begin{equation}
%     \left(x * \uppsi_{\vect{\lambda}}[\vect{n}]\right )_{\downarrow \vect{r}} = \left( \left(x[\vect{\lambda}] * \uppsi_{\lambda_i}[n_i]\right)_{\downarrow r_1} ... * \uppsi_{\lambda_m}[n_m]\right)_{\downarrow r_m}.
% \end{equation}.

% We abuse notation to indicate that $\downarrow r_i$ is an operator that downsamples the $i$'th dimension, and $\downarrow \vect{r}$ downsamples all dimensions each with their correponding downsample factor in $\vect{r} \in \mathbb{N}^m$.

% For $m$-dimensional sequences of dimensional lengths $N_1, ..., N_m$, the forward and inverse \acp{fft} computation time is reduced from $2\sum_{i} T \log N_i$ (no frequency periodisation), with $T = \prod_i N_i$ to $\sum_{i} \frac{T}{K_i} \log \frac{N_i}{K_i} + P \log P$, where $K_1 = 1$, $K_i = \prod_{j < i} r_j$ and $P = \prod_i \frac{N_i}{r_i}$. However, the number of convolution multiplications in the frequency domain increases from $\prod_i N_i$ to $\sum_i \prod_j \frac{N_j}{K_i} \leq m \prod_i N_i$.

% In contrast, optimal non-separable scattering transforms can achieve a computational cost of $\sum_{i} N_i \log N_i + P \log P$, since a full forward \ac{fft} is required to compute a convolution with a non-separable $m$-dimensional filter.

% The greatest computational advantage is during direct convolution computations, as is present in \acp{cnn}. For an $m$-dimensional filter impulse response with filter lengths $L_1, ..., L_m$, the computational cost reduces from $P\prod_i L_i$ to $P \sum_i L_i$.

\section{Results}

\subsection{MNIST}

% We benchmark separable \ac{ws} against 2D scattering by repeating the MNIST hand-written digit dataset experiment in \citep{2dscattering}.

The MNIST dataset \citep{mnist} contains 60000 training and 10000 test samples. Unlike in \citep{2dscattering}, which decorrelates scattering coefficients with a discrete cosine transform prior to classification, we perform classification on the scattering coefficients directly. Kymatio \citep{kymatio} is used to produce the 2D scattering coefficients. Our implementation of separable scattering is implemented similarly to Kymatio, with PyTorch \citep{pytorch} as a backend for FFT convolutions and \ac{nn} models.

Unless specified otherwise, all experiments have $\beta = \alpha = 2.5$. Scattering features are normalised prior to classification, according to the mean and \ac{std} calculated on the training set. No data augmentation is performed.

\begin{table}
\centering
\caption{MNIST classification error rate (\%) of separable and \ac{2d} scattering coefficients using a \ac{nn} classifier} \label{tab:mnistnn}
\begin{tabular}{|lr|l|} \hline
\multirow{2}{*}{2D WS + NN}        & $l=1$, $J=2$  & $0.64 \pm 0.05$    \\
          & $l=2$, $J=3$ & $0.50 \pm 0.03$     \\ \hline
\multirow{2}{*}{Separable WS + NN}  & $l=1$, $d=(4,4)$ & $0.63 \pm 0.05 $    \\
          & $l=2$, $d=(4,4)$  & $0.52 \pm 0.04$     \\ \hline
\end{tabular}
\end{table}

% Table \ref{tab:mnistlda} shows the results for MNIST using a \ac{lda} classifier from the Scikit-learn python package \citep{sklearn}, with a covariance shrinkage factor of $5\cdot 10^{-3}$. Scattering is performed for $l=1, 2$ levels.

% \begin{table}[!h] 
% \centering
% \caption{MNIST classification error rate (\%) of separable and \ac{2d} scattering coefficients  using an \ac{lda} classifier} \label{tab:mnistlda}
% \begin{tabular}{|r|rr|rr|} \hline
% \multirow{2}{*}{Training size}              & \multicolumn{2}{l}{2D}                               & \multicolumn{2}{l|}{Separable}                             \\
%  & \multicolumn{1}{l}{$l=1$} & \multicolumn{1}{l}{$l=2$} & \multicolumn{1}{l}{$l=1$} & \multicolumn{1}{l|}{$l=2$} \\ \hline
% 1000                              & 4.58                        & 2.69                        & 4.87                        & 4.58                        \\
% 2000                              & 3.89                        & 1.62                        & 4.31                        & 2.6                         \\
% 5000                              & 3.54                        & 1.12                        & 3.84                        & 1.7                         \\
% 10000                             & 3.35                        & 1.05                        & 3.54                        & 1.5                         \\
% 20000                             & 3.16                        & 0.87                        & 3.41                        & 1.41                        \\
% 40000                             & 2.92                        & 0.86                        & 3.34                        & 1.25                        \\
% 60000                             & 2.92                        & 0.81                        & 3.27                        & 1.34     \\ \hline                  
% \end{tabular}
% \end{table}

Due to its separability and non-angularly spaced filters, separable scattering does not perform as well compared to 2D scattering when using simple classifiers, such as \ac{lda} \citep{lda}. To illustrate that this performance discrepancy is not of significant consequence in a \ac{nn} setting, we test performance on the full dataset utilising a simple architecture. The neural network architecture used has an input layer with 256 neurons, followed by two hidden layers with 128 and 64 neurons respectively. The output layer has 10 neurons - one for each digit. Input and hidden layers are followed by a batch norm layer \citep{batchnorm} and ReLU activation function \citep{relu}. The output layer is followed by a softmax function. The Adam optimiser \citep{adam} is used with cross-entropy loss, a batch size of 256 and learning rate of $3 \cdot 10^{-5}$. 5000 of the 60000 training samples are reserved for validation and removed from the training set. Training is stopped when validation loss starts to increase. The \ac{nn} model is initialised with random weights, and the experiment is repeated 50 times. Different invariance scales were tested, and the best results are reported in table \ref{tab:mnistnn}. Tests are repeated for $l \in \{1, 2\}$ levels of scattering.

\subsection{MedMNIST3D}

The MedMNIST3D datasets are a subset of the MedMNIST dataset group \citep{medmnist}, where each 3D dataset contains $28\times 28 \times 28$ images with 2, 3 or 11 classes. Train, test and validation data partitions are provided by the authors. All datasets have on the order of 1000 training samples. We compare the baseline \ac{nn} results provided in \citep{medmnist} with separable scattering features combined with a simple \ac{nn} classifier.

We use \ac{nn} classifier with an input layer containing 1024 neurons, followed by two hidden layers with 512 and 256 neurons respectively. Input and hidden layers layers are each followed by a batch norm layer \citep{batchnorm} and a ReLU non-linearity \citep{relu}. For datasets with two classes, the output layer is a single neuron followed by a sigmoid activation function. For datasets with more than two classes, the output layer has a size equal to the number of classes, followed by a softmax activation. Binary cross-entropy loss are used for datasets with two classes, otherwise cross-entropy loss is used. The Adam optimiser \citep{adam} with a learning rate of $1 \cdot 10^{-5}$ is used. All other configuration parameters are identical to the model used for the MNIST dataset.

A single level of scattering coefficients are computed, with $Q = (2, 2, 2)$ and $d = (4, 4, 4)$. Many MedMNIST3D datasets tend to be unbalanced, implying that \ac{auc} is a more reliable metric to measure model performance. The results are shown in table \ref{tab:medmnist}, in which \ac{sota} \ac{auc} performance is achieved for the Organ, Adrenal and Vessel datasets. Table \ref{tab:medmnist} indicates the number of classes ($c$) for each of the datasets, with accuracy (ACC) also shown for reference. The performance of our method on non-\ac{sota} results are comparable to the other baseline \ac{nn} approaches presented in \citep{medmnist}. It is likely that better results can be achieved by the proposed method if the filters are made learnable and/or scattering parameters are uniquely optimised for each dataset.

### 2024-05-05 dissertation/content/paper3/original/main.tex
added dissertation

--- 2024-05-05 dissertation/content/paper3/main.tex
@@
-\chapter{Separable Wavelet Scattering}
+\documentclass[journal]{IEEEtran}
 
-\articleinfo{M.W.
-Rademan, D.J.J Verself, J.A.
-Du Preez}{IEEE Signal Processing Letters}{15 May 2024}{doi}{Wavelet scattering is a widely used feature extraction method due to it efficacy in extracting invariant features,
-while retaining any lost high frequency information resulting from averaging to obtain the desired amount of invariance. 
+\usepackage{graphicx} % Required for inserting images
+\usepackage{acronym}
+\usepackage{mathtools} 
+\usepackage{extarrows} 
+\usepackage{amsmath}
+\usepackage{amssymb}
+\usepackage{upgreek}
+\usepackage{algorithm}
+\usepackage{algpseudocode}
+\usepackage{mathtools}
+\usepackage{multirow}
+\usepackage{float}
+
+\acrodef{1d}[1D]{1-dimensional}
+\acrodef{2d}[2D]{2-dimensional}
+\acrodef{cnn}[CNN]{convolutional neural network}
+\acrodef{mfcc}[MFCC]{Mel-frequency cepstral coefficients}
+\acrodef{sift}[SIFT]{scale-invariant feature transform}
+\acrodef{ws}[WS]{wavelet scattering}
+\acrodef{tf}[TF]{time-frequency}
+\acrodef{lpf}[LPF]{low-pass filter}
+\acrodef{gpu}[GPU]{graphics processing unit}
+\acrodef{fir}[FIR]{finite impulse response}
+\acrodef{sota}[SOTA]{state-of-the-art}
+\acrodef{fft}[FFT]{fast Fourier transform}
+\acrodef{lda}[LDA]{linear discriminant analysis}
+\acrodef{nn}[NN]{neural network}
+\acrodef{auc}[AUC]{area under curve}
+\acrodef{std}[SD]{standard deviation}
+
+\usepackage{bm}
+\newcommand{\vect}[1]{\boldsymbol{\mathbf{#1}}}
+
+\DeclarePairedDelimiter\ceil{\lceil}{\rceil}
+\DeclarePairedDelimiter\floor{\lfloor}{\rfloor}
+
+\title{Separable Wavelet Scattering}
+\author{M.W. Rademan, D.J.J Versfeld, J.A. du Preez }
+\date{November 2023}
+
+\begin{document}
+
+\maketitle
+
+\begin{abstract}
+Wavelet scattering is a widely used feature extraction method due to it efficacy in extracting invariant features, while retaining any lost high frequency information resulting from averaging to obtain the desired amount of invariance.
 By generalising a 1-dimensional scattering transform, we extend its definition to an arbitrary number of independent dimensions.
-We show that, in a modern deep learning setting, separable wavelet scattering performs similarly to its non-separable counterparts 
-with the MNIST hand-written digit dataset. We additionally demonstrate state-of-the-art results for a subset of the MedMNIST3D datasets.}
+We show that, in a modern deep learning setting, separable wavelet scattering performs similarly to its non-separable counterparts with the MNIST hand-written digit dataset.
+We additionally demonstrate state-of-the-art results for a subset of the MedMNIST3D datasets.
+\end{abstract}
+
+\begin{IEEEkeywords}
+classification, MNIST, MedMNIST, wavelet scattering
+\end{IEEEkeywords}
 
@@
 It is an especially effective method for small datasets, since the feature filters are reminiscent of a \ac{cnn} front-end, while requiring no learning.
-The resulting features are invariant across all averaged dimensions, and exhibit separable class subspaces, allowing linear classifiers to be used with high efficacy \citep{2dscattering}.
+The resulting features are invariant across all averaged dimensions, and exhibit separable class subspaces, allowing linear classifiers to be used with high efficacy \cite{2dscattering}.
 
-Various forms of \ac{ws} exist, with \ac{1d} scattering first proposed by Anden and Mallat \citep{1dscattering1, 1dscattering2}, which was later extended to 2 \citep{2dscattering} and 3 dimensions \citep{3dscattering, harmonicscattering}.
-Additional and more specialised forms for \ac{ws} include rotation-invariant scattering \citep{groupinvariantscattering} for the classification of textures and joint-\ac{tf} scattering \citep{jointtfscattering1, jointtfscattering2}.
+Various forms of \ac{ws} exist, with \ac{1d} scattering first proposed by Anden and Mallat \cite{1dscattering1, 1dscattering2}, which was later extended to 2 \cite{2dscattering} and 3 dimensions \cite{3dscattering, harmonicscattering}.
+Additional and more specialised forms for \ac{ws} include rotation-invariant scattering \cite{groupinvariantscattering} for the classification of textures and joint-\ac{tf} scattering \cite{jointtfscattering1, jointtfscattering2}.
 Joint-\ac{tf} scattering utilises a separable \ac{2d} filter that operates on the first level scattering scalogram, denoted by the operator $\mathcal{U}_1$, which is the only apparent usage of separable filters in the current scattering literature.
 
-In a deep-learning setting, sensible initialisation of filters prior to learning can significantly improve performance and interpretability \citep{sincnet}.
-The greatest advantage of separable filters is their computation speed when computing convolutions directly, as is performed in \acp{cnn} \citep{separablecnn}.
+In a deep-learning setting, sensible initialisation of filters prior to learning can significantly improve performance and interpretability \cite{sincnet}.
+The greatest advantage of separable filters is their computation speed when computing convolutions directly, as is performed in \acp{cnn} \cite{separablecnn}.
 Additionally, separable filters reduce the number of parameters of a \ac{cnn} if the filters are configured to be learnable.
-Learnable filters can typically improve performance compared to their fixed-filters counterpart \citep{learnablewavelettransform}.
+Learnable filters can typically improve performance compared to their fixed-filters counterpart \cite{learnablewavelettransform}.
 
 At first glance, a \ac{ws} decomposition seems to be prohibitively expensive, but fast algorithms are possible due to the limited bandwidth of filter output.
-Fast algorithms utilise downsampling to take advantage of the demodulation of band-limited signals that result from the convolution of analytic wavelets \citep{2dscattering, 3dscattering, 1dscattering1}.
+Fast algorithms utilise downsampling to take advantage of the demodulation of band-limited signals that result from the convolution of analytic wavelets \cite{2dscattering, 3dscattering, 1dscattering1}.
 Path pruning is also used to ignore filter combinations which have negligible energy.
-These algorithms are the standard implementation on many platforms, such as the Kymatio python package \citep{kymatio} and MATLAB \citep{MATLAB}.
+These algorithms are the standard implementation on many platforms, such as the Kymatio python package \cite{kymatio} and MATLAB \cite{MATLAB}.
 
@@
 We propose flexible definitions for the Morlet bandwidth and filter overlap, which allows for an exact definition of near-optimal convolution computations with downsampling.
-Our implementation of separable scattering is \ac{gpu} accelerated, similar to the Kymatio implementations \citep{kymatio}.
+Our implementation of separable scattering is \ac{gpu} accelerated, similar to the Kymatio implementations \cite{kymatio}.
 
-We demonstrate using the MNIST handwritten digit dataset \citep{mnist} that separable wavelet filters still perform adequately in a \ac{nn} classification setting compared to the conventional 2D scattering filters.
-We demonstrate the arbitrarily scalable dimensionality of the separable scattering transform with three-dimensional medical datasets from MedMNIST3D \citep{medmnist}.
+We demonstrate using the MNIST handwritten digit dataset \cite{mnist} that separable wavelet filters still perform adequately in a \ac{nn} classification setting compared to the conventional 2D scattering filters.
+We demonstrate the arbitrarily scalable dimensionality of the separable scattering transform with three-dimensional medical datasets from MedMNIST3D \cite{medmnist}.
 Separable scattering coefficients show \ac{sota} results for some of the MedMNIST3D datasets when combined with a simple \ac{nn}, while performing comparably to \ac{cnn} benchmarks on datasets which do not indicate \ac{sota} results.
@@
 For an arbitrary bandwidth definition, we define the bandwidth-to-$\sigma$ ratio $\beta \in \mathbb{R}^+$, such that the one-sided bandwidth of a zero-mean Gaussian is $\beta \sigma_\omega$.
-For reference, Kymatio's implementation of 2D scattering indirectly defines $\beta \approx 2.5$ \citep{kymatio}.
+For reference, Kymatio's implementation of 2D scattering indirectly defines $\beta \approx 2.5$ \cite{kymatio}.
 
@@
 
-The provided definitions of filter clearance ($\alpha$) and bandwidth ($\beta$) may also be extended to non-separable wavelets, such as the rotational construction employed in \citep{2dscattering}.
+The provided definitions of filter clearance ($\alpha$) and bandwidth ($\beta$) may also be extended to non-separable wavelets, such as the rotational construction employed in \cite{2dscattering}.
 However, a conversion is required to find the bandwidth in each axis for a specific wavelet for rotationally constructed filterbanks, in order to employ the downsampling strategies proposed in this paper.
@@
 On-axis coverage is also required, in which each $\uppsi_\lambda$ must be multiplied with a Gaussian (zero-frequency wavelet), which the dilated wavelet definition in equation (\ref{eqn:dilwav}) defines as $\lambda=0$.
-A similar construction procedure is followed in \citep{jointtfscattering2}.
+A similar construction procedure is followed in \cite{jointtfscattering2}.
 
@@
 \end{gather}
-where $\cup$ indicates the set union operator and $\times$ the Cartesian product. For $m=1$, the provided definitions result in a conventional \ac{1d} scattering transform \citep{1dscattering1}.
+where $\cup$ indicates the set union operator and $\times$ the Cartesian product. For $m=1$, the provided definitions result in a conventional \ac{1d} scattering transform \cite{1dscattering1}.
 
@@
 % In this section, we describe the efficient computation of $m$-dimensional separable scattering transforms.
-The described downsampling strategy is utilised to a limited extent in Kymatio scattering implementations \citep{kymatio} and the original proposed fast algorithms \citep{2dscattering}, which is generalised further in this section.
+The described downsampling strategy is utilised to a limited extent in Kymatio scattering implementations \cite{kymatio} and the original proposed fast algorithms \cite{2dscattering}, which is generalised further in this section.
 
@@
 The \ac{lpf} $\phi$ remains constant throughout the transform.
-Note that the modulus/magnitude operator $|\cdot|$ demodulates the output of the filters, effectively extracting the Hilbert envelope from a band-limited signal \citep{waveletsandsubbandcoding}.
+Note that the modulus/magnitude operator $|\cdot|$ demodulates the output of the filters, effectively extracting the Hilbert envelope from a band-limited signal \cite{waveletsandsubbandcoding}.
 
@@
 \centering
-\caption{MedMNIST classification results of compared to baseline \ac{nn} approaches (\citep{medmnist})} \label{tab:medmnist}
+\caption{MedMNIST classification results of compared to baseline \ac{nn} approaches (\cite{medmnist})} \label{tab:medmnist}
 \begin{tabular}{|l|rr|rr|rr|rr|rr|rr|}
@@
 
-Given that $r|N$, we can express (\ref{eqn:circconv}) when downsampled as a periodised summation \citep{waveletsandsubbandcoding} in the frequency domain
+Given that $r|N$, we can express (\ref{eqn:circconv}) when downsampled as a periodised summation \cite{waveletsandsubbandcoding} in the frequency domain
 \begin{equation}
@@
 \end{equation}
-which can be efficiently implemented via shape manipulation of tensors in computational packages like MATLAB or PyTorch \citep{pytorch}.
+which can be efficiently implemented via shape manipulation of tensors in computational packages like MATLAB or PyTorch \cite{pytorch}.
 To compute valid convolutions, we must pad $x$ and $\uppsi$ to have a total length of $N = N_x + d + c$, where $c \in \mathbb{N}^+$ is a constant that ensures that $d | N$.
@@
 
-% We benchmark separable \ac{ws} against 2D scattering by repeating the MNIST hand-written digit dataset experiment in \citep{2dscattering}.
+% We benchmark separable \ac{ws} against 2D scattering by repeating the MNIST hand-written digit dataset experiment in \cite{2dscattering}.
 
-The MNIST dataset \citep{mnist} contains 60000 training and 10000 test samples.
-Unlike in \citep{2dscattering}, which decorrelates scattering coefficients with a discrete cosine transform prior to classification, we perform classification on the scattering coefficients directly.
-Kymatio \citep{kymatio} is used to produce the 2D scattering coefficients.
-Our implementation of separable scattering is implemented similarly to Kymatio, with PyTorch \citep{pytorch} as a backend for FFT convolutions and \ac{nn} models.
+The MNIST dataset \cite{mnist} contains 60000 training and 10000 test samples.
+Unlike in \cite{2dscattering}, which decorrelates scattering coefficients with a discrete cosine transform prior to classification, we perform classification on the scattering coefficients directly.
+Kymatio \cite{kymatio} is used to produce the 2D scattering coefficients.
+Our implementation of separable scattering is implemented similarly to Kymatio, with PyTorch \cite{pytorch} as a backend for FFT convolutions and \ac{nn} models.
 
@@
 
-% Table \ref{tab:mnistlda} shows the results for MNIST using a \ac{lda} classifier from the Scikit-learn python package \citep{sklearn}, with a covariance shrinkage factor of $5\cdot 10^{-3}$.
+% Table \ref{tab:mnistlda} shows the results for MNIST using a \ac{lda} classifier from the Scikit-learn python package \cite{sklearn}, with a covariance shrinkage factor of $5\cdot 10^{-3}$.
 Scattering is performed for $l=1, 2$ levels.
@@
 
-Due to its separability and non-angularly spaced filters, separable scattering does not perform as well compared to 2D scattering when using simple classifiers, such as \ac{lda} \citep{lda}.
+Due to its separability and non-angularly spaced filters, separable scattering does not perform as well compared to 2D scattering when using simple classifiers, such as \ac{lda} \cite{lda}.
 To illustrate that this performance discrepancy is not of significant consequence in a \ac{nn} setting, we test performance on the full dataset utilising a simple architecture.
@@
 The output layer has 10 neurons - one for each digit.
-Input and hidden layers are followed by a batch norm layer \citep{batchnorm} and ReLU activation function \citep{relu}.
+Input and hidden layers are followed by a batch norm layer \cite{batchnorm} and ReLU activation function \cite{relu}.
 The output layer is followed by a softmax function.
-The Adam optimiser \citep{adam} is used with cross-entropy loss, a batch size of 256 and learning rate of $3 \cdot 10^{-5}$.
+The Adam optimiser \cite{adam} is used with cross-entropy loss, a batch size of 256 and learning rate of $3 \cdot 10^{-5}$.
 5000 of the 60000 training samples are reserved for validation and removed from the training set.
@@
 
-The MedMNIST3D datasets are a subset of the MedMNIST dataset group \citep{medmnist}, where each 3D dataset contains $28\times 28 \times 28$ images with 2, 3 or 11 classes.
+The MedMNIST3D datasets are a subset of the MedMNIST dataset group \cite{medmnist}, where each 3D dataset contains $28\times 28 \times 28$ images with 2, 3 or 11 classes.
 Train, test and validation data partitions are provided by the authors.
 All datasets have on the order of 1000 training samples.
-We compare the baseline \ac{nn} results provided in \citep{medmnist} with separable scattering features combined with a simple \ac{nn} classifier.
+We compare the baseline \ac{nn} results provided in \cite{medmnist} with separable scattering features combined with a simple \ac{nn} classifier.
 
 We use \ac{nn} classifier with an input layer containing 1024 neurons, followed by two hidden layers with 512 and 256 neurons respectively.
-Input and hidden layers layers are each followed by a batch norm layer \citep{batchnorm} and a ReLU non-linearity \citep{relu}.
+Input and hidden layers layers are each followed by a batch norm layer \cite{batchnorm} and a ReLU non-linearity \cite{relu}.
 For datasets with two classes, the output layer is a single neuron followed by a sigmoid activation function.
@@
 Binary cross-entropy loss are used for datasets with two classes, otherwise cross-entropy loss is used.
-The Adam optimiser \citep{adam} with a learning rate of $1 \cdot 10^{-5}$ is used.
+The Adam optimiser \cite{adam} with a learning rate of $1 \cdot 10^{-5}$ is used.
 All other configuration parameters are identical to the model used for the MNIST dataset.
@@
 Table \ref{tab:medmnist} indicates the number of classes ($c$) for each of the datasets, with accuracy (ACC) also shown for reference.
-The performance of our method on non-\ac{sota} results are comparable to the other baseline \ac{nn} approaches presented in \citep{medmnist}.
+The performance of our method on non-\ac{sota} results are comparable to the other baseline \ac{nn} approaches presented in \cite{medmnist}.
 It is likely that better results can be achieved by the proposed method if the filters are made learnable and/or scattering parameters are uniquely optimised for each dataset.
+
+% \section{Conclusion}
+
+\newpage
+\bibliographystyle{ieeetr} 
+\bibliography{refs}
+
+\end{document}

### 2024-05-05 dissertation/main.tex
added dissertation

\documentclass[11pt,oneside,openany,a4paper, %... Global layout
afrikaans,UKenglish, %... Global language options
PhD, a5block %... stb-thesis options
]{stb-thesis}

\usepackage[english]{babel}
\usepackage{graphicx}
\usepackage[utf8]{inputenc}
\usepackage[onehalfspacing]{setspace}
\usepackage{color}
\usepackage{parskip}%......... Paragraph spacing + zero indent
\usepackage{amssymb} % only if necessary
\usepackage{bm} %.. Bold math characters (after fonts)
\usepackage{hyperref}

\usepackage[numbers]{stb-bib}%........................... Bibliography
%.......... Auhor-year style
\addto{\captionsafrikaans}{\renewcommand{\bibname}{Lys van Verwysings}}
\addto{\captionsUKenglish}{\renewcommand{\bibname}{List of References}}

% \iftrue
% \usepackage[backref=page]{hyperref}%.... Hyperlinks & backreferences
% \else
% \usepackage[pageref]{backref}%........... Citation backreferences
% \fi
% \renewcommand*{\backrefalt}[4]{%s
% \ifcase #1 (Not cited.)%
% \or (Cited on page~#2.)%
% \else (Cited on pages~#2.)%
% \fi}

\newcommand\articleinfo[5]{%
\noindent\rule{\textwidth}{0.5pt}
\begin{center}\small\textit{#1}\par
    \begin{spacing}{0.1}
    \small Accepted in #2 on #3\par
    \small doi: #4\noindent
\end{spacing}\noindent
\end{center}\noindent
\noindent\rule{\textwidth}{0.25pt}%
\section*{\centering \textsc{Abstract}}
\begin{spacing}{1.0}    
    \small  #5 \noindent
\end{spacing}
\noindent\rule{\textwidth}{0.5pt}\pagebreak
}

\usepackage{paper1}
\usepackage{paper2}
\usepackage{paper3}

% Subappendices
\usepackage{appendix}
% Start of subappendices environment
\AtBeginEnvironment{subappendices}{%
\section*{Appendices}
\addcontentsline{toc}{section}{Appendices}
\counterwithin{figure}{section}
\counterwithin{table}{section}
}

% End of subappendices environment
\AtEndEnvironment{subappendices}{%
\counterwithout{figure}{section}
\counterwithout{table}{section}
}

\graphicspath{{./content/paper2/fig/}{./content/paper1/fig/}}

\usepackage{iftex}
\ifxetex
\usepackage[math-style=TeX, bold-style=TeX]{unicode-math}
\setmainfont{Cambria}%........................ (Windows)
\setsansfont[Scale=MatchLowercase]{Calibri}
\setmonofont[Scale=MatchLowercase]{Consolas}
\setmathfont{Cambria Math}
\defaultfontfeatures{Ligatures=TeX}
\let\bm\symbfit%.............................. Bold math
\else
\usepackage{textcomp}%........................ Additional text symbols
\usepackage[T1]{fontenc}%..................... Type 1 outline fonts
\usepackage{bm}%.............................. Bold math fonts
\fi

\begin{document}
\title{\textbf{Wavelet Scattering Tranforms Applied to Whale Vocalisations}}
\author{M.W. Rademan}{Marco Wiehann Rademan}
\faculty{Electrical and Electronic Engineering}
\degree{PhD}{Doctor of Philosophy}
\supervisor{D.J.J. Versfeld}
\cosupervisor{J.A. Du Preez}
% \date{August}{2024} 
\SetCopyrightOff
\address{\AorE{Departement Siviele Ingenieurswese,\\ Stellenbosch Universiteit,\\
Privaatsak X1, 7602 Matieland, Suid Afrika.
}{%
Department of Civil Engineering,\\ Stellenbosch University,\\
Private Bag X1, 7602 Matieland, South Africa.}}

% \frontmatter
\begin{singlespace}
\TitlePage%..................... Single spaced title page
\end{singlespace}

\begin{singlespace}
\DeclarationDate{\today}
\DeclarationPage
\end{singlespace}
% \DeclarationPage

\begin{abstract}

\end{abstract}

\begin{abstract}[afrikaans]

\end{abstract}

\tableofcontents
\listoffigures
\listoftables
\chapter*{Nomenclature}
\label{chap:nomenclature}
\addcontentsline{toc}{chapter}{\nameref{chap:nomenclature}}

% \begin{singlespace}
\input{content/introduction}

\input{content/p1_intro}
\input{content/paper1/main}

\input{content/p2_intro}
\input{content/paper2/main}

\input{content/p3_intro}
\input{content/paper3/main}
% \end{singlespace}

\bibliographystyle{stb-bib-eng-a}
\bibliography{references}

\end{document}

### 2024-05-05 python/mnist_clf.py

@@ -27,7 +27,7 @@ torch.cuda.empty_cache()
 from kymatio.torch import Scattering2D
 
 
-TRAIN_SIZES = [300,1000,2000,5000,10000,20000,40000,60000]
+TRAIN_SIZES = [1000,2000,5000,10000,20000,40000,60000]
 Q_CONFIGS = [
     [[1, 1]],
     [[1, 1], [1, 1]]
@@ -76,25 +76,25 @@ for iq, Q in enumerate(Q_CONFIGS):
     print("Sep Scattering took {:.2f} ms".format((t1 - t0)*1000))
     print(S_train_sep.shape)
     
-    # ws_2d = Scattering2D(J=3, shape=(28, 28), max_order=iq+1)
-    # ws_2d.cuda()
+    ws_2d = Scattering2D(J=3, shape=(28, 28), max_order=iq+1)
+    ws_2d.cuda()
     
-    # t0 = time()
-    # S_train_2d: torch.Tensor = ws_2d.scattering(X_train)
-    # S_test_2d: torch.Tensor  = ws_2d.scattering(X_test)   
-    # torch.cuda.synchronize()
-    # t1 = time()
-    # print("2D Scattering took {:.2f} ms".format((t1 - t0)*1000))
-    # S_train_2d = S_train_2d.swapaxes(1, -1)
-    # S_test_2d = S_test_2d.swapaxes(1, -1)
-    # print(S_train_2d.shape)
-    # print('2D DEVICE', S_test_2d.device)
+    t0 = time()
+    S_train_2d: torch.Tensor = ws_2d.scattering(X_train.cuda())
+    S_test_2d: torch.Tensor  = ws_2d.scattering(X_test.cuda())   
+    torch.cuda.synchronize()
+    t1 = time()
+    print("2D Scattering took {:.2f} ms".format((t1 - t0)*1000))
+    S_train_2d = S_train_2d.swapaxes(1, -1)
+    S_test_2d = S_test_2d.swapaxes(1, -1)
+    print(S_train_2d.shape)
+    print('2D DEVICE', S_test_2d.device)
 
     # #to numpy
     S_train_sep: np.ndarray = S_train_sep.cpu().numpy()
     S_test_sep: np.ndarray = S_test_sep.cpu().numpy()
-    # S_train_2d: np.ndarray = S_train_2d.cpu().numpy()
-    # S_test_2d: np.ndarray = S_test_2d.cpu().numpy()
[…]

### 2024-05-05 python/mnist_clf_nn.py

@@ -286,7 +286,7 @@ t1 = time()
 print("Sep Scattering took {:.2f} ms".format((t1 - t0)*1000))
 print(S_train_sep.shape)
 
-ws_2d = Scattering2D(J=2, shape=(28, 28), max_order=1)
+ws_2d = Scattering2D(J=3, shape=(28, 28), max_order=2)
 ws_2d.cuda()
 
 t0 = time()
@@ -323,11 +323,11 @@ print(S_train_2d.shape)
 sep_err = []
 _2d_err = []
 
-for i in range(50):
+for i in range(1):
 
     net = DeepClassifier(S_train_sep.shape[1],[256, 128, 64], 10)
     trainer = LinearTrainer(net)
-    print(trainer.num_trainable_parameters())
+    print("SEP PARAMETERS", trainer.num_trainable_parameters())
     trainer.train(S_train_sep, y_train, S_val_sep, y_val, n_epochs=100, lr=3e-5)
     acc, _ = trainer.test_acc(S_test_sep, y_test)
     err_prc = (1-acc)*100
@@ -337,7 +337,7 @@ for i in range(50):
 
     net = DeepClassifier(S_train_2d.shape[1],[256, 128, 64], 10)
     trainer = LinearTrainer(net)
-    print(trainer.num_trainable_parameters())
+    print("2D PARAMETERS", trainer.num_trainable_parameters())
     trainer.train(S_train_2d, y_train, S_val_2d, y_val, n_epochs=100, lr=3e-5)
     acc, _ = trainer.test_acc(S_test_2d, y_test)
     err_prc = (1-acc)*100   

### 2024-05-05 python/sepws/scattering/filterbank.py

@@ -13,7 +13,8 @@ def calculate_padding_1d(N: int, d: int):
     left = d//2
     c = d - (N % d)
     right = left + c
-    return left, right, N + left + right
+    N_pad = N + left + right
+    return (N_pad - N)//2, (N_pad - N)//2, N_pad
 
 def calculate_sigma_psi_w(Q):
     alpha = cfg.get_alpha(Q)
@@ -291,8 +292,8 @@ def get_Lambda_set(fb: List, level: int, input_ds: List[int]):
     lambdas = [[0] + list(fb[level][dim][input_ds[dim]]['psi'].keys()) for dim in range(Ndims)] # list of all lambdas in each dimension, including the 0 filter
     Lambda = list(product(*lambdas)) #the cross product set, which still includes the 0 filter
     Lambda = Lambda[1:] #remove the 0 filter
-    # lambdas = [list(fb[level][dim][input_ds[dim]]['psi'].keys()) for dim in range(Ndims)] # list of all lambdas in each dimension, including the 0 filter
-    # Lambda = list(product(*lambdas)) #the cross product set, which still includes the 0 filter
+    # lambdas = [list(fb[level][dim][input_ds[dim]]['psi'].keys()) for dim in range(Ndims)] 
+    # Lambda = list(product(*lambdas)) 
     
     # Lambda_final = []
     # for L in Lambda:

### 2024-05-05 python/sepws/scattering/separable_scattering.py

@@ -26,7 +26,6 @@ class SeparableScattering:
         filterbank_to_tensor(self.fb)
         
     def _mul_and_downsample(self, X: Tensor, level: int, input_ds: List[int], lambdas: List[float]):
-        #TODO: Precalculate filters        
         mul1d = lambda x, y, dim: self.backend.mul1d(x, y, dim+1) #+1 to account for batch dim
         freqds = lambda x, d, dim: self.backend.freq_downsample1d(x, d, dim+1)
         for dim in range(self.Ndim):
@@ -34,6 +33,18 @@ class SeparableScattering:
             ds = get_output_downsample_factor(self.fb, level, dim, input_ds[dim], lambdas[dim])
             X = mul1d(X, filter, dim)
             if self.allow_ds: X = freqds(X, ds, dim)
+        return X
+    
+    def _ifft_mul_and_downsample(self, X: Tensor, level: int, input_ds: List[int], lambdas: List[float]):
+        mul1d = lambda x, y, dim: self.backend.mul1d(x, y, dim+1) #+1 to account for batch dim
+        freqds = lambda x, d, dim: self.backend.freq_downsample1d(x, d, dim+1)
+        ifft1d = lambda x, dim: self.backend.ifft1d(x, dim)
+        for dim in range(self.Ndim):
+            filter = get_wavelet_filter(self.fb, level, dim, input_ds[dim], lambdas[dim])  
+            ds = get_output_downsample_factor(self.fb, level, dim, input_ds[dim], lambdas[dim])
+            X = mul1d(X, filter, dim)
+            if self.allow_ds: X = freqds(X, ds, dim)
+            X = ifft1d(X, dim)
         return X
     
     def _ifft_all(self, X: Tensor):
@@ -68,7 +79,7 @@ class SeparableScattering:
             # these intervals must overlap, 
             # * is the centre of the spectrum, d is the significant bandwidth of the demodulated filter (via modulus), and f is the morlet filter under consideration which has a center x
             EPS = 1e-9 #for floating point error
-            if abs(lambda_filt[i]) > beta * sigma_psi_w_demod + EPS: return True 
+            if abs(lambda_filt[i]) > beta*sigma_psi_w_demod + EPS: return True 
             # if abs(lambda_filt[i]) > abs(lambda_demod[i]): return True 
             bws.append(abs(lambda_filt[i]) - sigma_psi_w_filt > sigma_psi_w_demod)           
         return False
@@ -121,6 +132,10 @@ class SeparableScattering:
[…]

### 2024-05-06 dissertation/content/nomenclature.tex
before reference consolodation

\section*{Notation}
Some notations are used interchangably, which is obvious from the context. Due to the nature of the publication format, notation is explained within the article it occurs.

\begin{tabular}{p{0.15\textwidth} p{0.75\textwidth}}
    $f(x)$ & a continuous function of $x$ \\
    $f(x, y)$ & a continuous multi-variable function of $x$ and $y$ \\
    $f[n]$ & a discrete-time (sampled) function indexed by $n$ \\
    $f[n, m]$ & a multi-variable discrete-time function indexed by $n$ and $m$ \\
    $\mathcal{F} \{f\}$ & the Fourier transform operator acting on $f$\\
    $\hat{f}$ & the Fourier transform of $f$ or an estimate of $f$, obvious from context\\
    $x^*$ & the complex conjugate of $x$\\
    $p(x)$ & a probability density \\
    $p(x|y)$ & a conditional probability density \\
    $\{a, b, ...\}$ & a set of numbers\\
    $A \setminus B$ & set minus: the set $A$ excluding the elements of $B$\\
    $A \times B$ & Cartesian product of sets\\
    $| x |$ & magnitude of a number (modulus), or size of a set\\
    $\mathbb{R}$ & the set of real numbers\\
    $\mathbb{C}$ & the set of complex numbers\\
    $\mathbb{N^+}$ & the set of natural numbers (excluding 0)\\
    $\mathbb{Z}$ & the set of integers\\
    $x \sim p$ & $x$ is drawn from the probability density $p$\\
    $\frac{d}{dx} f$ & derivative with respect to $x$ of a function $f$\\
    $\mathcal{U}_j x$ & scalogram of the $j$'th level of $x$\\
    $\mathcal{S}_j x$ & scattering coefficients of the $j$'th level of $x$\\
    $x * y$ & convolution of two functions\\
    $x \otimes y$ & circular convolution of two discrete functions\\
    $(x)_{\downarrow r}$ & a discrete function subsampled by a factor $r$\\
    $\mathbf{x}$ & a vector\\
\end{tabular}

\begin{tabular}{p{0.15\textwidth} p{0.75\textwidth}}
    $X[k,m]$ & short-time Fourier transform of $x$, with $k$ the frequency index and $m$ the time index \\
    $S[k,m]$ & power spectrum calculated from $X[k, m]$ \\
    $H[m]$ & spectral entropy calculated from $S[k, m]$ \\
    $X(s, b)$ & continuous wavelet transform of $x$, with $s$ as scale and $b$ as translation \\
    $\uppsi_\lambda$ & sometimes used as short-hand for  $\uppsi(\lambda t)$\\
    $\text{RM}^F_m (x)$ & rolling median filter operator with a window of size $F$ operating on index variable $m$ on a discrete function $x$\\
    $\text{RA}^F_m (x)$ & rolling average filter operator with a window of size $F$ operating on index variable $m$ on a discrete function $x$\\
\end{tabular}

\clearpage
\section*{Common Variables}
Some variables are used interchangably, which is obvious from the context. Due to the nature of the publication format, variables are explained within the article it occurs.

\begin{tabular}{p{0.15\textwidth} p{0.75\textwidth}}
    a & Hello \\
    a & Hello \\
    a & Hello \\
    a & Hello \\
\end{tabular}

\section*{Acronyms}
\begin{tabular}{p{0.15\textwidth} p{0.75\textwidth}}
    GMM & Gaussian mixture model \\
    POI & point(s) of interest \\
    EM & expectation maximization \\
    SE & spectral entropy \\
    PAM & passive acoustic monitoring \\
    PSD & power spectral density \\
    STFT & short-time Fourier transform \\
    TPR & true positive rate \\
    FPR & false positive rate \\
    PDF & probability density function \\
    SNR & signal-to-noise ratio \\
    AMFM & amplitude and frequency modulated \\
    BLED & band-limited energy detection \\
    TF & time-frequency \\
    CWT & continuous wavelet transform \\
    DWT & discrete wavelet transform \\
    SWT & stationary wavelet transform \\
    WPD & wavelet packet decomposition \\
    MODWT & maximal overlap discrete wavelet transform \\
    MFCC & Mel-frequency cepstral coefficient \\
    DNN & deep neural network \\
    CNN & convolutional neural network \\
    HMM & hidden Markov model \\
    SOTA & state-of-the-art \\
    LSTM & long-short-term memory \\
    EWT & Endangered Wildlife Trust \\
    WS & wavelet scattering \\
    WST & wavelet scattering transform \\
    FFT & fast Fourier transform \\
    GPU & graphics processing unit \\
    ROC & receiver operating characteristic \\
    CLS & cosine-log scattering \\
    LDA & linear discriminant analysis \\
    ML & machine learning \\
    NN & neural network \\
    SVM & support vector machine \\
\end{tabular}

\begin{tabular}{p{0.15\textwidth} p{0.75\textwidth}}

    1D & 1-dimensional \\
    2D & 2-dimensional \\
    SIFT & scale-invariant feature transform \\
    LPF & low-pass filter \\
    IR & impulse response \\
    FIR & finite impulse response \\
    AUC & area under curve \\
    SD & standard deviation \\
    STPR & soft true positive ratio \\
    STNR & soft true negative ratio \\

\end{tabular}

% TODO: paper 1 acronyms and consolodate

### 2024-05-06 dissertation/content/paper2/main.tex
before reference consolodation

@@
 \articleinfo{M.W.
-Rademan, D.J.J Verself, J.A.
-Du Preez}{Journal of the Acoustical Society of America (JASA)}{15 May 2024}{doi}{Current literature trends for whale classification in passive acoustic monitoring systems are mainly concerned with high accuracies obtained via deep learning approaches.
+Rademan, D.J.J Verseld, J.A.
+Du Preez}{the Journal of the Acoustical Society of America (JASA)}{15 May 2024}{doi}{Current literature trends for whale classification in passive acoustic monitoring systems are mainly concerned with high accuracies obtained via deep learning approaches.
 This requires preparation of large and accurate annotated datasets, which are often not available when surveying new regions, nor are they wholly accurate when they are.
@@
 
-% \begin{table}[h!]
-% \caption{Total number of annotations in the Casey 2017 dataset.}
-% \label{tab:numannotations}
+\begin{table}[h!]
+\caption{Total number of annotations in the Casey 2017 dataset.}
+\label{tab:numannotations}
 % \begin{ruledtabular}  
 % \renewcommand{\arraystretch}{0.8}
-% \begin{tabular}{lrr} 
-% \textbf{Vocalisation}      & \textbf{Number of annotations} & \textbf{Estimated False positives (\%)} \\ \hline
-% Bm\_Ant\_A & 1741  & 5 - 10               \\
-% Bm\_Ant\_B & 558  & 5 - 10                 \\
-% Bm\_Ant\_Z & 119    & 1 - 2               \\
-% Bm\_D      & 553     & 8 - 15              \\
-% Bp\_20Hz & 78        & 1 - 3             \\
-% Bp\_20Plus & 214     & 1 - 3               \\ 
-% Total & 3263 & 5 - 10\\ 
-% \end{tabular}
+\begin{tabular}{lrr} 
+\textbf{Vocalisation}      & \textbf{Number of annotations} & \textbf{Estimated False positives (\%)} \\ \hline
+Bm\_Ant\_A & 1741  & 5 - 10               \\
+Bm\_Ant\_B & 558  & 5 - 10                 \\
+Bm\_Ant\_Z & 119    & 1 - 2               \\
+Bm\_D      & 553     & 8 - 15              \\
+Bp\_20Hz & 78        & 1 - 3             \\
+Bp\_20Plus & 214     & 1 - 3               \\ 
+Total & 3263 & 5 - 10\\ 
+\end{tabular}
 % \end{ruledtabular}
-% \end{table}
+\end{table}
 
@@
 
-    % \clearpage
-
-    % \section{Summary of parameter values}
-
-    % \begin{table}[h]
-        % \caption{Summary of parameter values used in this study. Chosen values are indicated in boldface text, whereas non-bold parameters result from the chosen parameters.}
-        % \label{tab:parameters}
+    \clearpage
+
+    \section{Summary of parameter values}
+
+    \begin{table}[h]
+        \caption{Summary of parameter values used in this study. Chosen values are indicated in boldface text, whereas non-bold parameters result from the chosen parameters.}
+        \label{tab:parameters}
         % \begin{ruledtabular}    
-            % \renewcommand{\arraystretch}{0.8}% Tighter
-            % \begin{tabular}{>{\centering\arraybackslash}l>{\centering\arraybackslash}l>{\centering\arraybackslash}l}  \small
-                % \textbf{Parameter}      & \textbf{Description} & \textbf{Value} \\ \hline
-                % $T$ & \ac{ws} invariance scale & \textbf{1.5 seconds} \\
-                % $Q_1$ & Wavelets per octave in $\mathcal{S}_1$ & \textbf{16} \\
-                % $Q_2$ & Wavelets per octave in $\mathcal{S}_2$ & \textbf{4} \\
-                % $f_\text{min}$ & Minimum filterbank frequency in $\mathcal{S}_1$ & \textbf{15 Hz}\\
-                % $f_\text{max}$ & Maximum filterbank frequency in $\mathcal{S}_1$ & \textbf{120 Hz}\\
-                % $|\Lambda_1|$ & Resulting number of filters in $\mathcal{S}_1$ & 48 \\
-                % $|\Lambda_2|$ & Resulting number of filters in $\mathcal{S}_2$ & 14 \\
-                % $V$ & Wavelet scattering upsample factor & \textbf{2} \\
-                % $r$ & Downsampling factor of $\mathcal{U}_1$ due to maximum bandwidth & 8 \\
-                % $d$ & Downsampling factor of $\mathcal{S}_1$ & 5 \\
-                % $f_{ss}$ & Resulting scattering transform sample frequency & 6.25 Hz \\
-                % $B$ & Median filter length for noise coefficient suppression & \textbf{9} \\
-                % $D$ & Median filter length for transient coefficient suppression  & \textbf{5 seconds} \\
-                % $F$ & Average filter length for noise coefficient estimation  & \textbf{60 seconds} \\
-                % $R$ & Median filter length for stabilizing entropy  & \textbf{1.5 seconds} \\
-                % $G$ & Maximum filter length for boundary extension  & \textbf{6 seconds} \\
-                % $p_t$ & Probability threshold for the whitened \ac{se} \ac{gmm} detector & \textbf{0.3} \\ 
-                % $k$ & Dimensions in the classification feature vector $v$ & 521\\ 
-                % \end{tabular}
+            \renewcommand{\arraystretch}{0.8}% Tighter
+            \begin{tabular}{lll}  \small
+                \textbf{Parameter}      & \textbf{Description} & \textbf{Value} \\ \hline
+                $T$ & \ac{ws} invariance scale & \textbf{1.5 seconds} \\
+                $Q_1$ & Wavelets per octave in $\mathcal{S}_1$ & \textbf{16} \\
+                $Q_2$ & Wavelets per octave in $\mathcal{S}_2$ & \textbf{4} \\
+                $f_\text{min}$ & Minimum filterbank frequency in $\mathcal{S}_1$ & \textbf{15 Hz}\\
+                $f_\text{max}$ & Maximum filterbank frequency in $\mathcal{S}_1$ & \textbf{120 Hz}\\
+                $|\Lambda_1|$ & Resulting number of filters in $\mathcal{S}_1$ & 48 \\
+                $|\Lambda_2|$ & Resulting number of filters in $\mathcal{S}_2$ & 14 \\
+                $V$ & Wavelet scattering upsample factor & \textbf{2} \\
+                $r$ & Downsampling factor of $\mathcal{U}_1$ due to maximum bandwidth & 8 \\
+                $d$ & Downsampling factor of $\mathcal{S}_1$ & 5 \\
+                $f_{ss}$ & Resulting scattering transform sample frequency & 6.25 Hz \\
+                $B$ & Median filter length for noise coefficient suppression & \textbf{9} \\
+                $D$ & Median filter length for transient coefficient suppression  & \textbf{5 seconds} \\
+                $F$ & Average filter length for noise coefficient estimation  & \textbf{60 seconds} \\
+                $R$ & Median filter length for stabilizing entropy  & \textbf{1.5 seconds} \\
+                $G$ & Maximum filter length for boundary extension  & \textbf{6 seconds} \\
+                $p_t$ & Probability threshold for the whitened \ac{se} \ac{gmm} detector & \textbf{0.3} \\ 
+                $k$ & Dimensions in the classification feature vector $v$ & 521\\ 
+                \end{tabular}
                 % \end{ruledtabular}
-                % \end{table}
+                \end{table}
 

### 2024-05-06 dissertation/scripts/consolodate_refs.py

@@ -1,0 +1,41 @@
+import bibtexparser as bib
+
+with open('references.bib', 'r') as refs:
+    lib: bib.bibdatabase.BibDatabase = bib.load(refs)
+    
+entries = lib.get_entry_list()
+
+title = lambda x: x['title'].lower().replace("{", "").replace("}", "").strip()
+
+entry_consol = {}
+
+# for e in entries:
+#     print(e.keys())
+
+#check titles
+print('------------')
+for entry in entries:
+    k = title(entry)
+    if k in entry_consol.keys():
+        print(f"{entry['ID']} is a duplicate of {entry_consol[k]['ID']}")
+    else:
+        entry_consol[k] = entry
+print('------------')        
+#remove abracts
+for k, v in entry_consol.items():
+    if 'abstract' in v.keys():
+        del v["abstract"]
+        
+import pprint
+
+#sort according to authors
+entry_sorted = sorted(entry_consol.values(), key=lambda x: x['author'].lower().replace("{", "").replace("}", "").strip())
+
+# pprint.pprint(entry_sorted)
+
+sorted_lib = bib.bibdatabase.BibDatabase()
+sorted_lib.entries = entry_sorted
+
+with open('references-consolodated.bib', 'w') as file:
+    bib.dump(sorted_lib, file)
+    

### 2024-05-06 dissertation/content/paper2/main.tex
consolodated references

@@
 Underwater \ac{pam} systems are an effective non-invasive tool to monitor marine activity, with many research endeavours to automate cetacean monitoring and detection over the last decades.
-\Ac{ml} is a particularly prevalent approach to detecting and classifying cetacean calls from \ac{pam} audio \citep{MLreview}, although some recent publications remain focused on traditional approaches that seem to oppose \ac{ml} trends \citep{jacques1, jacques2}.
+\Ac{ml} is a particularly prevalent approach to detecting and classifying cetacean calls from \ac{pam} audio \citep{MLreview}, although some recent publications remain focused on traditional approaches that seem to oppose \ac{ml} trends \citep{internalsignaldetection, jacques2}.
 The disparity between the approaches are often due to a lack of data or a high uncertainty of the data quality \citep{casey2019}, which are detrimental to \ac{nn} and deep learning models.
@@
 They do not necessarily classify signals, and instead pinpoint \acp{poi}, which suffer from many false positive detections.
-On the other hand, very accurate \ac{nn} approaches exist for a variety of species \citep{alexnet_killer_whales, cnn_multiple_whale_classes, narw_cn_denoising, cnn_fin_whale}, but require large datasets and careful data preparation.
-
-Non-\ac{nn} approaches have also obtained very accurate results, most of which utilise \acp{mfcc} as features \citep{MFCC_HMM_birds, mfcc_hmm_brydes}.
+On the other hand, very accurate \ac{nn} approaches exist for a variety of species \citep{MLalexnet, cnn_multiple_whale_classes, narw_cn_denoising, cnn_fin_whale}, but require large datasets and careful data preparation.
+
+Non-\ac{nn} approaches have also obtained very accurate results, most of which utilise \acp{mfcc} as features \citep{MFCC_HMM_birds, hmm2}.
 A \ac{hmm} is deployed in the aforementioned studies, thereby allowing the modelling of time-dependent sound characteristics without requiring a \ac{nn}.
@@
 Automated classification is often the end goal of many \ac{pam} systems, but identifying \acp{poi} without classification is a necessary step to assist in the annotation and sound exploration process, especially when surveying new and unknown oceanic regions.
-As such, generalised detection is often performed with \acp{bled} and \ac{se} detectors \citep{se_dolphin}.
-\Ac{se} detectors have consistently shown better performance than energy detectors \citep{se_erbe_king}.
+As such, generalised detection is often performed with \acp{bled} and \ac{se} detectors \citep{entropyOCEANS}.
+\Ac{se} detectors have consistently shown better performance than energy detectors \citep{entropyJASA}.
 The previous approach by \citet{mypaper} modifies \ac{se} detectors to stabilize the entropy measure and provide pseudo-probabilistic output, thereby improving endpoint detection accuracy and the interpretability of the detector's output.
@@
 In Figure \ref{fig:det_calls}, we show that the proposed detector is able to detect the majority calls while detecting fewer noise.
-As illustrated in many previous studies, \ac{se} greatly outperforms a traditional \ac{bled} detector \citep{se_erbe_king}.
+As illustrated in many previous studies, \ac{se} greatly outperforms a traditional \ac{bled} detector \citep{entropyJASA}.
 Whitening is also shown to greatly improve the capabilities of the \ac{bled} detector.
@@
 Typical \ac{bled} with a fixed threshold performs very poorly due to varying noise conditions.
-\ac{bled} has repeatedly been shown to not be a reliable \ac{poi} detector \citep{jacques1, se_erbe_king}.
+\ac{bled} has repeatedly been shown to not be a reliable \ac{poi} detector \citep{internalsignaldetection, entropyJASA}.
 
@@
 
-% \begin{table}[h!]
-% \caption{Number of labeled detections for all vocalisations of interest.
+\begin{table}[h!]
+\caption{Number of labeled detections for all vocalisations of interest.
 The detections are produced by the whitened \ac{gmm} \ac{se} detector.
 A significant number of detections are discarded in the training data due to overlapping.}
-% \label{tab:numdetections}
+\label{tab:numdetections}
 % \begin{ruledtabular}
 % \renewcommand{\arraystretch}{0.8}
-% \begin{tabular}{lc} 
-% \textbf{Vocalisation}      & \textbf{Number of detections}  \\ \hline
-% Bm\_Ant\_A & 1514              \\
-% Bm\_Ant\_B & 474              \\
-% Bm\_Ant\_Z & 91               \\
-% Bm\_D      & 328               \\
-% Bp & 50                \\
-% Noise      & 25638          \\
-% Overlapping & 150 \footnotemark[1]      \\ 
-% \end{tabular}
+\begin{tabular}{lc} 
+\textbf{Vocalisation}      & \textbf{Number of detections}  \\ \hline
+Bm\_Ant\_A & 1514              \\
+Bm\_Ant\_B & 474              \\
+Bm\_Ant\_Z & 91               \\
+Bm\_D      & 328               \\
+Bp & 50                \\
+Noise      & 25638          \\
+Overlapping & 150 \footnotemark[1]      \\ 
+\end{tabular}
 % \end{ruledtabular}
-% \footnotetext[1]{Overlapping detections contain at least 2 overlapping annotations of different classes.
+\footnotetext[1]{Overlapping detections contain at least 2 overlapping annotations of different classes.
 This implies that more than 300 vocalisations are discarded as a result of overlap, thereby drastically reducing training data for the \textit{Bm\_D} and \textit{Bp\_20Plus} classes, which have a higher tendency for overlap.}
-% \end{table}
+\end{table}
 
@@
 
-Our disposal of the time dimension by using the maximum naturally lends our classification model being replaced with a \ac{hmm} to incorporate time-dependent relationships similar to that of \citet{mfcc_hmm_brydes} and \citet{dmd_hmm}.
+Our disposal of the time dimension by using the maximum naturally lends our classification model being replaced with a \ac{hmm} to incorporate time-dependent relationships similar to that of \citet{hmm2} and \citet{hmm1}.
 However, instead of classifying on a frame-by-frame basis such as in \citet{otherScattering, ws_audio, ws_ecg}, we utilise the scattering coefficients to first detect a variable-length segment of audio which may contain whale vocalisations, thereby reducing the amount of data in the noise class.

### 2024-05-06 dissertation/content/paper3/main.tex
consolodated references

@@
 
-Various forms of \ac{ws} exist, with \ac{1d} scattering first proposed by Anden and Mallat \citep{1dscattering1, 1dscattering2}, which was later extended to 2 \citep{2dscattering} and 3 dimensions \citep{3dscattering, harmonicscattering}.
-Additional and more specialised forms for \ac{ws} include rotation-invariant scattering \citep{groupinvariantscattering} for the classification of textures and joint-\ac{tf} scattering \citep{jointtfscattering1, jointtfscattering2}.
+Various forms of \ac{ws} exist, with \ac{1d} scattering first proposed by Anden and Mallat \citep{1dscattering1, ws}, which was later extended to 2 \citep{2dscattering} and 3 dimensions \citep{3dscattering, harmonicscattering}.
+Additional and more specialised forms for \ac{ws} include rotation-invariant scattering \citep{groupinvariantscattering} for the classification of textures and joint-\ac{tf} scattering \citep{ws_joint_tf, jointtfscattering2}.
 Joint-\ac{tf} scattering utilises a separable \ac{2d} filter that operates on the first level scattering scalogram, denoted by the operator $\mathcal{U}_1$, which is the only apparent usage of separable filters in the current scattering literature.
@@
 Additionally, separable filters reduce the number of parameters of a \ac{cnn} if the filters are configured to be learnable.
-Learnable filters can typically improve performance compared to their fixed-filters counterpart \citep{learnablewavelettransform}.
+Learnable filters can typically improve performance compared to their fixed-filters counterpart \citep{scattering_birdsong}.
 

### 2024-05-06 dissertation/content/paper3/original/main.tex
consolodated references

@@
 
-Various forms of \ac{ws} exist, with \ac{1d} scattering first proposed by Anden and Mallat \cite{1dscattering1, 1dscattering2}, which was later extended to 2 \cite{2dscattering} and 3 dimensions \cite{3dscattering, harmonicscattering}.
-Additional and more specialised forms for \ac{ws} include rotation-invariant scattering \cite{groupinvariantscattering} for the classification of textures and joint-\ac{tf} scattering \cite{jointtfscattering1, jointtfscattering2}.
+Various forms of \ac{ws} exist, with \ac{1d} scattering first proposed by Anden and Mallat \cite{1dscattering1, ws}, which was later extended to 2 \cite{2dscattering} and 3 dimensions \cite{3dscattering, harmonicscattering}.
+Additional and more specialised forms for \ac{ws} include rotation-invariant scattering \cite{groupinvariantscattering} for the classification of textures and joint-\ac{tf} scattering \cite{ws_joint_tf, jointtfscattering2}.
 Joint-\ac{tf} scattering utilises a separable \ac{2d} filter that operates on the first level scattering scalogram, denoted by the operator $\mathcal{U}_1$, which is the only apparent usage of separable filters in the current scattering literature.
@@
 Additionally, separable filters reduce the number of parameters of a \ac{cnn} if the filters are configured to be learnable.
-Learnable filters can typically improve performance compared to their fixed-filters counterpart \cite{learnablewavelettransform}.
+Learnable filters can typically improve performance compared to their fixed-filters counterpart \cite{scattering_birdsong}.
 

### 2024-05-06 dissertation/scripts/consolodate_refs.py

@@ -21,10 +21,11 @@ for entry in entries:
     else:
         entry_consol[k] = entry
 print('------------')        
-#remove abracts
+#remove abracts and add title brackets
 for k, v in entry_consol.items():
     if 'abstract' in v.keys():
         del v["abstract"]
+    v['title'] = "{" + v['title'] + "}"
         
 import pprint
 
@@ -36,6 +37,6 @@ entry_sorted = sorted(entry_consol.values(), key=lambda x: x['author'].lower().r
 sorted_lib = bib.bibdatabase.BibDatabase()
 sorted_lib.entries = entry_sorted
 
-with open('references-consolodated.bib', 'w') as file:
+with open('references_consolodated.bib', 'w') as file:
     bib.dump(sorted_lib, file)
     

### 2024-05-14 dissertation/content/nomenclature.tex
started intro to p3

@@
 \begin{tabular}{p{0.15\textwidth} p{0.75\textwidth}}
-    $X[k,m]$ & short-time Fourier transform of $x$, with $k$ the frequency index and $m$ the time index \\
+    $X[k,m]$ & a time-frequency decomposition of $x$, with $k$ the frequency index and $m$ the time index \\
     $S[k,m]$ & power spectrum calculated from $X[k, m]$ \\
@@
     $\text{RA}^F_m (x)$ & rolling average filter operator with a window of size $F$ operating on index variable $m$ on a discrete function $x$\\
+    $x \triangleq y$ & $x$ is defined as $y$\\
 \end{tabular}
@@
     GMM & Gaussian mixture model \\
+    DT & discrete-time \\
     POI & point(s) of interest \\
@@
     STNR & soft true negative ratio \\
+    BPF & band-pass filter \\
 

### 2024-05-14 dissertation/content/p1_intro.tex
before reference consolodation
started intro to p3

@@
-\chapter{The Relevance of Time-Frequency Analyses}
+\acrodef{dt}[DT]{discrete-time}
+\acrodef{ct}[CT]{continuous-time}
+\acrodef{bpf}[BPF]{band-pass filter}
+\acrodef{dtft}[DTFT]{discrete-time Fourier transform}
+
+\chapter{Time-Frequency Analyses}
+\label{chap:p1i}
+
+A \ac{tf} analysis is one of the most important tools in audio classification, data preparation and data manipulation.
+The main idea behind \ac{tf} decompositions is to split a signal into frequency bands, which can viewed as time-varying functions.
+
+Formally, given a \ac{1d} \ac{dt} signal $x[n]$, with an index variable $n$, a \ac{tf} analysis decomposes $x$ as 
+\begin{equation}
+    X[k, n] = \sum_i x[n] \upsilon_k[i - n] = x * \upsilon_k,
+\end{equation}
+where $\upsilon_k \in \Upsilon$ is the $k$'th \ac{bpf} in the decomposition filter dictionary $\Upsilon$. Note that a \ac{lpf} can be considered to be a \ac{bpf} with a centre-frequency of 0.
+
+Generally, the entire convolution $X[n,k]$ is not computed.
+Instead, only select samples are computed by speficying a hop-distance or downsample factor $d$.
+We indicate this by tranforming the index variable $m = dn$ as
+\begin{equation}
+    \label{eqn:gentf}
+    X[k, m] \triangleq X[k, dn] = \left(x * \upsilon_k\right)_{\downarrow d}.
+\end{equation}
+
+The is equivalent to applying a downsample operator $\left(\cdot\right)_{\downarrow d}$ to the convolution.
+
+\section{Fourier Transforms}
+
+A \ac{ct} Fourier transform of $x(t)$ is given as 
+\begin{equation}
+    \mathcal{F} \{x\} \triangleq \hat{x}(\omega) = \int_{-\infty}^{\infty} x(t) e^{-j\omega t} dt = \left(x * e^{j\omega t}\right)\bigg|_{t=0}.
+\end{equation}
+
+Derived from the conventional Fourier transform, a \ac{dt} signal $x[n], n \in \{0, ..., N-1\}$ (length $N$) has a \ac{dtft}
+\begin{equation}
+    \label{eqn:dtft}
+    \text{DTFT}\{x\} \triangleq \hat{x}[k] = \sum_{n=0}^{N} x(t) e^{-j \frac{2\pi k n}{N}} = \left(x \otimes \upsilon_k\right)\bigg|_{n=0},
+\end{equation}
+for which $\upsilon_k = e^{j \frac{2\pi k n}{N}}, \ k \in \{0, ..., N-1\}$ forms an orthogonal basis of $x$.
+The circular convolution operator $\otimes$ implies the infinite periodic extension of $x$ beyond its boundaries of $n \in \{0, ..., N-1\}$.
+
+Equation (\ref{eqn:dtft}) describes the \ac{dtft} as convolution evaluated at the index 0 as a useful link to the generalised notion of \ac{tf} analyses: Fourier transforms can be considered as complex sinusoidal filters being applied to the signal of interest, with periodic boundary extension.
+
+The \ac{dtft} is commonly used to quickly compute convolutions for which the \ac{fft} is a fast alorithm with $N \log N$ computational complexity.
+In particular, to circularly convolve signals $x$ and $y$, a fast computation is implemented as
+\begin{equation}
+    \label{eqn:fftconv}
+    x \otimes y = \text{FFT}^{-1}\{\hat{x} \cdot \hat{y}\}.
+\end{equation}
+
+With the correct padding procedures, one can compute $x * y$ using (\ref{eqn:fftconv}).
+
+Fourier transforms as decribed above are invertible operators, since they are orthogonal decompositions.
+However, for \ac{tf} analyses, invertibility is not necessarily a requirement of the application.
+As such, this disseration does not consider invertibility for the remainder of the content.
+
+\section{Constructing Filters}
+\label{sec:p1i:tfa}
+
+The most popular \ac{tf} analysis is the \ac{stft}.
+
+In terms of equation (\ref{eqn:gentf}), a \ac{stft} is computed with 
+\begin{equation}
+    \label{eqn:stftfilter}
+    \upsilon_k[n] =  w[n] e^{j \frac{2\pi k n}{N}},
+\end{equation}
+where $w$ is a window function defined by the user.
+Various \acp{fft} are use to compute this transform, so the properties of $w$ is typically specified by the \ac{fft} length, and the window type (Hann, Hamming, Blackman, etc.).
+The hop-length (or downsample factor) $d$ is an additional parameter specified by the user.
+
+The frequency domain content $\hat{w}$ entirely specifies the filter behaviour of $\upsilon_k$, since the factor $e^{j \frac{2\pi k n}{N}}$ only serves to specify the centre frequency of $\upsilon_k$.
+The resulting filter dictionary $\Upsilon$ is therefore comprised of constant-bandwidth analytic (complex) filters, linearly spaced in the frequency domain.
+As such, depending on the definition of filter bandwidth, there exists an optimal $d$ which captures all of the necessary information of $|X[k, m]|$ such that no aliasing occurs across the index dimension $m$.
+This type of optimality is discussed, at length, for wavelet filterbanks in chapter \ref{chap:p3}.
+
+The linearly-spaced filters and constant-bandwidth does not necessarily reflect and/or capture the natural phenomena typically present in audio signals.
+In particular, an audio signal can be expressed as a series of harmonics above its fundamental.
+This motivates the use of exponentially placed filters.
+
+Formally, given a signal with a fundamental frequency most accurately captured in the frequency bin $k_0$, any change in frequency by a factor $a$ will move the fundamental frequency content to the bin $a k_0$.
+This has dire consequences for lower frequency signals - the frequency resolution of lower frequency signals is significantly poorer compared to the frequency resolution of its own harmonics.
+As such, a small frequency shift of a lower frequency signal with ample harmonic content which cause large index shifts for its harmonics, i.e., the $b$'th harmonic will be shifted to index $a b k_0$.
+
+Suppose that we design $\Upsilon$ to contain $Q$ filters per octave.
+A frequency change by a factor $a$ shifts the frequency index of the fundamental to $(Q\log_2 (a) + 1) k_0$.
+Similarly, the $b$'th harmonic's frequency index is shifted to $(Q\log_2 (a) + 1) (Q\log_2 (b) + 1) k_0$.
+For example, if $a=2$, the fundamental's frequency index is shifted to $k_0 + Q$ (as opposed to $2 k_0$ for the \ac{stft}).
+This property is referred to a frequency-shift invariance, since small frequency deformations of the order $Q$ do not significantly change the frequency index structure of the \ac{tf} decomposition.
+The \ac{stft} is much more sensitive to such deformations across its entire structure for harmonically rich signals.
+
+An exponential filterbank design has increased stability to small frequency fluctuations, which can make it more useful for feature-extraction methods in a \ac{ml} pipeline, since feature vectors are more stable against frequency deformations.
+The exponential filterbank is, in fact, the way wavelet filters are typically constructed.
+Additionally, the \ac{mfcc} filterbank attributes much of its sucess (compared to raw \ac{stft} magnitude coefficients) due to its exponentially placed filters.
+
+\section{Wavelet Filters}
+\label{sec:p1i:wavelets}
+
+To construct a wavelet filterbank with $\upsilon_k \in \Upsilon$ which has the exponentially-spaced frequency properties discussed in section \ref{sec:p1i:tfa}, we use a ``template'' wavelet $\uppsi(t)$, referred to as the mother wavelet.
+Conventionally, wavelet filters are defined as \ac{ct} filters and then discretised in software implementations.
+
+The mother wavelet is a \ac{bpf} with the following properties:
+\begin{enumerate}
+    \item Centre frequency of 1 rad/s.
+    \item Zero mean: $\hat{\uppsi}(0) = 0$.
+    \item Bandwidth defined by the user.
+\end{enumerate}
+
+$uppsi$ can the be scaled by a factor $\lambda$ such that
+\begin{equation}
+    \upsilon_k(t) = \lambda_k \uppsi(\lambda_k t).
+\end{equation}
+
+Notational shorthand is often used: 
+\begin{equation}
+    \uppsi_{\lambda_k}(t) \triangleq \lambda_k \uppsi(\lambda_k t).
+\end{equation}
+
+We construct a set $\lambda_k \in \Lambda$ such that $\lambda_{k+1} = 2^\frac{1}{Q}\lambda_{k}$, which ensures $Q$ wavelets per octave.
+Since $\uppsi$ has a frequency of 1 rad/s, $\lambda_k$ becomes the centre frequency of $\uppsi_{\lambda_k}$, while also scaling its bandwidth by a factor of $\lambda_k$.
+
+An additional advantage resulting from bandwidth scaling is that higher frequencies become more localised in time, allowing for better localisation precision as frequency increases.
+This results from the fact that $\uppsi_{\lambda_k}$ maintains the same number of oscillations over scale, which is not true for the \ac{stft}.
+
+Chapter \ref{chap:p1} discusses the applications of such a filterbank in terms of the \ac{cwt}, obtaining a scalogram.
+Note that traditional \ac{cwt} notation typically uses the scale variable $s = \frac{1}{\lambda}$, where $s$ is made continuous.
+Discretisations of the \ac{cwt} approximate the transform by limiting $s$ to the set $s_k \in \frac{1}{\lambda_k}, \ \lambda_k \in \Lambda$.
+Chapter \ref{chap:p1} specifically considers the application of the \ac{cwt} for signal detectors, while also using other techniques to improve traditional methods.

### 2024-05-14 dissertation/content/p2_intro.tex
started intro to p3

@@
+\acrodef{dct}[DCT]{discrete cosine transform}
+
 \chapter{Feature Extraction with Wavelet Transforms}
+\label{chap:p2i}
+
+\ac{tf} decompositions are often used to extract features for signals.
+In audio, \acp{mfcc} are features calculated from the \ac{stft} of the signal.
+A ``\ac{tf}'' decomposition, as referred to in this dissertation, does not necesarrily require a time variable: spatial variables are often used.
+In this sense, a \ac{cnn} is can also be considered as a ``\ac{tf}'' decomposition, where multi-dimensional index variables (or index vectors) are used for ``time'' and frequency:
+\begin{equation}
+    X[\vect{k}, \vect{m}] = \left( x * \upsilon_{\vect{k}} \right)_{\downarrow \vect{d}}, \ \upsilon_{\vect{k}} \in \Upsilon.
+\end{equation}
+
+In this case, we abuse notation to indicate that the operator $(\cdot)_{\downarrow \vect{d}}$ downsamples by different amounts for each dimension of $\vect{m}$.
+
+The downsampling factor $\vect{d}$ limits the bandwidth of $X$, which, as discussed in chapter \ref{chap:p1i}, is selected so as to retain all information carried by the filters in $\Upsilon$.
+Although not notated, different filters (different $\vect{k}$) may have different values of $\vect{d}$.
+Notation to support such a case is neglected for clarity, but is later expanded upon in chapter \ref{chap:p3}.
+
+\section{The Importance of Invariance}
+
+The decomposition $X$ is often further processed by additional filters $\Phi_i[\vect{k}, \vect{m}]$, where $i$ is the index of the post-processing filter:
+\begin{equation}
+    \tilde{X_i}[\vect{k}, \vect{m}] = X * \Phi_i.
+\end{equation}
+The filters $\Phi_i$ are not necesarrily multi-dimensional, and may only operate along a single dimension of $X$.
+If $\Phi_i$ is an averaging (low-pass) filter, some additional stability can be observed in $\tilde{X_i}$ compared with $X$.
+This allows for the introduction of invariances to various deformations applied to $X$, which can be greatly beneficial for some applications.
+In a sense, cascased structures of $X$ and $\tilde{X_i}$ are defined in a \ac{cnn} for which the network learns the required invariances.
+
+It is widely known that enforcing invariance in feature extraction and \ac{ml} pipelines increase performance.
+Enforcing invariance in \acp{nn} has been shown to improve the network, while also improving the interpretability of the network.
+For example, when classifying images, invariance to scale, shear, rotation and translation deformations, depending on image properties, may be useful when extracting features.
+Specifically, deformations of a specified order should approximately map to the same point in feature space:
+\begin{equation}
+    \Gamma\left(\delta\left(x\left[\vect{n}\right]\right)\right) \approx \Gamma\left(x\left[\vect{n}\right]\right),
+\end{equation}
+where $\Gamma$ is the feature extraction operator, and $\delta$ is the deformation operator applied to $x$.
+
+Invariance filters ($\Phi_i$) can also have additional properties that makes it favourable for audio applications.
+For example, many applications ``blur'' (low-pass) the \ac{stft} magnitude or power spectrum, which can reduce the impact of noise fluctuations, and can also serve to suppress small frequency fluctuations across adjascent frequency bins (if the blurring kernel $\Phi_i$ filters across frequency as well).
+The blurring operation can improve the performance of signal detectors and classifiers.
+However, as a cost, time localisation and/or frequency localisation is reduced.
+
+A new problem is introduced when enforcing invariances via the $\Phi_i$ filters - information is lost via averaging.
+Various techniques can be used to further retain information, while also retaining invariance.
+Wavelet scattering is such a method.
+In a sense, multilayer \acp{cnn} topologies with skip connections can be though of as a method to recover lost information in the deeper layers.
+
+\section{MFCCs Reframed as Shift-Invariant Features}
+
+Since \acp{mfcc} are the most widely used feature extraction method for audio signals, save for \ac{cnn} frontends which also act as a feature extractor when operating on the \ac{stft}.
+For this reframing, we ignore the \ac{dct} applied to the filtered values, since this linear transformation only serves to decorrelate the filter coefficients.
+
+The magnitude spectrum used in \ac{mfcc} calculation is expressed as a \ac{tf} decomposition as defined by equations (\ref*{eqn:gentf}) and (\ref*{eqn:stftfilter}):
+\begin{equation}
+    X[l, m] = |\upsilon_l * x|_{\downarrow d},
+\end{equation}
+where $\upsilon_l$ is the $l$'th \ac{stft} filter.
+
+The Mel spectrogram \ac{tf} decomposition $X_\text{Mel}$ is then calculated as
+\begin{equation}
+    X_\text{Mel}[k, m] = \log\left(X[l, m] * \Phi_k[l]\right) \bigg|_{l = 0},
+\end{equation}
+where $\Phi_k$ is the $k$'th MFFC triangular filter operating along the STFT index $l$. The evaluation at $l=0$ indicates that this operation is a multiplication and summation only.
+
+We can view this process as a \ac{tf} decompistion with STFT filters $\upsilon_l \in \Upsilon$, which is modified with the Mel-scale triangular filters $\Phi_k$.
+$\upsilon_l$ provides time-shift invariance due to the STFT window, whereas $\Phi_k$ selectively averages frequency content of the STFT, thereby providing some frequency-shift invariance.
+
+\section{Wavelets as a MFCC Generalisation}
+
+We can construct a wavelet filterbank which, similar to MFCCs, introduces time and frequency-shift invariance in a similar manner to the Mel spectrogram.
+This process is constructed in the opposite order (frequency-shift, then time-shift invariance) to the Mel spectrogram.
+However, its properties remain similar \cite{ws}.
+
+We construct wavelet band-pass filters $\upsilon_k \in \Upsilon$ which is spaced in frequency such that the desired invariance properties is obtained (as discussed in section \ref*{sec:p1i:wavelets}).
+We then utilise a single low-pass filter $\Phi[\vect{m}]$ which then provides invariance in the time and/or spatial dimensions:
+\begin{gather}
+    \label{eqn:general_scalogram}
+    X[\vect{k}, \vect{m}] = \left|x * \upsilon_{\vect{k}}\right|_{\downarrow \vect{d}}, \ \upsilon_{\vect{k}} \in \Upsilon; \\
+    \label{eqn:general_scattering}
+    \tilde{X}[\vect{k}, \vect{m}] = X[\vect{k}, \vect{m}] * \Phi[\vect{m}]
+\end{gather}
+
+The notation in equations (\ref*{eqn:general_scalogram}) and (\ref*{eqn:general_scattering}) conforms the general \ac{tf} description of chapter \ref*{chap:p1i}, although the specific wavelet scattering operators in chapters \ref*{chap:p2} to \ref*{chap:p3} follow a different standard.
+These operators are known as wavelet scattering, which formalise the notation of ``averaging a \ac{tf}-decomposition'' in terms of digital signal processing.
+Chapters \ref*{chap:p2} to \ref*{chap:p3} discuss the specifics, flavours and implementation of scattering operators at length.

### 2024-05-14 dissertation/content/p3_intro.tex
started intro to p3

@@
 \chapter{Wavelet Scattering in Higher Dimensions}
+\label{chap:p3i}
+
+In chapter \ref*{chap:p2i}, we reviewed a generalised definition of a multi-dimensional TF decomposition $X$, which is processed by a set of invariance-enforcing filters $\Phi_i$.
+Wavelet scattering in 1 dimension utilising a single invariance-enforcing filter, $\phi$, as discussed in chapter \ref*{chap:p2}.
+
+Wavelet scattering filterbanks have been extended to multiple dimensions using various different wavelets and filterbank construction techniques.
+This chapter provides an overview of the existing filterbanks and filters and their properties.
+We then identify a gap in the literature, which gives rise to a generalisation of cascasded 1D scattering transforms -- separable wavelet scattering.
+
+\section{Types of Scattering Transforms}
+Existing scattering transforms are listed below convenience:
+\begin{enumerate}
+    \item 1D scattering \citep{1dscattering1, ws}.
+This is the first-introduced scattering transform, which serves as a generalisation of MFCCs.
+It improves on MFCCs by recovering the information lost due to invariance filter (averaging) operations.
+Morlet wavelets are used, with cascaded scattering levels having its own number of filter per octave.
+    \item 2D scattering \citep{2dscattering}.
+Gabor filters modified to 0-mean, are used to construct the filterbank.
+The filterbank consists of exponentially placed filters in one axis, which is then rotated by equally spaced angles to cover the frequency plane.
+This filterbank only has a single filter per octave ($Q=1$).
+In particular, in polar coordinates, filters are placed exponentially along $r$, but linearly along $\theta$.
+    \item 3D scattering -- harmonic scattering \citep{harmonicscattering}.
+The spherical harmonics are used as filters which consitute an orthogonal basis in 3 dimensions.
+These filters have ``lobes'' that extend in a multitude of dimensions, depending on the harmonic, in the frequency domain.
+    \item 3D scattering -- extension of 2D scattering \citep{3dscattering}.
+The filterbank is constructed similarly to 2D scattering, where Gabor filters are spaced with linearly varying angles (in spherical coordinates) in 3 dimensions.
+    \item Joint time-frequency \citep{ws_joint_tf,jointtfscattering2} -- extending 1D to 2D.
+A first level of scattering in performed in 1D to produce a scalogram ($X[k, m]$).
+2D scattering is then performed on subsequent levels, directly on the scalogram.
+The 2D wavelets are separable - a product of 1D wavelets.
+This allows 1D scattering on audio to have features that are also dependent on adjascent frequency bins.
+It is conceptually similar to a \ac{cnn} operating on a \ac{stft} spectrogram.
+\end{enumerate}
+
+\section{A Detailed view of Joint Time-Frequency Scattering}
+
+\section{The Benefits of Generalising Joint Time-Frequency}

### 2024-05-14 dissertation/content/p3_intro.tex
section updates in p3i

@@
 
+Previous chapters have been specifically targeted at underwater bioacoustic applications.
+We deviate from specific applications in chapter \ref*{chap:p3} in order to provide a generalised definition of a new type of scattering.
+This is considered to be the main and largest contribution of this dissertation.
+The new proposed scattering is intended for use in multiple applications which can benefit from filter separability properties (see section \ref*{sec:p3i:sep}).
+
 Wavelet scattering filterbanks have been extended to multiple dimensions using various different wavelets and filterbank construction techniques.
@@
 \section{Types of Scattering Transforms}
-Existing scattering transforms are listed below convenience:
+Existing scattering transforms are listed and described below for convenience:
 \begin{enumerate}
@@
 This filterbank only has a single filter per octave ($Q=1$).
-In particular, in polar coordinates, filters are placed exponentially along $r$, but linearly along $\theta$.
-    \item 3D scattering -- harmonic scattering \citep{harmonicscattering}.
+In particular, in polar coordinates $(r, \theta)$ in the frequency domain, filters are placed exponentially along $r$, but linearly along $\theta$.
+    \item Harmonic scattering \citep{harmonicscattering} (3 dimensions).
 The spherical harmonics are used as filters which consitute an orthogonal basis in 3 dimensions.
@@
     \item 3D scattering -- extension of 2D scattering \citep{3dscattering}.
-The filterbank is constructed similarly to 2D scattering, where Gabor filters are spaced with linearly varying angles (in spherical coordinates) in 3 dimensions.
-    \item Joint time-frequency \citep{ws_joint_tf,jointtfscattering2} -- extending 1D to 2D.
+The filterbank is constructed similar to 2D scattering.
+In spherical coordinates $(\rho, \theta, \phi)$ in the frequency domain, filters are spaced exponentially along $\rho$, and linearly along $\theta$ and $\phi$.
+    \item Joint time-frequency \citep{ws_joint_tf,jointtfscattering2} scattering -- 2D filters on 1D scattering scalogram.
 A first level of scattering in performed in 1D to produce a scalogram ($X[k, m]$).
-2D scattering is then performed on subsequent levels, directly on the scalogram.
+2D scattering is then performed on subsequent levels, with $X[k, m]$ as input.
 The 2D wavelets are separable - a product of 1D wavelets.
@@
 It is conceptually similar to a \ac{cnn} operating on a \ac{stft} spectrogram.
+In the frequency domain with axes $(\omega_1, \omega_2)$, the 2D filters are constructed as the cross-product between filter sets that placed exponentially in both $\omega_1$ and $\omega_2$.
+    This is the only literature considering separable filters.
 \end{enumerate}
 
-\section{A Detailed view of Joint Time-Frequency Scattering}
-
-\section{The Benefits of Generalising Joint Time-Frequency}
+\section{Benefits of Separable Filters}
+\label{sec:p3i:sep}

### 2024-05-18 dissertation/content/introduction.tex
started intro

\chapter{Introduction}

\Ac{ml} models have become mainstay in software products and automation pipelines. Often, when there is processing power available, a \ac{ml} model will be deployed to increase efficiency, accuracy and limit human interaction. Various models, whether they are statistical, Bayesian or probabilistic, decision trees or \ac{nn}, are configured to assist or perform various tasks across many industries.

Modern research is very focussed on advancing \ac{ml} integration, improving model performance and investigating new techniques. A large branch of this research is particularly focussed on data-driven approaches using \acp{nn}. The collection of data therefore becomes an important part of model training and deployment. \Acp{nn} present some challenges and concerns, since they require vast amounts of data, and their operation is often unpredictable and not well understood. They are typically treated as ``black box'' functions, which researchers improve and design through trial and error via hyper-parameter optimisation and architecture exploration.

In the field of \ac{pam}, which focuses on monitoring wildlife using non-invasive acoustic sensors, the majority of recent \ac{sota} models are \ac{nn}-based. However, metrics of real-life deployment and practical model evaluation are often ignored, since the acoustical landscape of natural environments can be surprisingly varied. For example, one cannot expect a \ac{nn} model which was not trained on data that contains rainfall noise, to effectively operate when it starts raining.

Long moments of silence, weather patterns and migrating animals contribute to an ever-changing soundscape which can cause many false positives in naively trained and deployed models. Systems which aim to track specific species are often faced with this ``needle in a haystack'' problem, in which the sounds you are looking for are rare compared to the lifetime of the deployed sensors. As such, human intervention with recorded audio is a must-have when trying to reliably track species via audio signals, which can be a very laborious task indeed.

As such, designing models and systems which are robust to all the aforementioned factors are critical in automating wildlife conservation studies based on non-invasive audio sensors. This leads to the natural question of the suitability of \acp{nn} in this field and to what extent they can be used to be effective.

At the core of many audio-based systems lies a \ac{tf}-decomposition, typically \acp{mfcc} or the \ac{stft}. This dissertation focuses on \ac{tf} decompositions, and how they can be leveraged in order to reduce the usage or even entirely replace \acp{nn} in \ac{pam} systems. At the end of this investigative journey, we naturally arrive an entirely new generalisation of a feature extraction method based on \ac{ct} wavelets, which can be utilised as a drop-in replacement for a \ac{cnn} front-end. The integration and investigation of wavelet-based processing in \ac{pam} systems are relatively scarce, even though it is very prevalent in other fields, which we aim to remedy in the publications presented in this dissertation.

\section{Problem Statement}

In summary, some of the many challenges posed by \ac{pam} systems are:
\begin{enumerate}
    \item Long periods of silence or noise, leading to unbalanced data classes which can have a tendency for creating many false positives in models if not trained correctly.
    \item Constantly evolving noise landscape, with the interference of other bioacoustical sounds which we are often not interested in classifying.
    \item Large audio databases and recordings, which motivates the use of fast algorithms and processing techniques.
    \item Poor \ac{snr}, often less than 0 dB.
    \item Incomplete or mislabeled training data, due to poor \ac{snr} and human error.
    \item Small datasets, which only cover a portion of environment variability, making it unsuitable for a general model.
    \item Non-white, non-stationary background noise, which makes classification and signal detection non-trivial tasks.
    \item Poorly understood species behaviour in terms of their bioacoustics and movements.
\end{enumerate}

All the above factors seem demotivate the use of \acp{nn} for the following reasons: principle of operation can be unpredictable and not well understood; can take long to compute; requires reliable training data or special techniques to mitigate problematic data. Although it is possible to address these challenges, it is worth exploring non-\ac{nn} methods.

As a reponse to the challenges posed by \ac{pam} systems, we focus specifically on alternative aproaches to \acp{nn}, by considering the modification of traditional classifiers and detectors with wavelet-based \ac{tf}-decompositions.

\section{Layout}

As this is a dissertation by publication, each paper can be read as a stand-alone item, with all the necessary literature and knowledge capture within. As such, this dissertation is structured with additional contextual chapters (chapters \ref*{chap:p1i}, \ref*{chap:p2i}, \ref*{chap:p3i}) prior to each publication (chapters \ref*{chap:p1}, \ref*{chap:p2}, \ref*{chap:p3}).

Each contextual chapter is in itself an introduction to the literature and concepts presented by the paper in the following paper. Contextual chapters serve to unify the content of the published papers and show the natural progression of the analysis and application of \ac{tf} decompositions. All relevant literature is therefore gradually introduced, due to each paper conceptually expanding on the previous paper's ideas and contributions.

Although this dissertation is mostly focussed on the application to \ac{pam} systems, the theory and findings it presents are general. As such, the contextual chapters are generalised mathematical descriptions of the concepts present in each paper, whereas the papers 1 and 2 (chapters \ref*{chap:p1}, \ref*{chap:p2}) are focussed on applying it to \ac{pam} systems. The final paper (chapter \ref*{chap:p3}) shows a generalised result from all the knowledge and investigations presented in prior papers.

Each paper is presented as-is, with no modifications, save for unifying mathematical notation and the renumbering of items.

\section{Journal Papers and Contributions}

### 2024-05-18 dissertation/content/p1_intro.tex
started intro

@@
 
-The most popular \ac{tf} analysis is the \ac{stft}.
-
-In terms of equation (\ref{eqn:gentf}), a \ac{stft} is computed with 
+The most popular \ac{tf} analysis is the \ac{stft}. In terms of equation (\ref{eqn:gentf}), a \ac{stft} is computed with 
 \begin{equation}
@@
 The hop-length (or downsample factor) $d$ is an additional parameter specified by the user.
+It is possible to choose $w$ and $d$ such that the \ac{stft} becomes an invertible operator -- the original signal can be entirely recovered.
 
@@
 For example, if $a=2$, the fundamental's frequency index is shifted to $k_0 + Q$ (as opposed to $2 k_0$ for the \ac{stft}).
-This property is referred to a frequency-shift invariance, since small frequency deformations of the order $Q$ do not significantly change the frequency index structure of the \ac{tf} decomposition.
+This property is referred to a frequency-shift invariance, since small frequency deformations of the order $Q^{-1}$ do not significantly change the frequency index structure of the \ac{tf} decomposition.
 The \ac{stft} is much more sensitive to such deformations across its entire structure for harmonically rich signals.
@@
 
-$uppsi$ can the be scaled by a factor $\lambda$ such that
+$\uppsi$ can the be scaled by a factor $\lambda$ such that
 \begin{equation}

### 2024-05-18 dissertation/content/p2_intro.tex
started intro

@@
 
-We can construct a wavelet filterbank which, similar to MFCCs, introduces time and frequency-shift invariance in a similar manner to the Mel spectrogram.
+We can construct a wavelet filterbank which introduces time and frequency-shift invariance in a similar manner to the Mel spectrogram.
 This process is constructed in the opposite order (frequency-shift, then time-shift invariance) to the Mel spectrogram.
@@
 The notation in equations (\ref*{eqn:general_scalogram}) and (\ref*{eqn:general_scattering}) conforms the general \ac{tf} description of chapter \ref*{chap:p1i}, although the specific wavelet scattering operators in chapters \ref*{chap:p2} to \ref*{chap:p3} follow a different standard.
-These operators are known as wavelet scattering, which formalise the notation of ``averaging a \ac{tf}-decomposition'' in terms of digital signal processing.
+These operators are known as wavelet scattering, which formalise the notion of ``averaging a \ac{tf}-decomposition'' in terms of signal processing theory.
 Chapters \ref*{chap:p2} to \ref*{chap:p3} discuss the specifics, flavours and implementation of scattering operators at length.

### 2024-05-18 dissertation/content/p3_intro.tex
started intro

@@
 The filterbank consists of exponentially placed filters in one axis, which is then rotated by equally spaced angles to cover the frequency plane.
-This filterbank only has a single filter per octave ($Q=1$).
-In particular, in polar coordinates $(r, \theta)$ in the frequency domain, filters are placed exponentially along $r$, but linearly along $\theta$.
+In particular, in polar coordinates $(r, \theta)$ in the frequency domain, filters are placed exponentially along $r$ with a single filter per octave, but linearly along $\theta$.
     \item Harmonic scattering \citep{harmonicscattering} (3 dimensions).
@@
 The filterbank is constructed similar to 2D scattering.
-In spherical coordinates $(\rho, \theta, \phi)$ in the frequency domain, filters are spaced exponentially along $\rho$, and linearly along $\theta$ and $\phi$.
+In spherical coordinates $(\rho, \theta, \phi)$ in the frequency domain, filters are spaced exponentially along $\rho$ with a single filter per octave, and linearly along $\theta$ and $\phi$.
     \item Joint time-frequency \citep{ws_joint_tf,jointtfscattering2} scattering -- 2D filters on 1D scattering scalogram.
@@
 \label{sec:p3i:sep}
+
+Recently, wavelet scattering operations implemented as \acp{cnn}, referred to as a scattering network, have been of increasing interest in order to increase the effectiveness of \ac{cnn} filters.
+Results indicate that making some aspects of scattering networks learnable yields significant improvements compared to non-learnable filters.
+Particularly, fixed spatial and/or time filters with learnable across-channel filters seem to obtain the best results.
+
+For gradient descent, it may be beneficial to perform convolutions directly, and not in the Fourier domain.
+Separable filters have been shown to be effective in a \ac{cnn} settings, while also allowing for much quicker convolution operations.
+Separable filters also allow for the independance of each dimension, which can be particularly useful for optimisations such as downsampling.
+This type of scheme is illustrated by higher-dimensional dyadic wavelet transforms such as the \ac{dwt}.
+
+Downsampling schemes are trivial when the axes are independant -- a filterbank can be configured independantly for each axis and all operations for following axes do not dependent on previous filtering operations.
+When a mix of time and spatial dimensions are present, techniques such as 2D, 3D and harmonic wavelet scattering may not be suitable, since their invariance filters cannot provide a certain amount of invariance for each dimension independently.
+Additionally, 2D and 3D scattering tradionally allow the user to create a filterbank with more than 1 filter per octave across $r$ and $\rho$ respectively.

### 2024-05-21 dissertation/content/conclusion.tex
added astract and conclusion

\chapter{Conclusion}

In this dissertation, we investigate the viability of wavelet transforms as alternative \ac{tf} decompositions in for detecting and classifying whale calls.

In chapter \ref*{chap:p1}, we show that the \ac{cwt} can be used for \ac{se} detectors, which can, under certain conditions, yield marginally better performance in terms of time-localisation compared to the \ac{stft}. Median filtering of the \ac{se} statistic and k-means clustering further improve the interpretability and accuracy of the \ac{se} detector. We show that this detector is much better than \acp{bled}, which are the most commonly used signal detectors in \ac{pam}.

Chapter \ref*{chap:p2} further expands on the ideas of chapter \ref*{chap:p1}, by including additional time-shift invariance to the \ac{tf} decomposition via first-level wavelet scattering coefficients. We additionally model the entropy as a \ac{gmm}, which is a further improvement on k-means. We lay the groundwork for efficient downsampling strategies in for wavelet scattering computations. Since wavelet scattering can be used a features, we combine the detector output with a classifier to investigate the practicality of such a system. We provide critical discussions on the dataset that is used, and highlight the real-world problems faced in evaluation \ac{pam} detections and classification systems.

From the foundation established in chapter \ref*{chap:p2} and inspired by joint-\ac{tf} scattering, we generalise the notation of using separable wavelet filters in a multi-dimensional scattering transform in chapter \ref*{chap:p3}. We show that, in a modern \ac{ml}-setting, it can be equally effective compared to other wavelet scattering filterbanks, yet provides significant advantages due to filter separability, which includes greater configuration freedoms and computational advantages.

Further work includes deploying separable wavelet scattering and the improved \ac{se} detector from chapter \ref*{chap:p2} in a \ac{pam} detection and classification system, while evaluating its performance compared to \ac{nn}-based approaches.

The papers presented in this dissertation naturally introduce wavelet scattering to both signal detection and classification for \ac{pam} systems thereby addressing some of the shortcomings of \acp{nn}. We develop a new type of scattering which is particularly well suited to the requirements of \ac{pam}.

### 2024-05-21 dissertation/content/introduction.tex
added astract and conclusion

@@
 Often, when there is processing power available, a \ac{ml} model will be deployed to increase efficiency, accuracy and limit human interaction.
-Various models, whether they are statistical, Bayesian or probabilistic, decision trees or \ac{nn}, are configured to assist or perform various tasks across many industries.
+Various models, whether they are statistical, Bayesian or probabilistic, decision trees or \acfp{nn}, are configured to assist or perform various tasks across many industries.
 
@@
 The collection of data therefore becomes an important part of model training and deployment.
-\Acp{nn} present some challenges and concerns, since they require vast amounts of data, and their operation is often unpredictable and not well understood.
+\Acp{nn} present some challenges and concerns, since they require vast amounts of data, and their operation is often unpredictable and not well understood \citep{nnsurvey}.
 They are typically treated as ``black box'' functions, which researchers improve and design through trial and error via hyper-parameter optimisation and architecture exploration.
 
-In the field of \ac{pam}, which focuses on monitoring wildlife using non-invasive acoustic sensors, the majority of recent \ac{sota} models are \ac{nn}-based.
+In the field of \ac{pam}, which focuses on monitoring wildlife using non-invasive acoustic sensors, the majority of recent \ac{sota} models are \ac{nn}-based \citep{pamsurvey}.
 However, metrics of real-life deployment and practical model evaluation are often ignored, since the acoustical landscape of natural environments can be surprisingly varied.
-For example, one cannot expect a \ac{nn} model which was not trained on data that contains rainfall noise, to effectively operate when it starts raining.
+For example, one cannot expect a \ac{nn} model which was trained on data that does not contain rainfall noise, to effectively operate when it starts raining.
+This is especially relevant for \acp{cnn}, since the learnt filters are often unexpectedly shaped in the frequency domain \cite{sincnet}.
 
-Long moments of silence, weather patterns and migrating animals contribute to an ever-changing soundscape which can cause many false positives in naively trained and deployed models.
+Long moments of silence, weather patterns and migrating animals contribute to an ever-changing soundscape which can cause many false positives in naively trained and deployed models \citep{pamguide}.
 Systems which aim to track specific species are often faced with this ``needle in a haystack'' problem, in which the sounds you are looking for are rare compared to the lifetime of the deployed sensors.
@@
 At the end of this investigative journey, we naturally arrive an entirely new generalisation of a feature extraction method based on \ac{ct} wavelets, which can be utilised as a drop-in replacement for a \ac{cnn} front-end.
-The integration and investigation of wavelet-based processing in \ac{pam} systems are relatively scarce, even though it is very prevalent in other fields, which we aim to remedy in the publications presented in this dissertation.
+Wavelet scattering features be further improved with deep learning across filter channels \cite{nnphasecollapse}.
+The integration and investigation of wavelet-based processing in \ac{pam} systems are relatively scarce, even though it is very prevalent in other fields \citep{denoising,ws_fault_diag, waveletfault, waveletflight, waveletflow, waveletimagedenoising}, which we aim to remedy in the publications presented in this dissertation.
 
 \section{Problem Statement}
+\label{sec:problem}
 
@@
 
-All the above factors seem demotivate the use of \acp{nn} for the following reasons: principle of operation can be unpredictable and not well understood; can take long to compute; requires reliable training data or special techniques to mitigate problematic data.
+All the above factors seem demotivate the use of \acp{nn} for the following reasons: principle of operation can be unpredictable; computation time; requires reliable training data or special techniques to mitigate problematic data.
 Although it is possible to address these challenges, it is worth exploring non-\ac{nn} methods.
 
-As a reponse to the challenges posed by \ac{pam} systems, we focus specifically on alternative aproaches to \acp{nn}, by considering the modification of traditional classifiers and detectors with wavelet-based \ac{tf}-decompositions.
+As a reponse to the challenges posed by \ac{pam} systems, we focus specifically on alternative approaches to \acp{nn}, by considering the modification of traditional classifiers and detectors with wavelet-based \ac{tf}-decompositions.
 
@@
 
-As this is a dissertation by publication, each paper can be read as a stand-alone item, with all the necessary literature and knowledge capture within.
+As this is a dissertation by publication, each paper can be read as a stand-alone item, with all the necessary literature and knowledge captured within.
 As such, this dissertation is structured with additional contextual chapters (chapters \ref*{chap:p1i}, \ref*{chap:p2i}, \ref*{chap:p3i}) prior to each publication (chapters \ref*{chap:p1}, \ref*{chap:p2}, \ref*{chap:p3}).
 
-Each contextual chapter is in itself an introduction to the literature and concepts presented by the paper in the following paper.
-Contextual chapters serve to unify the content of the published papers and show the natural progression of the analysis and application of \ac{tf} decompositions.
+Each contextual chapter is in itself an introduction to the literature and concepts presented in the following chapter.
+Contextual chapters serve to unify the content of the published papers and show the natural progression \ac{tf} decompositions, which finally arrives at a new wavelet scattering transform.
 All relevant literature is therefore gradually introduced, due to each paper conceptually expanding on the previous paper's ideas and contributions.
@@
 Although this dissertation is mostly focussed on the application to \ac{pam} systems, the theory and findings it presents are general.
-As such, the contextual chapters are generalised mathematical descriptions of the concepts present in each paper, whereas the papers 1 and 2 (chapters \ref*{chap:p1}, \ref*{chap:p2}) are focussed on applying it to \ac{pam} systems.
+As such, the contextual chapters are generalised mathematical descriptions of the concepts present in each paper.
+Papers 1 and 2 (chapters \ref*{chap:p1}, \ref*{chap:p2}) are focussed on applying it to \ac{pam} systems.
 The final paper (chapter \ref*{chap:p3}) shows a generalised result from all the knowledge and investigations presented in prior papers.
@@
 \section{Journal Papers and Contributions}
+
+The three papers presented in this dissertation and their contribution in relation to \ac{pam} and general \ac{ml} is described below.
+Note that the actual journal article titles are used, which are different from the chapter titles.
+
+\begin{enumerate}
+    \item \textbf{Chapter \ref*{chap:p1}: Soft-output Signal Detection for Cetacean Vocalizations using Spectral Entropy, K-means Clustering and the Continuous Wavelet Transform} -- we consider the use of the \ac{cwt} as a \ac{tf} decomposition for \ac{se} detectors, where the \ac{stft} is typically used.
+We then extend the conventional detector to be of a probabilistic nature.
+This paper introduces wavelet-based \ac{tf}decompositions in this dissertation and lays the groundwork for further extension and investigation.
+    \item \textbf{Chapter \ref*{chap:p2}: Detecting and classifying Whale Calls with Wavelet Scattering and Spectral Entropy} -- we build on the work of chapter \ref*{chap:p1}, and instead consider wavelet scattering as a \ac{tf} decompistion, for which the first-level coefficients can be considered as a ``blurred'' \ac{cwt}.
+We further improve the \ac{se} detector using \acp{gmm}, and integrate the detector system with a classifier using wavelet scattering features.
+This paper focusses on the practicality of combining and improving older methods, while also introducing wavelet scattering to the field of \ac{pam}.
+    \item \textbf{Chapter \ref*{chap:p3}: Separable Wavelet Scattering} -- following from the ideas presented in chapters \ref*{chap:p1} and \ref*{chap:p2}, especially with reference to computational efficiency discussed in chapter \ref*{chap:p2}, an entirely new type of wavelet scattering is generalised.
+We evaluate the new wavelet scattering technique with the MNIST and MedMNIST datasets for quick evaluation of the method compared to other similar wavelet scattering decompositions.
+We consider this paper to be the largest contribution presented in this dissertation, due to its general nature, whose design stemmed from the needs of \ac{pam} systems as discussed in section \ref*{sec:problem}.
+\end{enumerate}
+
+I am the main author in all the papers presented in this dissertation, in which I conducted all relevant research, performed the writing and submitted to the journals.
+Each chapter containing a publication declares detailed author contributions in terms of the CRediT system.

### 2024-05-21 dissertation/content/p1_intro.tex
added astract and conclusion

@@
 
-To construct a wavelet filterbank with $\upsilon_k \in \Upsilon$ which has the exponentially-spaced frequency properties discussed in section \ref{sec:p1i:tfa}, we use a ``template'' wavelet $\uppsi(t)$, referred to as the mother wavelet.
+To construct a wavelet filterbank with $\upsilon_k \in \Upsilon$ which has the exponentially-spaced frequency properties discussed in section \ref{sec:p1i:tfa}, we use a ``template'' wavelet $\uppsi(t)$, referred to as the mother wavelet \cite{waveletbook}.
 Conventionally, wavelet filters are defined as \ac{ct} filters and then discretised in software implementations.

### 2024-05-21 dissertation/main.tex
before reference consolodation
consolodated references
started intro to p3
started intro
added astract and conclusion

@@
 \noindent\rule{\textwidth}{0.25pt}%
+% \pagebreak
 \section*{\centering \textsc{Abstract}}
@@
 \end{spacing}
-\noindent\rule{\textwidth}{0.5pt}\pagebreak
+\clearpage
 }
@@
 \addcontentsline{toc}{section}{Appendices}
-\counterwithin{figure}{section}
-\counterwithin{table}{section}
 }
@@
 \AtEndEnvironment{subappendices}{%
-\counterwithout{figure}{section}
-\counterwithout{table}{section}
 }
+
+% Chapter format
+\usepackage{titlesec}
+\titleformat{\chapter}
+  {\LARGE\bfseries}{\thechapter}{1em}{}
+\titlespacing*{\chapter}{0pt}{3.5ex plus 1ex minus .2ex}{2.3ex plus .2ex}
 
@@
 \begin{document}
-\title{\textbf{Wavelet Scattering Tranforms Applied to Whale Vocalisations}}
+\title{\AorE{Wavelet Scattering Transformasies Toegepas op Walvis Vokaliserings}{Wavelet Scattering Transforms Applied to Whale Vocalisations}}
 \author{M.W. Rademan}{Marco Wiehann Rademan}
@@
 \SetCopyrightOff
-\address{\AorE{Departement Siviele Ingenieurswese,\\ Stellenbosch Universiteit,\\
+\address{\AorE{Departement Elektriese en Elektroniese Ingenieurswese,\\ Stellenbosch Universiteit,\\
 Privaatsak X1, 7602 Matieland, Suid Afrika.
 }{%
-Department of Civil Engineering,\\ Stellenbosch University,\\
+Department of Electrical and Electronic Engineering,\\ Stellenbosch University,\\
 Private Bag X1, 7602 Matieland, South Africa.}}
@@
 
+In this dissertation by publication, we investigate the application of wavelet scattering transforms on whale vocalisations, by following the natural progression of 3 consecutive papers.
+
+We consider how wavelets can be used to replace the short-time Fourier transform (STFT) in signal detection.
+Wavelet scattering is introduced to the field of passive acoustic monitoring as a convolutional neural network (CNN) alternative for small datasets, which is combined with an improved entropy-based signal detector.
+
+For 1-dimensional (1D) signals, wavelet scattering can be considered as a generalisation of Mel-frequency cepstral coefficients (MFCCs), and can typically outperform MFCCs in audio classification applications.
+In general, wavelet scattering filtering operations resemble the first layers of a CNN.
+An effective feature extractor may be constructed with wavelet scattering filters without requiring any learning.
+The main contribution of this dissertation is the introduction of separable filters for a wavelet scattering transform, which is generalised to an arbitrary number of dimensions.
+
 \end{abstract}
@@
 \begin{abstract}[afrikaans]
+
+  In hierdie publikasie proefskrif, ondersoek ons die toepassing van wavelet scattering transformasies op walvis vokaliserings, deur middel van opeenvolgende joernaal arikels.
+
+  Ons orweeg hoe wavelets gebruik kan word om die kort-tyd Fourier transformasie (KTFT) in sein deteksie te vervang.
+Wavelet scattering is bekendgestel in die veld van passiewe akoustiese monitering, as an alternatief to konvolusionele neurale netwerke (KNNs).
+Verder, kombineer ons wavelet scattering met 'n klassieke entropie-gebasseerde sein detektor wat ons verbeter met addisionele tegnieke.
+
+  Vir eendimensionele seine kan wavelet scattering as 'n veralgemening van Mel-frekwensie kepstrale koeffisi\"{e}nte (MFKKs) beskou word, altans 'n verbeterde weergawe daarvan.
+In die algemeen, is wavelet scattering vergelykbaar met die eerste lae van KNNs.
+Die hoof bydrag van hierdie proefskrif is die bekendstelling van skeibare filters vir die wavelet scattering transform, wat veralgemeen is vir 'n arbitr\^{e}re aantal intreedimensies.
 
@@
 \label{chap:nomenclature}
+\markboth{Nomenclature}{\uppercase{Nomenclature}}
 \addcontentsline{toc}{chapter}{\nameref{chap:nomenclature}}
+\input{content/nomenclature.tex}
 
@@
 \input{content/paper3/main}
+
+\input{content/conclusion}
 % \end{singlespace}
@@
 \bibliographystyle{stb-bib-eng-a}
-\bibliography{references}
+\bibliography{references_consolodated}
 

### 2024-05-27 dissertation/content/conclusion.tex
draft edit nearly complete, need to do refs

@@
 Chapter \ref*{chap:p2} further expands on the ideas of chapter \ref*{chap:p1}, by including additional time-shift invariance to the \ac{tf} decomposition via first-level wavelet scattering coefficients.
-We additionally model the entropy as a \ac{gmm}, which is a further improvement on k-means.
+We additionally model the entropy as a \ac{gmm}, which is a further improvement of the k-means implementation in the first paper.
 We lay the groundwork for efficient downsampling strategies in for wavelet scattering computations.
@@
 
-The papers presented in this dissertation naturally introduce wavelet scattering to both signal detection and classification for \ac{pam} systems thereby addressing some of the shortcomings of \acp{nn}.
+The papers presented in this dissertation naturally introduce wavelet scattering to both signal detection and classification for \ac{pam} systems, thereby addressing some of the shortcomings of \acp{nn}.
 We develop a new type of scattering which is particularly well suited to the requirements of \ac{pam}.

### 2024-05-27 dissertation/content/introduction.tex
draft edit nearly complete, need to do refs

@@
 
-\Ac{ml} models have become mainstay in software products and automation pipelines.
+\Ac{ml} models have become mainstay in software products and auto\-mation pipelines.
 Often, when there is processing power available, a \ac{ml} model will be deployed to increase efficiency, accuracy and limit human interaction.
@@
 
-Modern research is very focussed on advancing \ac{ml} integration, improving model performance and investigating new techniques.
+Modern research is very focussed on advancing \ac{ml} integration, improving model per\-formance and investigating new techniques.
 A large branch of this research is particularly focussed on data-driven approaches using \acp{nn}.
@@
 \Acp{nn} present some challenges and concerns, since they require vast amounts of data, and their operation is often unpredictable and not well understood \citep{nnsurvey}.
-They are typically treated as ``black box'' functions, which researchers improve and design through trial and error via hyper-parameter optimisation and architecture exploration.
+They are typically treated as ``black box'' functions, which researchers improve and design through trial and error via hyper-para\-meter optimisation and architecture exploration.
 
 In the field of \ac{pam}, which focuses on monitoring wildlife using non-invasive acoustic sensors, the majority of recent \ac{sota} models are \ac{nn}-based \citep{pamsurvey}.
-However, metrics of real-life deployment and practical model evaluation are often ignored, since the acoustical landscape of natural environments can be surprisingly varied.
+However, metrics of real-life deployment and practical model evaluation are often ignored, since the acoustical landscape of natural environments can be surpri\-singly varied.
 For example, one cannot expect a \ac{nn} model which was trained on data that does not contain rainfall noise, to effectively operate when it starts raining.
@@
 
-As such, designing models and systems which are robust to all the aforementioned factors are critical in automating wildlife conservation studies based on non-invasive audio sensors.
+As such, designing models and systems which are robust to all the aforementioned factors are critical in automating wildlife conservation studies with non-invasive audio sensors.
 This leads to the natural question of the suitability of \acp{nn} in this field and to what extent they can be used to be effective.
@@
 At the end of this investigative journey, we naturally arrive an entirely new generalisation of a feature extraction method based on \ac{ct} wavelets, which can be utilised as a drop-in replacement for a \ac{cnn} front-end.
-Wavelet scattering features be further improved with deep learning across filter channels \cite{nnphasecollapse}.
+Wavelet scattering features can be further improved with deep learning across filter channels \cite{nnphasecollapse}, still making it useful in a deep learning context.
 The integration and investigation of wavelet-based processing in \ac{pam} systems are relatively scarce, even though it is very prevalent in other fields \citep{denoising,ws_fault_diag, waveletfault, waveletflight, waveletflow, waveletimagedenoising}, which we aim to remedy in the publications presented in this dissertation.
@@
 \begin{enumerate}
-    \item Long periods of silence or noise, leading to unbalanced data classes which can have a tendency for creating many false positives in models if not trained correctly.
-    \item Constantly evolving noise landscape, with the interference of other bioacoustical sounds which we are often not interested in classifying.
+    \item Long periods of silence or noise, leading to unbalanced data classes which can have a tendency for creating many false positives in ML models if not trained correctly.
+    \item Constantly evolving noise landscape, with the interference of other unpredictable bioacoustical sounds which we are often not interested in classifying.
     \item Large audio databases and recordings, which motivates the use of fast algorithms and processing techniques.
@@
     \item Small datasets, which only cover a portion of environment variability, making it unsuitable for a general model.
-    \item Non-white, non-stationary background noise, which makes classification and signal detection non-trivial tasks.
+    \item Non-white, non-stationary background noise, which can make classification and signal detection non-trivial tasks.
     \item Poorly understood species behaviour in terms of their bioacoustics and movements.
@@
 
-All the above factors seem demotivate the use of \acp{nn} for the following reasons: principle of operation can be unpredictable; computation time; requires reliable training data or special techniques to mitigate problematic data.
+All the above factors seem demotivate the use of \acp{nn} for the following reasons: principle of operation can be unpredictable; computation time; it requires reliable training data or special techniques to mitigate problematic data.
 Although it is possible to address these challenges, it is worth exploring non-\ac{nn} methods.
@@
 Each contextual chapter is in itself an introduction to the literature and concepts presented in the following chapter.
-Contextual chapters serve to unify the content of the published papers and show the natural progression \ac{tf} decompositions, which finally arrives at a new wavelet scattering transform.
-All relevant literature is therefore gradually introduced, due to each paper conceptually expanding on the previous paper's ideas and contributions.
+Contextual chapters serve to unify the content of the published papers and show the natural progression of analysing the function and purpose \ac{tf} decompositions.
+All relevant literature is therefore gradually introduced, since each paper conceptually expanding on the previous paper's ideas and contributions.
 
@@
 As such, the contextual chapters are generalised mathematical descriptions of the concepts present in each paper.
-Papers 1 and 2 (chapters \ref*{chap:p1}, \ref*{chap:p2}) are focussed on applying it to \ac{pam} systems.
+Papers 1 and 2 (chapters \ref*{chap:p1}, \ref*{chap:p2}) are focussed on applying wavelet \ac{tf} decompositions to \ac{pam} systems.
 The final paper (chapter \ref*{chap:p3}) shows a generalised result from all the knowledge and investigations presented in prior papers.
@@
 
+\clearpage
 \begin{enumerate}
-    \item \textbf{Chapter \ref*{chap:p1}: Soft-output Signal Detection for Cetacean Vocalizations using Spectral Entropy, K-means Clustering and the Continuous Wavelet Transform} -- we consider the use of the \ac{cwt} as a \ac{tf} decomposition for \ac{se} detectors, where the \ac{stft} is typically used.
+    \item \textbf{Chapter \ref*{chap:p1}: Soft-output Signal Detection for Cetacean Vocalizations using Spectral Entropy, K-means Clustering and the Continuous Wavelet Trans\-form} -- we consider the use of the \ac{cwt} as a \ac{tf} decomposition for \ac{se} detectors, where the \ac{stft} is typically used.
 We then extend the conventional detector to be of a probabilistic nature.
-This paper introduces wavelet-based \ac{tf}decompositions in this dissertation and lays the groundwork for further extension and investigation.
-    \item \textbf{Chapter \ref*{chap:p2}: Detecting and classifying Whale Calls with Wavelet Scattering and Spectral Entropy} -- we build on the work of chapter \ref*{chap:p1}, and instead consider wavelet scattering as a \ac{tf} decompistion, for which the first-level coefficients can be considered as a ``blurred'' \ac{cwt}.
-We further improve the \ac{se} detector using \acp{gmm}, and integrate the detector system with a classifier using wavelet scattering features.
+This paper introduces wavelet-based \ac{tf} decompositions in this dissertation and lays the groundwork for further extension and investigation.
+    \item \textbf{Chapter \ref*{chap:p2}: Detecting and classifying Whale Calls with Wavelet Scatter\-ing and Spectral Entropy} -- we build on the work of chapter \ref*{chap:p1}, and instead consider wavelet scattering as a \ac{tf} decomposition, for which the first-level coeffi\-cients can be considered as a ``blurred'' \ac{cwt}.
+We further improve the \ac{se} detector using \acfp{gmm}, and integrate the detector system with a classifier using wavelet scattering features.
 This paper focusses on the practicality of combining and improving older methods, while also introducing wavelet scattering to the field of \ac{pam}.
     \item \textbf{Chapter \ref*{chap:p3}: Separable Wavelet Scattering} -- following from the ideas presented in chapters \ref*{chap:p1} and \ref*{chap:p2}, especially with reference to computational efficiency discussed in chapter \ref*{chap:p2}, an entirely new type of wavelet scattering is generalised.
-We evaluate the new wavelet scattering technique with the MNIST and MedMNIST datasets for quick evaluation of the method compared to other similar wavelet scattering decompositions.
+We evaluate the new wavelet scattering technique with the MNIST and MedMNIST datasets to benchmark the performance of the features provided by separable scattering.
 We consider this paper to be the largest contribution presented in this dissertation, due to its general nature, whose design stemmed from the needs of \ac{pam} systems as discussed in section \ref*{sec:problem}.

### 2024-05-27 dissertation/content/nomenclature.tex
added astract and conclusion
draft edit nearly complete, need to do refs

@@
 \section*{Notation}
-Some notations are used interchangably, which is obvious from the context. Due to the nature of the publication format, notation is explained within the article it occurs.
+Some notations are used interchangably, which is obvious from the context. Due to the nature of the publication format, notation is additionally explained within the article it occurs.
 
@@
 
+\section*{Notation (continued)}
 \begin{tabular}{p{0.15\textwidth} p{0.75\textwidth}}
-    $X[k,m]$ & a time-frequency decomposition of $x$, with $k$ the frequency index and $m$ the time index \\
+    $(\vect{x})_{\downarrow \vect{r}}$ & a discrete multidimensional signal subsampled in each dimension by the components of $\vect{r}$ \\
+    $X[k,m]$ & a time-frequency decomposition of a discrete signal $x$, with $k$ the frequency index and $m$ the time index \\
     $S[k,m]$ & power spectrum calculated from $X[k, m]$ \\
@@
     $X(s, b)$ & continuous wavelet transform of $x$, with $s$ as scale and $b$ as translation \\
-    $\uppsi_\lambda$ & sometimes used as short-hand for  $\uppsi(\lambda t)$\\
-    $\text{RM}^F_m (x)$ & rolling median filter operator with a window of size $F$ operating on index variable $m$ on a discrete function $x$\\
-    $\text{RA}^F_m (x)$ & rolling average filter operator with a window of size $F$ operating on index variable $m$ on a discrete function $x$\\
+    $\uppsi_\lambda$ & used as short-hand for a wavelet dilation -- $\lambda\uppsi(\lambda t)$\\
     $x \triangleq y$ & $x$ is defined as $y$\\
-\end{tabular}
-
-\clearpage
-\section*{Common Variables}
-Some variables are used interchangably, which is obvious from the context. Due to the nature of the publication format, variables are explained within the article it occurs.
-
-\begin{tabular}{p{0.15\textwidth} p{0.75\textwidth}}
-    a & Hello \\
-    a & Hello \\
-    a & Hello \\
-    a & Hello \\
 \end{tabular}
@@
 
+\section*{Acronyms (continued)}
 \begin{tabular}{p{0.15\textwidth} p{0.75\textwidth}}

### 2024-05-27 dissertation/content/p1_intro.tex
draft edit nearly complete, need to do refs

@@
 \end{equation}
-where $\upsilon_k \in \Upsilon$ is the $k$'th \ac{bpf} in the decomposition filter dictionary $\Upsilon$. Note that a \ac{lpf} can be considered to be a \ac{bpf} with a centre-frequency of 0.
+where $\upsilon_k \in \Upsilon$ is the $k$'th \ac{bpf} in the decomposition filter dictionary $\Upsilon$. Note that we consider a \ac{lpf} to be a \ac{bpf} with a centre-frequency of 0.
 
 Generally, the entire convolution $X[n,k]$ is not computed.
-Instead, only select samples are computed by speficying a hop-distance or downsample factor $d$.
-We indicate this by tranforming the index variable $m = dn$ as
+Instead, only select samples are computed by specifying a hop-distance or downsample factor $d$.
+We indicate this by transforming the index variable $m = dn$ as
 \begin{equation}
@@
 
-The is equivalent to applying a downsample operator $\left(\cdot\right)_{\downarrow d}$ to the convolution.
+Hop-distance is equivalently expressed by applying the downsample operator $\left(\cdot\right)_{\downarrow d}$ to the convolution.
 
@@
 \begin{equation}
+    \label{eqn:fouriertransform}
     \mathcal{F} \{x\} \triangleq \hat{x}(\omega) = \int_{-\infty}^{\infty} x(t) e^{-j\omega t} dt = \left(x * e^{j\omega t}\right)\bigg|_{t=0}.
@@
 
-Derived from the conventional Fourier transform, a \ac{dt} signal $x[n], n \in \{0, ..., N-1\}$ (length $N$) has a \ac{dtft}
+Derived from the conventional Fourier transform given by equation (\ref*{eqn:fouriertransform}), a \ac{dt} signal $x[n], n \in \{0, ..., N-1\}$ of length $N$ samples has a \ac{dtft}
 \begin{equation}
@@
 for which $\upsilon_k = e^{j \frac{2\pi k n}{N}}, \ k \in \{0, ..., N-1\}$ forms an orthogonal basis of $x$.
-The circular convolution operator $\otimes$ implies the infinite periodic extension of $x$ beyond its boundaries of $n \in \{0, ..., N-1\}$.
+The circular convolution operator $\otimes$ implies the infinite periodic extension of $x$ beyond its boundaries of $n \in \{0, ..., N-1\}$ \citep{dspbook}.
 
-Equation (\ref{eqn:dtft}) describes the \ac{dtft} as convolution evaluated at the index 0 as a useful link to the generalised notion of \ac{tf} analyses: Fourier transforms can be considered as complex sinusoidal filters being applied to the signal of interest, with periodic boundary extension.
+Equation (\ref{eqn:dtft}) describes the \ac{dtft} as a convolution evaluated at the index 0 as a useful link to the generalised notion of \ac{tf} analyses: Fourier transforms can be considered as complex sinusoidal filters being applied to the signal of interest, with periodic boundary extension.
 
@@
 
-With the correct padding procedures, one can compute $x * y$ using (\ref{eqn:fftconv}).
+With the correct padding procedures, one can compute $x * y$ using equation (\ref{eqn:fftconv}).
 
@@
 
-The most popular \ac{tf} analysis is the \ac{stft}. In terms of equation (\ref{eqn:gentf}), a \ac{stft} is computed with 
+The most popular \ac{tf} analysis is the \ac{stft}. In terms of equation (\ref{eqn:gentf}), a \ac{stft} is computed with a set of \ac{tf} filters
 \begin{equation}
@@
 where $w$ is a window function defined by the user.
-Various \acp{fft} are use to compute this transform, so the properties of $w$ is typically specified by the \ac{fft} length, and the window type (Hann, Hamming, Blackman, etc.).
+Various \acp{fft} are use to compute this transform, so the properties of $w$ is typically specified by the \ac{fft} length, and the window type (Hann, Hamming, Blackman, etc.) \citep{dspbook}.
 The hop-length (or downsample factor) $d$ is an additional parameter specified by the user.
-It is possible to choose $w$ and $d$ such that the \ac{stft} becomes an invertible operator -- the original signal can be entirely recovered.
+It is possible to choose $w$ and $d$ such that the \ac{stft} becomes an invertible operator -- the original signal can be entirely recovered \citep{stftbook}.
 
@@
 For example, if $a=2$, the fundamental's frequency index is shifted to $k_0 + Q$ (as opposed to $2 k_0$ for the \ac{stft}).
-This property is referred to a frequency-shift invariance, since small frequency deformations of the order $Q^{-1}$ do not significantly change the frequency index structure of the \ac{tf} decomposition.
+This property is referred to a frequency-shift invariance, since small frequency deformations of the order $Q^{-1}$ do not significantly change the frequency index structure of the \ac{tf} decomposition \citep{1dscattering1}.
 The \ac{stft} is much more sensitive to such deformations across its entire structure for harmonically rich signals.
@@
 
-$\uppsi$ can the be scaled by a factor $\lambda$ such that
+$\uppsi$ can be scaled by a factor $\lambda$ such that
 \begin{equation}
@@
 \end{equation}
+
+The multiplicate scaling, in addition to time scaling, ensures that $\uppsi_{\lambda_k}$ retains constant amplitude in the frequency domain, also referred to as L1 normalisation \cite{waveletbook}.
 

### 2024-05-27 dissertation/content/p2_intro.tex
added astract and conclusion
draft edit nearly complete, need to do refs

@@
 
-\ac{tf} decompositions are often used to extract features for signals.
+\ac{tf} decompositions are often used to extract features from signals.
 In audio, \acp{mfcc} are features calculated from the \ac{stft} of the signal.
-A ``\ac{tf}'' decomposition, as referred to in this dissertation, does not necesarrily require a time variable: spatial variables are often used.
+Note that a ``\ac{tf}'' decomposition, as referred to in this dissertation, does not necesarrily require a time variable: spatial variables are often present instead.
 In this sense, a \ac{cnn} is can also be considered as a ``\ac{tf}'' decomposition, where multi-dimensional index variables (or index vectors) are used for ``time'' and frequency:
@@
 If $\Phi_i$ is an averaging (low-pass) filter, some additional stability can be observed in $\tilde{X_i}$ compared with $X$.
-This allows for the introduction of invariances to various deformations applied to $X$, which can be greatly beneficial for some applications.
-In a sense, cascased structures of $X$ and $\tilde{X_i}$ are defined in a \ac{cnn} for which the network learns the required invariances.
+This allows for the introduction of invariances to various deformations applied to $X$, which can be greatly beneficial for some applications \citep{cnninvariance}.
+In a sense, cascased structures of $X$ and $\tilde{X_i}$ are defined in a \ac{cnn} for which the network learns the required invariances \cite{2dscattering}.
 
 It is widely known that enforcing invariance in feature extraction and \ac{ml} pipelines increase performance.
-Enforcing invariance in \acp{nn} has been shown to improve the network, while also improving the interpretability of the network.
-For example, when classifying images, invariance to scale, shear, rotation and translation deformations, depending on image properties, may be useful when extracting features.
-Specifically, deformations of a specified order should approximately map to the same point in feature space:
+Enforcing invariance in \acp{nn} has been shown to improve the network, while also improving the interpretability of the network \cite{cnninvariance2}.
+For example, when classifying images, invariance to scale, shear, rotation and translation deformations, de\-pending on image properties, may be useful when extracting features.
+Specifically, defor\-mations of a specified order should approximately map to the same point in feature space:
 \begin{equation}
@@
 Invariance filters ($\Phi_i$) can also have additional properties that makes it favourable for audio applications.
-For example, many applications ``blur'' (low-pass) the \ac{stft} magnitude or power spectrum, which can reduce the impact of noise fluctuations, and can also serve to suppress small frequency fluctuations across adjascent frequency bins (if the blurring kernel $\Phi_i$ filters across frequency as well).
+For example, many applications ``blur'' (low-pass) the TF decomposition \citep{seyicwt, speechmodulation}, which can reduce the impact of noise fluctuations, and can also serve to suppress small frequency fluctuations across adjascent frequency bins (if the blurring kernel $\Phi_i$ filters across frequency as well).
 The blurring operation can improve the performance of signal detectors and classifiers.
@@
 
-A new problem is introduced when enforcing invariances via the $\Phi_i$ filters - information is lost via averaging.
+A new problem is introduced when enforcing invariances via the $\Phi_i$ filters - information is lost via averaging \cite{1dscattering1}.
 Various techniques can be used to further retain information, while also retaining invariance.
-Wavelet scattering is such a method.
-In a sense, multilayer \acp{cnn} topologies with skip connections can be though of as a method to recover lost information in the deeper layers.
+Wavelet scattering is such a method \cite{ws,2dscattering}.
+In a sense, multilayer \acp{cnn} topologies with skip connections can be thought of as a method to recover lost information in the deeper layers.
 
@@
 
-Since \acp{mfcc} are the most widely used feature extraction method for audio signals, save for \ac{cnn} frontends which also act as a feature extractor when operating on the \ac{stft}.
-For this reframing, we ignore the \ac{dct} applied to the filtered values, since this linear transformation only serves to decorrelate the filter coefficients.
+\acp{mfcc} are the most widely used feature extraction method for audio signals \citep{mfccreview}, save for \ac{cnn} frontends which also act as a feature extractor when operating on the \ac{stft}.
+For this reframing, we ignore the \ac{dct} applied to the filtered values, since this linear transformation only serves to decorrelate the filter coefficients \citep{dctdecorrelation}.
 
@@
 
-We can view this process as a \ac{tf} decompistion with STFT filters $\upsilon_l \in \Upsilon$, which is modified with the Mel-scale triangular filters $\Phi_k$.
+We can view this process as a \ac{tf} decomposition with STFT filters $\upsilon_l \in \Upsilon$, which is modified with the Mel-scale triangular filters $\Phi_k$.
 $\upsilon_l$ provides time-shift invariance due to the STFT window, whereas $\Phi_k$ selectively averages frequency content of the STFT, thereby providing some frequency-shift invariance.
@@
 
-The notation in equations (\ref*{eqn:general_scalogram}) and (\ref*{eqn:general_scattering}) conforms the general \ac{tf} description of chapter \ref*{chap:p1i}, although the specific wavelet scattering operators in chapters \ref*{chap:p2} to \ref*{chap:p3} follow a different standard.
+The notation in equations (\ref*{eqn:general_scalogram}) and (\ref*{eqn:general_scattering}) conforms the general \ac{tf} description of chapter \ref*{chap:p1i}, although the specific wavelet scattering operators in chapters \ref*{chap:p2} to \ref*{chap:p3} follow standard notation \citep{waveletbook}.
 These operators are known as wavelet scattering, which formalise the notion of ``averaging a \ac{tf}-decomposition'' in terms of signal processing theory.

### 2024-05-27 dissertation/content/p3_intro.tex
draft edit nearly complete, need to do refs

--- 2024-05-14 dissertation/content/p3_intro.tex
@@
 In chapter \ref*{chap:p2i}, we reviewed a generalised definition of a multi-dimensional TF decomposition $X$, which is processed by a set of invariance-enforcing filters $\Phi_i$.
-Wavelet scattering in 1 dimension utilising a single invariance-enforcing filter, $\phi$, as discussed in chapter \ref*{chap:p2}.
+Wavelet scattering in 1 dimension utilises a single invariance-enforcing filter, $\phi$, as discussed in chapter \ref*{chap:p2}.
 
@@
 Wavelet scattering filterbanks have been extended to multiple dimensions using various different wavelets and filterbank construction techniques.
-This chapter provides an overview of the existing filterbanks and filters and their properties.
+This chapter provides an over\-view of the existing wavelet filterbanks and filters and their properties.
 We then identify a gap in the literature, which gives rise to a generalisation of cascasded 1D scattering transforms -- separable wavelet scattering.
@@
     \item 2D scattering \citep{2dscattering}.
-Gabor filters modified to 0-mean, are used to construct the filterbank.
+Gabor filters modified to have 0-mean are used to construct the filterbank.
 The filterbank consists of exponentially placed filters in one axis, which is then rotated by equally spaced angles to cover the frequency plane.
-This filterbank only has a single filter per octave ($Q=1$).
-In particular, in polar coordinates $(r, \theta)$ in the frequency domain, filters are placed exponentially along $r$, but linearly along $\theta$.
+In particular, in polar coordinates $(r, \theta)$ in the frequency domain, filters are placed exponentially along $r$ with a single filter per octave, but linearly along $\theta$.
     \item Harmonic scattering \citep{harmonicscattering} (3 dimensions).
 The spherical harmonics are used as filters which consitute an orthogonal basis in 3 dimensions.
-These filters have ``lobes'' that extend in a multitude of dimensions, depending on the harmonic, in the frequency domain.
+These filters have ``lobes'' that extend in the along various directions in the frequency domain, depending on the harmonic.
     \item 3D scattering -- extension of 2D scattering \citep{3dscattering}.
 The filterbank is constructed similar to 2D scattering.
-In spherical coordinates $(\rho, \theta, \phi)$ in the frequency domain, filters are spaced exponentially along $\rho$, and linearly along $\theta$ and $\phi$.
+In spherical coordinates $(\rho, \theta, \phi)$ in the frequency domain, filters are spaced exponentially along $\rho$ with a single filter per octave, and linearly along $\theta$ and $\phi$.
     \item Joint time-frequency \citep{ws_joint_tf,jointtfscattering2} scattering -- 2D filters on 1D scattering scalogram.
-A first level of scattering in performed in 1D to produce a scalogram ($X[k, m]$).
+A first level of scattering is performed on a 1D time signal to produce a 2D scalogram ($X[k, m]$).
 2D scattering is then performed on subsequent levels, with $X[k, m]$ as input.
 The 2D wavelets are separable - a product of 1D wavelets.
-This allows 1D scattering on audio to have features that are also dependent on adjascent frequency bins.
+This allows 1D scattering on audio to have features that are also dependent on adjacent frequency bins, which is not possible with 1D scattering only.
+The extracted features can therefore correlate with frequency modulations while retaining the benefits from 1D.
 It is conceptually similar to a \ac{cnn} operating on a \ac{stft} spectrogram.
@@
 \label{sec:p3i:sep}
+
+Recently, wavelet scattering operations implemented as \acp{cnn}, referred to as a scattering network, have been of increasing interest in order to increase the effectiveness of \ac{cnn} filters \cite{2dscattering, scattering_birdsong}.
+Results indicate that making some aspects of scattering networks learnable yields significant improvements compared to non-learnable filters.
+Particularly, fixed spatial and/or time filters with learnable across-channel filters seem to obtain the best results \cite{nnphasecollapse}.
+
+For gradient descent, it may be beneficial to perform convolutions directly, and not in the Fourier domain.
+Separable filters have been shown to be effective in a \ac{cnn} settings, while also allowing for much quicker convolution operations \cite{separablecnn,separablecnn2} when computed directly.
+Separable filters also allow for the independance of each dimension, which can be particularly useful for optimisations such as downsampling.
+This type of scheme is illustrated by higher-dimensional dyadic wavelet transforms such as the \ac{dwt} \cite{waveletbook}.
+
+Downsampling schemes are trivial when the axes are independent -- a filterbank can be configured independently for each axis, where subsequent axes remain oblivious to previous operations.
+When a mix of time and spatial dimensions are present, existing techniques such as 2D, 3D and harmonic wavelet scattering may not be suitable, since their invariance filters cannot provide a certain amount of invariance for each dimension independently, due to their filterbank construction.
+Additionally, 2D and 3D scattering traditionally do not allow the user to create a filterbank with more than 1 filter per octave across $r$ and $\rho$ respectively.
+Signals with different time/spatial supports (i.e., a signal may be much longer in one dimension compared to another) will likely require different invariance scales across dimensions.
+As such, conventional higher dimensional scattering transforms may not be appropriate.
+
+The type of scattering transform used is usually highly dependent on the signal dimen\-sionality -- there is no unifiying multi-dimensional analogy of 1D scattering as is present with dyadic wavelet transforms \citep{waveletbook}.
+We construct this generalisation in the chapter \ref*{chap:p3} via separable filters.

### 2024-05-27 dissertation/content/paper1/main.tex
started intro to p3
draft edit nearly complete, need to do refs

@@
 \chapter{Wavelet Transforms as a Fourier Alternative}
+\label{chap:p1}
 
-\articleinfo{M.W. Rademan, D.J.J Verself, J.A. Du Preez}{Ecological Informatics}{15 May 2024}{doi}{Underwater passive acoustic monitoring systems record many hours of audio data for marine research, 
-making fast and reliable non-causal signal detection paramount. Such detectors assist in reducing the amount of 
-labor required for signal annotations, which often contain large portions devoid of signals.\par
+\articleinfo{Soft-output Signal Detection for Cetacean Vocalizations using
+Spectral \break Entropy, K-means Clustering and the Continuous Wavelet Transform}{Ecological Informatics}{15 May 2024}{doi}
+
+\section*{CRediT}
+\textbf{M.W. Rademan}: conceptualisation; methodology; software; formal analysis; writing -- original draft.
+
+\textbf{D.J.J. Verseld}: data collection; writing -- review and editing; supervision.
+
+\textbf{J.A. Du Preez}: writing -- review and editing; supervision.
+
+\section*{Abstract}
+Underwater passive acoustic monitoring systems record many hours of audio data for marine research, 
+making fast and reliable non-causal signal detection paramount.
+Such detectors assist in reducing the amount of labor required for signal annotations, which often contain large portions devoid of signals.
+
 Cetacean vocalization detection based on spectral entropy is investigated as a means of vocalization discovery.
-Previous techniques using spectral entropy mostly consider time-frequency enhancement of the entropy measure, and 
-utilize the short time Fourier transform (STFT) as its time-frequency (TF) decomposition. Spectral entropy methods 
-also requires the user to set a detection threshold manually, which call for knowledge of the produced entropy measures.\par
-This paper considers median filtering as a simple, effective way to provide temporal stabilization to the entropy measure, 
-and considers the continuous wavelet transform (CWT) as an alternative TF decomposition. K-means clustering is used to 
-determine the threshold required to accurately separate the signal/no-signal entropy measures, resulting in a  one-dimensional, 
-two-class classification problem. The class means are used to perform pseudo-probabilistic soft class assignment, which is a 
-useful metric in algorithmic development. The effect of median filtering, signal-to-noise ratio and the chosen TF decomposition are investigated.\par 
-The accuracy and specificity measures of the proposed detection technique are simulated using a pulsed frequency modulated sweep, 
-corrupted by a sample of ocean noise. The results show that median filtering is particularly effective for low signal-to-noise ratios. 
+Previous techniques using spectral entropy mostly consider time-frequency enhancement of the entropy measure, and utilize the short time Fourier transform (STFT) as its time-frequency (TF) decomposition.
+Spectral entropy methods also requires the user to set a detection threshold manually, which call for knowledge of the produced entropy measures.
+
+This paper considers median filtering as a simple, effective way to provide temporal stabilization to the entropy measure, and considers the continuous wavelet transform (CWT) as an alternative TF decomposition.
+K-means clustering is used to determine the threshold required to accurately separate the signal/no-signal entropy measures, resulting in a  one-dimensional, two-class classification problem.
+The class means are used to perform pseudo-probabilistic soft class assignment, which is a useful metric in algorithmic development.
+The effect of median filtering, signal-to-noise ratio and the chosen TF decomposition are investigated.
+
+The accuracy and specificity measures of the proposed detection technique are simulated using a pulsed frequency modulated sweep, corrupted by a sample of ocean noise.
+The results show that median filtering is particularly effective for low signal-to-noise ratios.
 Both the STFT and CWT prove to be effective TF analyses for signal detection purposes, each presenting with different advantages and drawbacks.
-The simulated results provide insight into configuring the proposed detector, which is compared to a conventional STFT-based spectral entropy 
-detector using manually annotated humpback whale (\textit{Megaptera novaeangliae}) songs recorded in False Bay, South Africa, July 2021.\par
-The proposed method shows a significant improvement in detection accuracy and specificity, while also providing a more interpretable 
-detection threshold setting via soft class assignment, providing a detector for use in development of adaptive algorithms.%
-}
+The simulated results provide insight into configuring the proposed detector, which is compared to a conventional STFT-based spectral entropy detector using manually annotated humpback whale (\textit{Megaptera novaeangliae}) songs recorded in False Bay, South Africa, July 2021.
+
+The proposed method shows a significant improvement in detection accuracy and specificity, while also providing a more interpretable detection threshold setting via soft class assignment, providing a detector for use in development of adaptive algorithms.
 
@@
 
-		\node (cs)[right=0.3cm of Sigmoid] {\scriptsize $c_s(H[m])$};
+		\node (cs)[right=0.3cm of Sigmoid] {\scriptsize $p_\text{soft}(H[m])$};
 
@@
 		\draw[->] (MF.south) -- (KM.north);
-		\draw[<-]  (Sigmoid.south) |- (KM.east) node[above, near end]{\scriptsize $\mu_S$, $\mu_{NS}$};
+		\draw[<-]  (Sigmoid.south) |- (KM.east) node[above, near end]{\scriptsize $\mu_s$, $\mu_n$};
 
@@
 	\includegraphics[width=0.6\textwidth]{fbdata1}
-	\caption{Audio snippet and spectrogram of a humpback song recorded in False Bay.
+	\caption[Audio snippet and spectrogram of a humpback song recorded in False Bay.]{Audio snippet and spectrogram of a humpback song recorded in False Bay.
 The waveform shown is high-passed at 100 Hz, resampled and normalized.
@@
 	\includegraphics[width=0.7\textwidth]{real_data_hist}
-	\caption{Histograms indicating the combined call length and distance between calls of the two humpback songs, for use in determining median filter window length limits.}
+	\caption[Humpback call length and time spread.]{Histograms indicating the combined call length and distance between calls of the two humpback songs, for use in determining median filter window length limits.}
 	\label{fig:fbhist}
@@
 	\includegraphics[width=0.7\textwidth]{cwt_fb_example}
-	\caption{An example of a CWT filterbank, shown as the L1 and L2 normalized versions , with $\gamma = 10$, $\beta=50$ and 4 voices per octave.}
+	\caption[An example of a CWT filterbank.]{An example of a CWT filterbank, shown as the L1 and L2 normalized versions , with $\gamma = 10$, $\beta=50$ and 4 voices per octave.}
 	\label{fig:cwt_fb_example}
@@
 
-Given the class means $\mu_S$ (signal), $\mu_{NS}$ (no signal) determined by k-means clustering, the k-means decision boundary $\beta$ is the midpoint:
+Given the class means $\mu_s$ (signal), $\mu_n$ (no signal) determined by k-means clustering, the k-means decision boundary $\beta$ is the midpoint:
 \begin{equation*}
-	\beta = \mu_S + \frac{\mu_{NS} - \mu_S}{2}
+	\beta = \mu_s + \frac{\mu_n - \mu_s}{2}
 \end{equation*}
 
-In classical K-means, the assigned class $c_h(H[m])$ results in a hard output:
+In classical K-means, the assigned class $p_\text{hard}(H[m])$ results in a hard output:
 \begin{equation*}
-	c_h(H[m]) = \begin{cases}
+	p_\text{hard}(H[m]) = \begin{cases}
 		1, & H[m] \leq \beta \\
@@
 
-To provide a soft output, a function which smoothly transitions from one class to another in the interval $H[m] \in [\mu_S, \mu_{NS}]$ is required.
-Since an entropy measurement which is lower than $\mu_S$ is more probable to contain signal and, conversely, an entropy measurement which is greater than $\mu_{NS}$ is more probable to be noise, the transition function $\alpha(t)$ should satisfy the following criteria:
+To provide a soft output, a function which smoothly transitions from one class to another in the interval $H[m] \in [\mu_s, \mu_n]$ is required.
+Since an entropy measurement which is lower than $\mu_s$ is more probable to contain signal and, conversely, an entropy measurement which is greater than $\mu_n$ is more probable to be noise, the transition function $\alpha(t)$ should satisfy the following criteria:
 \begin{enumerate}
@@
 A higher gain will result in a steeper shape, thus attributing more certainty to entropy measurements closer to the class means, and vice versa for lower gains.
-It may also be beneficial to normalize the entropy measure such that $H[m] = \mu_S$ and $H[m] = \mu_{NS}$ map to $\pm1$.
+It may also be beneficial to normalize the entropy measure such that $H[m] = \mu_s$ and $H[m] = \mu_n$ map to $\pm1$.
 The normalization will ensure consistency in the behaviour of the hyper parameter $g$ over a range of class means.
@@
 \begin{equation*}
-	c_{s}(H[m]) = \sigma\left(	-2g	\frac{H[m] - \beta}{\mu_{NS} - \mu_S}	\right).
+	c_{s}(H[m]) = \sigma\left(	-2g	\frac{H[m] - \beta}{\mu_n - \mu_s}	\right).
 \end{equation*}
 
-To assign a specific pseudo-probability $c_{s}(H[m]) = p$ that corresponds to signal presence when $H[m] = \mu_S$, then the gain hyper parameter $g$ is calculated as
+To assign a specific pseudo-probability $c_{s}(H[m]) = p$ that corresponds to signal presence when $H[m] = \mu_s$, then the gain hyper parameter $g$ is calculated as
 \begin{equation}
 	\label{eqn:sigmoidg}
-	g = \frac{1}{2} \frac{\mu_{NS} - \mu_S}{\mu_S - \beta} \ln\left(\frac{1}{p} - 1\right)
+	g = \frac{1}{2} \frac{\mu_n - \mu_s}{\mu_s - \beta} \ln\left(\frac{1}{p} - 1\right)
 \end{equation}
@@
 	\includegraphics[width=0.6\textwidth]{sigmoid}
-	\caption{Example of the sigmoid function for soft output classification for various gains ($\mu_S=2$, $\mu_{NS}=4$).}
+	\caption[Example of the sigmoid function.]{Example of the sigmoid function for soft output classification for various gains ($\mu_s=2$, $\mu_n=4$).}
 	\label{fig:sigmoid}
@@
 \begin{equation*}
-	y[n] = \text{med}(x[n - R], ..., x[0], ..., x[n + R] ),
+	y[n] = \text{median}(x[n - R], ..., x[0], ..., x[n + R] ),
 \end{equation*}
@@
 	\includegraphics[width=0.6\textwidth]{entropymf}
-	\caption{Median filtered entropy (STFT) with $M=355$.}
+	\caption[Median filtered entropy.]{Median filtered entropy (STFT) with $M=355$.}
 	\label{fig:entropymf}
@@
 	\includegraphics[width=0.6\textwidth]{kmeans1}
-	\caption{The resulting class means from the K-means algorithm on median filtered entropy with $M=355$.}
+	\caption[K-means class means on entropy.]{The resulting class means from the K-means algorithm on median filtered entropy with $M=355$.}
 	\label{fig:kmeans1}
@@
 
-The soft-output classification function $c_s(H[m])$ is used to determine the soft true positive (STPR) and true negative (STNR) discovery rates:
+The soft-output classification function $p_\text{soft}(H[m])$ is used to determine the soft true positive (STPR) and true negative (STNR) discovery rates:
 \begin{equation*}
-	\text{STPR} = \frac{1}{|S_x|} \sum_{n \in S_x} c_s(H[m])
+	\text{STPR} = \frac{1}{|S_x|} \sum_{n \in S_x} p_\text{soft}(H[m])
 \end{equation*}
 \begin{equation*}
-	\text{STNR} = \frac{1}{N - |S_x|} \sum_{n \notin S_x} 1 - c_s(H[m])
+	\text{STNR} = \frac{1}{N - |S_x|} \sum_{n \notin S_x} 1 - p_\text{soft}(H[m])
 \end{equation*}
 
-This is equivalent to calculating the conventional TPR and TNR discovery rates if $c_h(H[m])$ is used instead of $c_s(H[m])$.
+This is equivalent to calculating the conventional TPR and TNR discovery rates if $p_\text{hard}(H[m])$ is used instead of $p_\text{soft}(H[m])$.
 
@@
  	\includegraphics[width=0.6\textwidth]{res_snr_mf}
- 	\caption{STFT and CWT soft discovery rates over a range of SNR's and median filter window lengths.
+ 	\caption[STFT and CWT soft discovery rates over SNR.]{STFT and CWT soft discovery rates over a range of SNR's and median filter window lengths.
 The left-hand side of the figure shows the STPR and STNR of the STFT respectively.
@@
 	\includegraphics[width=0.6\textwidth]{res_mf_snr}
-	\caption{STFT and CWT soft discovery rates over a range median filter window lengths for low SNRs. The left and right sides of the figure show the results for the STFT and CWT respectively.}
+	\caption[STFT and CWT soft discovery rates over median filter length.]{STFT and CWT soft discovery rates over a range median filter window lengths for low SNRs.
+The left and right sides of the figure show the results for the STFT and CWT respectively.}
 	\label{fig:res_mf_snr}
@@
 	\includegraphics[width=0.6\textwidth]{res_classmeans}
-	\caption{STFT and CWT difference in class means ($\mu_{NS} - \mu_S$) over a range of SNRs and median filter window lengths.}
+	\caption[STFT and CWT difference in class means.]{STFT and CWT difference in class means ($\mu_n - \mu_s$) over a range of SNRs and median filter window lengths.}
 	\label{fig:res_classmeans}
@@
 	\includegraphics[width=0.6\textwidth]{res_snr_mf_L2}
-	\caption{Comparing L1 and L2 CWT normalization for spectral entropy calculation.
+	\caption[Comparing L1 and L2 CWT normalization for spectral entropy calculation.]{Comparing L1 and L2 CWT normalization for spectral entropy calculation.
 The left side of the figure shows the results without MF ($M=1$).
@@
 	\includegraphics[width=1\textwidth]{res_realdata_snippet1}
-	\caption{A snippet showing the measures produced by the entropy and the proposed soft-output signal detector.}
+	\caption[A snippet showing the measures produced by the proposed entropy detector.]{A snippet showing the measures produced by the entropy and the proposed soft-output signal detector.}
 	\label{fig:res_realdata_snippet1}
@@
 	\includegraphics[width=0.5\textwidth]{res_realdata_roc}
-	\caption{The ROC curves comparing the proposed technique and the baseline STFT SE.
+	\caption[The ROC curves comparing the proposed technique and the baseline STFT SE.]{The ROC curves comparing the proposed technique and the baseline STFT SE.
 Median filter window lengths of 1, 201 and 501 are shown.
@@
 	\includegraphics[width=0.55\textwidth]{res_realdata_thresh}
-	\caption{The thresholds used to produce the ROC curve and the TPR/FPR of the baseline SE and the proposed technique.
+	\caption[The thresholds used to produce the ROC curve.]{The thresholds used to produce the ROC curve and the TPR/FPR of the baseline SE and the proposed technique.
 Median filter window length of 1, 201 and 501 are shown on the right.
@@
 In practice, the detector should be tuned such that the FPR is below 5\%.
-From figure \ref{fig:res_realdata_thresh}, this corresponds to a threshold of $c_s(H[m]) > 0.1$ for $p=0.95$.
+From figure \ref{fig:res_realdata_thresh}, this corresponds to a threshold of $p_\text{soft}(H[m]) > 0.1$ for $p=0.95$.
 This threshold will vary depending on the choice of $p$.
@@
 
-\changes{The addition of the k-means algorithm presents another advantage: utilizing the class-mean separation ($\mu_{NS} - \mu_S$) as metric to quickly determine whether there are points of interest within an audio segment.
+\changes{The addition of the k-means algorithm presents another advantage: utilizing the class-mean separation ($\mu_n - \mu_s$) as metric to quickly determine whether there are points of interest within an audio segment.
 Furthermore, the k-means algorithm may be restricted to converge within a set number of iterations.
@@
 
+\pagebreak
 \begin{subappendices}
@@
 
-	Given the class means $\mu_S$, $\mu_{NS}$ and the entropy measure $H[m]$ of length $N$, the assignment function assigns the class with the closest mean:
+	Given the class means $\mu_s$, $\mu_n$ and the entropy measure $H[m]$ of length $N$, the assignment function assigns the class with the closest mean:
 
@@
 		w[m] = \begin{cases}
-			1, & (H[m] - \mu_S)^2 \leq (H[m] - \mu_{NS})^2 \\
+			1, & (H[m] - \mu_s)^2 \leq (H[m] - \mu_n)^2 \\
 			0, & \text{otherwise}
@@
 		\begin{equation*}
-			\mu_S = \min\limits_{m}\left(H[m]\right);
+			\mu_s = \min\limits_{m}\left(H[m]\right);
 		\end{equation*}
 		\begin{equation*}
-			\mu_{NS} = \max\limits_{m}\left(H[m]\right).
+			\mu_n = \max\limits_{m}\left(H[m]\right).
 		\end{equation*}
@@
 			\begin{equation*}
-				\mu_S = \frac{\sum_{m=0}^{N-1} w[m] H[m]}{\sum_{m=0}^{N-1} w[m] };
+				\mu_s = \frac{\sum_{m=0}^{N-1} w[m] H[m]}{\sum_{m=0}^{N-1} w[m] };
 			\end{equation*}
 			\begin{equation*}
-				\mu_{NS} = \frac{\sum_{m=0}^{N-1} (1-w[m]) H[m]}{\sum_{m=0}^{N-1} 1 - w[m] }.
+				\mu_n = \frac{\sum_{m=0}^{N-1} (1-w[m]) H[m]}{\sum_{m=0}^{N-1} 1 - w[m] }.
 			\end{equation*}

### 2024-05-27 dissertation/content/paper2/main.tex
started intro to p3
added astract and conclusion
draft edit nearly complete, need to do refs

@@
 \chapter{Introducing Wavelet Scattering}
-
-\articleinfo{M.W.
-Rademan, D.J.J Verseld, J.A.
-Du Preez}{the Journal of the Acoustical Society of America (JASA)}{15 May 2024}{doi}{Current literature trends for whale classification in passive acoustic monitoring systems are mainly concerned with high accuracies obtained via deep learning approaches.
+\label{chap:p2}
+
+\articleinfo{Detecting and classifying Whale Calls with Wavelet Scattering
+and Spectral Entropy}{the Journal of the Acoustical Society of America (JASA)}{15 May 2024}{doi}
+
+\section*{CRediT}
+\textbf{M.W. Rademan}: conceptualisation; methodology; software; formal analysis; writing -- original draft.
+
+\textbf{D.J.J. Verseld}: writing -- review and editing; supervision
+
+\textbf{J.A. Du Preez}: writing -- review and editing; supervision.
+
+\section*{Abstract}
+Current literature trends for whale classification in passive acoustic monitoring systems are mainly concerned with high accuracies obtained via deep learning approaches.
 This requires preparation of large and accurate annotated datasets, which are often not available when surveying new regions, nor are they wholly accurate when they are.
@@
 The purpose of this work is to provide a practical classification and detection system for new regions in which very little annotated data is available.
-}
 
@@
     \includegraphics[width=0.9\textwidth]{system_diagram_v2.png}
-    \caption{The detection and classifying system.
+    \caption[The detection and classifying system.]{The detection and classifying system.
 The audio database constists of 1-hour long recordings, which are segmented for computational efficiency.
@@
 \end{figure}
-
-\clearpage
 
@@
      \end{subfigure}
-        \caption{High SNR examples of the first level scattering transform to illustrate the characteristics of blue whale calls.
+        \caption[High SNR examples of the first level scattering transform to illustrate the characteristics of blue whale calls.]{High SNR examples of the first level scattering transform to illustrate the characteristics of blue whale calls.
 Note the logarithmic frequency scale.
@@
     \centering
-\caption{Estimated erroneous (false positives) proportion of annotations per class that could be considered noise.
+\caption[Estimated erroneous (false positives) proportion of annotations per class that could be considered noise.]{Estimated erroneous (false positives) proportion of annotations per class that could be considered noise.
 This table does not reflect on whether annotations are of the correct class (i.e., misclassifications).
@@
 
-\clearpage
-
 \section{Wavelet Scattering}
@@
     \includegraphics[width=0.9\textwidth]{scattering.png}
-    \caption{An illustration of the successive filtering operations performed by a scattering transform.
+    \caption[An illustration of the successive filtering operations performed by a scattering transform.]{An illustration of the successive filtering operations performed by a scattering transform.
 Nodes in colour represent the extracted scattering coefficients, which is down-sampled according to an invariance scale specified by the user.}
@@
     \includegraphics[width=1\textwidth]{fb.jpg}
-    \caption{The filter banks used in this study, with $Q_1 = 16$ and $Q_2 = 4$.
+    \caption[The filter banks used in this study.]{The filter banks used in this study, with $Q_1 = 16$ and $Q_2 = 4$.
 The filter bank used to calculate $\mathcal{U}_1$ is shown on the left and the $\mathcal{U}_2$ filter bank on the right.
@@
 
-\clearpage
 \section{Improved Spectral Entropy Detector}
@@
     \includegraphics[width=1\textwidth]{detector_example_2.eps}
-    \caption{An example of the output of the detector taken from a densely populated vocalisation region of the dataset.
+    \caption[An example of the output of the detector.]{An example of the output of the detector taken from a densely populated vocalisation region of the dataset.
 The annotation bounding boxes are also indicated (top).
@@
      \end{subfigure}
-     \caption{Results of the evaluated detectors.
+     \caption[ROC curves of the evaluated detectors.]{Results of the evaluated detectors.
 \ac{se} detectors are indicated as a solid line, while \ac{bled} is dashed.
@@
 
-\clearpage
 \section{Classification}
@@
 \begin{table}[h!]
-\caption{Number of labeled detections for all vocalisations of interest.
+\caption[Number of labeled detections for all vocalisations of interest.]{Number of labeled detections for all vocalisations of interest.
 The detections are produced by the whitened \ac{gmm} \ac{se} detector.
@@
 % \renewcommand{\arraystretch}{0.8}
+\centering
 \begin{tabular}{lc} 
@@
 \begin{table}[]
-\caption{Average model classification accuracies for varying train-test splits with chi-squared feature selection.
+\caption[Average model classification accuracies for varying train-test splits with chi-squared feature selection.]{Average model classification accuracies for varying train-test splits with chi-squared feature selection.
 An 80\% training split without chi-square feature selection is included as a baseline.
@@
 
-\clearpage
 \section{Conclusion}
@@
 
-    \clearpage
-
     \section{Summary of parameter values}
@@
     \begin{table}[h]
-        \caption{Summary of parameter values used in this study. Chosen values are indicated in boldface text, whereas non-bold parameters result from the chosen parameters.}
+        \caption[Summary of parameter values used in this study.
+]{Summary of parameter values used in this study.
+Chosen values are indicated in boldface text, whereas non-bold parameters result from the chosen parameters.}
         \label{tab:parameters}
         % \begin{ruledtabular}    
-            \renewcommand{\arraystretch}{0.8}% Tighter
-            \begin{tabular}{lll}  \small
+            % \renewcommand{\arraystretch}{0.8}% Tighter
+            \begin{tabular}{lll} 
                 \textbf{Parameter}      & \textbf{Description} & \textbf{Value} \\ \hline

### 2024-05-27 dissertation/content/paper3/main.tex
started intro to p3
started intro
draft edit nearly complete, need to do refs

@@
 \chapter{Separable Wavelet Scattering}
+\label{chap:p3}
 
-\articleinfo{M.W.
-Rademan, D.J.J Verself, J.A.
-Du Preez}{IEEE Signal Processing Letters}{15 May 2024}{doi}{Wavelet scattering is a widely used feature extraction method due to it efficacy in extracting invariant features,
-while retaining any lost high frequency information resulting from averaging to obtain the desired amount of invariance. 
-By generalising a 1-dimensional scattering transform, we extend its definition to an arbitrary number of independent dimensions. 
-We show that, in a modern deep learning setting, separable wavelet scattering performs similarly to its non-separable counterparts 
-with the MNIST hand-written digit dataset. We additionally demonstrate state-of-the-art results for a subset of the MedMNIST3D datasets.}
+\articleinfo{Seperable Wavelet Scattering}{IEEE Signal Processing Letters}{15 May 2024}{doi}
+
+\section*{CRediT}
+\textbf{M.W. Rademan}: conceptualisation; methodology; software; formal analysis; writing -- original draft.
+
+\textbf{D.J.J. Verseld}: writing -- review and editing; supervision.
+
+\textbf{J.A. Du Preez}: writing -- review and editing; supervision.
+
+\section*{Abstract}
+Wavelet scattering is a widely used feature extraction method due to it efficacy in extracting invariant features.
+It retains lost high frequency information resulting from the averaging operations required to obtain invariance.
+Different wavelet scattering transforms exist for a varying number of input dimensions, each having different choices of wavelets and filterbank construction techniques.
+In this work, we propose a new scattering transform by generalising the one-dimensional case to an arbitrary number of independent dimensions.
+Additionally, we provide concise implementation definitions for the Morlet wavelet.
+We show that, in a modern deep learning setting, separable wavelet scattering performs similarly to its non-separable counterparts with the MNIST hand-written digit dataset.
+We additionally demonstrate state-of-the-art results for a subset of the MedMNIST3D datasets.
 
@@
 It is an especially effective method for small datasets, since the feature filters are reminiscent of a \ac{cnn} front-end, while requiring no learning.
-The resulting features are invariant across all averaged dimensions, and exhibit separable class subspaces, allowing linear classifiers to be used with high efficacy \citep{2dscattering}.
+The resulting features are invariant across all averaged dimensions, and exhibit separable class subspaces, allowing linear classifiers to be used with high efficacy \cite{2dscattering}.
 
-Various forms of \ac{ws} exist, with \ac{1d} scattering first proposed by Anden and Mallat \citep{1dscattering1, ws}, which was later extended to 2 \citep{2dscattering} and 3 dimensions \citep{3dscattering, harmonicscattering}.
-Additional and more specialised forms for \ac{ws} include rotation-invariant scattering \citep{groupinvariantscattering} for the classification of textures and joint-\ac{tf} scattering \citep{ws_joint_tf, jointtfscattering2}.
+Various forms of \ac{ws} exist, with \ac{1d} scattering first proposed by Anden and Mallat \cite{1dscattering1, ws}, which was later extended to 2 \cite{2dscattering} and 3 dimensions \cite{3dscattering, harmonicscattering}.
+Additional and more specialised forms for \ac{ws} include rotation-invariant scattering \cite{groupinvariantscattering} for the classification of textures and joint-\ac{tf} scattering \cite{ws_joint_tf, jointtfscattering2}.
 Joint-\ac{tf} scattering utilises a separable \ac{2d} filter that operates on the first level scattering scalogram, denoted by the operator $\mathcal{U}_1$, which is the only apparent usage of separable filters in the current scattering literature.
 
-In a deep-learning setting, sensible initialisation of filters prior to learning can significantly improve performance and interpretability \citep{sincnet}.
-The greatest advantage of separable filters is their computation speed when computing convolutions directly, as is performed in \acp{cnn} \citep{separablecnn}.
+In a deep-learning setting, sensible initialisation of filters prior to learning can significantly improve performance and interpretability \cite{sincnet}.
+The greatest advantage of separable filters is their computation speed when computing convolutions directly, as is performed in \acp{cnn} \cite{separablecnn}.
 Additionally, separable filters reduce the number of parameters of a \ac{cnn} if the filters are configured to be learnable.
-Learnable filters can typically improve performance compared to their fixed-filters counterpart \citep{scattering_birdsong}.
+Learnable filters can typically improve performance compared to their fixed-filters counterpart \cite{scattering_birdsong}.
 
-At first glance, a \ac{ws} decomposition seems to be prohibitively expensive, but fast algorithms are possible due to the limited bandwidth of filter output.
-Fast algorithms utilise downsampling to take advantage of the demodulation of band-limited signals that result from the convolution of analytic wavelets \citep{2dscattering, 3dscattering, 1dscattering1}.
+At first glance, a \ac{ws} decomposition seems to be prohibitively expensive, but fast algorithms are possible due to the limited bandwidth of the filter output.
+Fast algorithms utilise downsampling to take advantage of the demodulation of band-limited signals that result from the convolution of analytic wavelets \cite{2dscattering, 3dscattering, 1dscattering1}.
 Path pruning is also used to ignore filter combinations which have negligible energy.
-These algorithms are the standard implementation on many platforms, such as the Kymatio python package \citep{kymatio} and MATLAB \citep{MATLAB}.
+These algorithms are the standard implementation on many platforms, such as the Kymatio python package \cite{kymatio} and MATLAB \cite{MATLAB}.
 
@@
 
-\subsection{Morlet}
+\subsection{Morlet Wavelet}
 
-We define a \ac{1d} zero-mean Gaussian as
+A \ac{1d} zero-mean Gaussian is given by
 \begin{equation}
@@
 For an arbitrary bandwidth definition, we define the bandwidth-to-$\sigma$ ratio $\beta \in \mathbb{R}^+$, such that the one-sided bandwidth of a zero-mean Gaussian is $\beta \sigma_\omega$.
-For reference, Kymatio's implementation of 2D scattering indirectly defines $\beta \approx 2.5$ \citep{kymatio}.
+For reference, Kymatio's implementation of 2D scattering indirectly utilises $\beta \approx 2.5$ \cite{kymatio}.
 
@@
 
-A \ac{1d} Morlet $\uppsi$ has energy concentrated around 1 rad/s, with most of its energy contained in the interval $\omega \in [1 - \frac{1}{Q}, 1 + \frac{1}{Q}]$.
+A \ac{1d} Morlet wavelet $\uppsi$ has energy concentrated around 1 rad/s, with most of its energy contained in the interval $\omega \in [1 - \frac{1}{Q}, 1 + \frac{1}{Q}]$.
 $Q \in \mathbb{R}^+$ is defined as the number of wavelets per octave.
@@
         \uppsi(\lambda t), \ \lambda \neq 0 \\
-        \phi_j(t), \ \lambda = 0
+        \phi_i(t), \ \lambda = 0
     \end{cases},
 \end{equation}
-where $\phi$ represents the 1D \ac{lpf} utilised in scattering computations for the current dimension ($j$).
+where $\phi$ represents the 1D \ac{lpf} utilised in scattering computations for the current dimension ($i$).
 
-A $n$-dimensional separable wavelet may be constructed similarly, with
+An $n$-dimensional separable wavelet may be constructed similarly, with
 \begin{equation}
@@
 \label{sec:filterbank}
- Suppose a 1D Morlet has a centre frequency $\lambda_0$.
+ Suppose a 1D Morlet wavelet has a centre frequency $\lambda_0$.
 The following filter is placed at $\lambda_0 2^{\frac{1}{Q}}$.
-To define the amount of overlap between filters, it is useful express the corresponding filter's frequency \ac{std} $\sigma_\omega$ in terms of the distance between filters in the frequency domain.
-If the wavelet at $\lambda_0$ requires a decay equivalent to $\alpha \sigma_\omega$ \ac{std}s at the next wavelet at $\lambda_0 2^{\frac{1}{Q}}$.
+To define the amount of overlap between filters, it is useful to express the corresponding filter's frequency \ac{std} $\sigma_\omega$ in terms of the distance between filters in the frequency domain.
+Suppose a wavelet at centre frequency $\lambda_0$ requires a decay equivalent to $\alpha \sigma_\omega$ \ac{std}s at the next wavelet at $\lambda_0 2^{\frac{1}{Q}}$.
 It follows that $\sigma_\omega = \frac{1}{\sigma_t} = \frac{1}{\alpha}\left( 2^\frac{1}{Q} - 1 \right)$.
@@
 
-The provided definitions of filter clearance ($\alpha$) and bandwidth ($\beta$) may also be extended to non-separable wavelets, such as the rotational construction employed in \citep{2dscattering}.
-However, a conversion is required to find the bandwidth in each axis for a specific wavelet for rotationally constructed filterbanks, in order to employ the downsampling strategies proposed in this paper.
+The provided definitions of filter clearance ($\alpha$) and bandwidth ($\beta$) may also be extended to non-separable wavelets, such as the rotational construction employed in \cite{2dscattering}.
+Other scattering transforms may also benefit from the downsampling strategies proposed in this paper.
 
@@
 The scattering transform requires that the time support of all filters do not exceed the time support of $\phi$, ensuring no filters contain time information exceeding the invariance scale $d$.
-$d$ is equivalently defined as the total downsample factor of the scattering tranform, and is unique for each dimension.
+The invariance scale $d$ is equivalently defined as the total downsample factor of the scattering tranform, and is unique for each dimension.
 Equivalently, the frequency \ac{std} of the filters may not exceed the \ac{std} of $\phi(\omega)$.
@@
 On-axis coverage is also required, in which each $\uppsi_\lambda$ must be multiplied with a Gaussian (zero-frequency wavelet), which the dilated wavelet definition in equation (\ref{eqn:dilwav}) defines as $\lambda=0$.
-A similar construction procedure is followed in \citep{jointtfscattering2}.
+A similar construction procedure is followed in \cite{jointtfscattering2}.
 
-Given $m$ 1D filterbanks, with $m \ge 2$, each having Morlet filters with a positive set of lambdas $\Lambda_i^+$ and invariance scales $\vect{d} \in \mathbb{N}^m$, with $i = 1, ..., m$ indexing the dimension, we construct the $m$-dimensional filterbank with
+Given $m$ 1D filterbanks, with $m \ge 2$, each having Morlet filters with a positive set of centre frequencies $\Lambda_i^+$ and invariance scales $\vect{d} \in \mathbb{N}^m$, with $i = 1, ..., m$ indexing the dimension, we construct the $m$-dimensional filterbank with
 \begin{equation}
@@
 
-% The dilated wavelet definition provided by (\ref{eqn:dilwav}) therefore provides a concise set construction, with which to construct the \ac{lpf} and on-axis wavelets simultaneously.
-
-% All on-axis filters will be highly correlated with their negative counterpart due to the symmetry of the multidimensional Fourier transform of real signals.
-For example, the 2D filter specified by $\lambda = [0, x]^T$ will be highly correlated with the filter at $[0, -x]^T$.
-As the invariance scale $d$ increases, the on-axis filters will become more narrow in the zero coordinate dimension, thereby becoming increasingly correlated.
-
-% \subsection{Discretisation}
-
-% Morlets do not have compact support, as such, they must be truncated for computation.
-The invariance scale $d$ specifies the maximum allowable time support of all filters.
-Since input signals must be at least of length $d$, filters can be precomputed using the padded signal length, while also guaranteeing a suitable level of truncation.
-The padding required is on the order of $d$ samples.
-
-% Note that, due to their separability, all $\uppsi_{\vect{\lambda}} \in \mathbb{F}$ can be stored as $d$ one-dimensional filters, and not as $N_1 \times ...
-\times N_d$ tensors, thereby saving on the storage required for pre-computed filters.
-The savings on storage space allow implementations to compute the filters at multiple input sample frequencies, allowing for optimal downsampling strategies without incurring a significant storage cost.
-
 \section{Separable Scattering Transform} \label{sec:wst}
 % In this section, we describe the efficient computation of $m$-dimensional separable scattering transforms.
-The described downsampling strategy is utilised to a limited extent in Kymatio scattering implementations \citep{kymatio} and the original proposed fast algorithms \citep{2dscattering}, which is generalised further in this section.
+The described downsampling strategy is utilised to a limited extent in Kymatio scattering implementations \cite{kymatio} and the original proposed fast algorithms \cite{2dscattering}, which is generalised further in this section.
 
@@
 
-A downsample factor $d$ which decomposes into a set of $n$ prime factors $\{p_1, ..., p_n\}$ with a corresponding multiset $\mathbb{M} = \left\{ m_1, ..., m_n \right\}$, where $m_i$ is the multiplicity of the prime $p_i$.
-We can find an optimal $d$ by maximising the sum $\sum\limits_{m \in M} m$.
+A downsample factor $d$ decomposes into a set of $n$ prime factors $\{p_1, ..., p_n\}$ with a corresponding multiset $\mathbb{M} = \left\{ m_1, ..., m_n \right\}$, where $m_i$ is the multiplicity of the prime $p_i$.
+We can find an optimal $d$ by maximising the sum $\sum\limits_{m \in \mathbb{M}} m$.
 
@@
 such that $d = d_1 \cdot d_2, \ d_1, d_2 \in \mathbb{N}^+$, with $d_1 | d$ and $d_1 \le  d_{\uppsi_1}$.
+The notation $x | y$ indicates that $x$ evenly divides $y$.
 In order to find $d_1$, we decrement $d_{\uppsi_1}$ until it divides $d$ evenly.
@@
 
-% Since the filterbank configuration is known, all required combinations of downsampled filters and their downsampling factors can be precomputed.
+\subsection{Convolutions}
+Optimal \ac{fft} convolutions can be achieved by performing downsampling in the frequency domain instead of the time domain.
+It is straightforward to verify that $|x * \uppsi|_{\downarrow r} = \left|(x * \uppsi)_{\downarrow r}\right|$, since the modulus is an element-wise operation, for some downsampling factor $r$.
+Given the signal and filter Fourier transforms $\hat{x}$ and $\hat{\uppsi}$, we then have
+\begin{equation}
+\label{eqn:circconv}
+    x \otimes \uppsi [n] \xleftrightarrow{\mathcal{FFT}}  \hat{x} \cdot \hat{\uppsi} [k],
+\end{equation}
+where the $\otimes$ operator represents a circular convolution and $k$ is the frequency index.
 
-\begin{table*}[t!]
+Given that $r|N$, we can express (\ref{eqn:circconv}) when downsampled as a periodised summation \cite{waveletsandsubbandcoding} in the frequency domain
+\begin{equation}
+    (x \otimes \uppsi [n])_{\downarrow r} \xleftrightarrow{\mathcal{FFT}}  \frac{1}{r} \sum_{i=0}^{r-1} \hat{x} \cdot \hat{\uppsi} [k + iN/r], %https://citeseerx.ist.psu.edu/document?repid=rep1&type=pdf&doi=5216ea733e562541b33a7f97dab0de072b2e8827
+\end{equation}
+which can be efficiently implemented via shape manipulation of tensors in computational packages like MATLAB or PyTorch \cite{pytorch}.
+To compute valid convolutions, we must pad $x$ and $\uppsi$ to have a total length of $N = N_x + d + c$, where $c \in \mathbb{N}^+$ is a constant that ensures that $d | N$.
+
+\section{Results}
+
+\subsection{MNIST}
+
+% We benchmark separable \ac{ws} against 2D scattering by repeating the MNIST hand-written digit dataset experiment in \cite{2dscattering}.
+
+The MNIST dataset \cite{mnist} contains 60000 training and 10000 test samples.
+Unlike in \cite{2dscattering}, which decorrelates scattering coefficients with a discrete cosine transform prior to classification, we perform classification on the scattering coefficients directly.
+Kymatio \cite{kymatio} is used to produce the 2D scattering coefficients.
+Our implementation of separable scattering is implemented similarly to Kymatio, with PyTorch \cite{pytorch} as a backend for FFT convolutions and \ac{nn} models.
+
+Unless specified otherwise, all experiments have $\beta = \alpha = 2.5$.
+Scattering features are normalised prior to classification, according to the mean and \ac{std} calculated on the training set.
+No data augmentation is performed.
+
+\begin{table}
 \centering
-\caption{MedMNIST classification results of compared to baseline \ac{nn} approaches (\citep{medmnist})} \label{tab:medmnist}
+\caption{MNIST classification error rate (\%) of separable and \ac{2d} scattering coefficients using a NN classifier.} \label{tab:mnistnn}
+\begin{tabular}{|lr|l|} \hline
+\multirow{2}{*}{2D WS + NN}        & $l=1$, $J=2$  & $0.64 \pm 0.05$    \\
+          & $l=2$, $J=3$ & $0.50 \pm 0.03$     \\ \hline
+\multirow{2}{*}{Separable WS + NN}  & $l=1$, $d=(4,4)$ & $0.63 \pm 0.05 $    \\
+          & $l=2$, $d=(4,4)$  & $0.52 \pm 0.04$     \\ \hline
+\end{tabular}
+\end{table}
+
+Due to its separability and non-angularly spaced filters, separable scattering does not perform as well compared to 2D scattering when using simple classifiers, such as \ac{lda} \cite{lda}.
+To illustrate that this performance discrepancy is not of significant consequence in a \ac{nn} setting, we test performance on the full dataset utilising a simple architecture.
+The neural network architecture used has an input layer with 256 neurons, followed by two hidden layers with 128 and 64 neurons respectively.
+The output layer has 10 neurons - one for each digit.
+Input and hidden layers are followed by a batch norm layer \cite{batchnorm} and ReLU activation function \cite{relu}.
+The output layer is followed by a softmax function.
+The Adam optimiser \cite{adam} is used with cross-entropy loss, a batch size of 256 and learning rate of $3 \cdot 10^{-5}$.
+5000 of the 60000 training samples are reserved for validation and removed from the training set.
+Training is stopped when validation loss starts to increase.
+The \ac{nn} model is initialised with random weights, and the experiment is repeated 50 times.
+Different invariance scales were tested, and the best results are reported in table \ref{tab:mnistnn}.
+Tests are repeated for $l \in \{1, 2\}$ levels of scattering.
+
+\subsection{MedMNIST3D}
+The MedMNIST3D datasets are a subset of the MedMNIST dataset group \cite{medmnist}, where each 3D dataset contains $28\times 28 \times 28$ images with 2, 3 or 11 classes.
+Train, test and validation data partitions are provided by the authors.
+All datasets have on the order of 1000 training samples.
+We compare the baseline \ac{nn} results provided in \cite{medmnist} with separable scattering features combined with a simple \ac{nn} classifier.
+
+We use \ac{nn} classifier with an input layer containing 1024 neurons, followed by two hidden layers with 512 and 256 neurons respectively.
+Input and hidden layers layers are each followed by a batch norm layer \cite{batchnorm} and a ReLU non-linearity \cite{relu}.
+For datasets with two classes, the output layer is a single neuron followed by a sigmoid activation function.
+For datasets with more than two classes, the output layer has a size equal to the number of classes, followed by a softmax activation.
+Binary cross-entropy loss are used for datasets with two classes, otherwise cross-entropy loss is used.
+The Adam optimiser \cite{adam} with a learning rate of $1 \cdot 10^{-5}$ is used.
+All other configuration parameters are identical to the model used for the MNIST dataset.
+
+A single level of scattering coefficients are computed, with $Q = (2, 2, 2)$ and $d = (4, 4, 4)$.
+Many MedMNIST3D datasets tend to be unbalanced, implying that \ac{auc} is a more reliable metric to measure model performance.
+The results are shown in table \ref{tab:medmnist}, in which \ac{sota} \ac{auc} performance is achieved for the Organ, Adrenal and Vessel datasets.
+Table \ref{tab:medmnist} indicates the number of classes ($c$) for each of the datasets, with accuracy (ACC) also shown for reference.
+Non-\ac{sota} results are comparable to the other baseline \ac{nn} approaches presented in \cite{medmnist}.
+It is likely that better results can be achieved by the proposed method if the filters are made learnable and/or scattering parameters and network architectures are uniquely optimised for each dataset.
+
+\begin{table*}[]
+    \centering
+    \caption[MedMNIST classification results of compared to baseline \ac{nn} approaches.]{MedMNIST classification results of compared to baseline NN approaches (\citep{medmnist}).} \label{tab:medmnist}
+    \begin{adjustbox}{angle=90}
 \begin{tabular}{|l|rr|rr|rr|rr|rr|rr|}
@@
 \end{tabular}
+\end{adjustbox}
 \end{table*}
-
-\subsection{Convolutions}
-Optimal \ac{fft} convolutions can be achieved by performing downsampling in the frequency domain instead of the time domain.
-It is straightforward to verify that $|x * \uppsi|_{\downarrow r} = \left|(x * \uppsi)_{\downarrow r}\right|$, since the modulus is an element-wise operation, for some downsampling factor $r$.
-Given the signal and filter Fourier transforms $\hat{x}$ and $\hat{\uppsi}$, we then have
-\begin{equation}
-\label{eqn:circconv}
-    x \otimes \uppsi [n] \xleftrightarrow{\mathcal{FFT}}  \hat{x} \cdot \hat{\uppsi} [k],
-\end{equation}
-where the $\otimes$ operator represents a circular convolution and $k$ is the frequency index.
-
-Given that $r|N$, we can express (\ref{eqn:circconv}) when downsampled as a periodised summation \citep{waveletsandsubbandcoding} in the frequency domain
-\begin{equation}
-    (x \otimes \uppsi [n])_{\downarrow r} \xleftrightarrow{\mathcal{FFT}}  \frac{1}{r} \sum_{i=0}^{r-1} \hat{x} \cdot \hat{\uppsi} [k + iN/r], %https://citeseerx.ist.psu.edu/document?repid=rep1&type=pdf&doi=5216ea733e562541b33a7f97dab0de072b2e8827
-\end{equation}
-which can be efficiently implemented via shape manipulation of tensors in computational packages like MATLAB or PyTorch \citep{pytorch}.
-To compute valid convolutions, we must pad $x$ and $\uppsi$ to have a total length of $N = N_x + d + c$, where $c \in \mathbb{N}^+$ is a constant that ensures that $d | N$.
-
-% Note that only $d$ additional samples are required instead of $2d$, since $d$ is, by definition, the time support of $\phi$ in number of samples.
-Padding is applied equally to the left and right of $x$.
-For further reduction of boundary effects, reflection padding is utilised.
-Kymatio's 2D scattering implementation utilises a similar padding scheme.
-
-% Thus far, the techniques discussed may be equivalently implemented on many scattering transforms.
-Separable scattering benefits from reduced computations resulting from filter separability.
-In particular, given a $m$-dimensional signal $x[\vect{n}]$ indexed by $\vect{n} = (n_1, ..., n_m)^T$, with each dimension having downsample factors expressed as a vector $\vect{r} = (r_1, ..., r_m)^T$, we can chain downsampling for each dimension:
-% \begin{equation}
-%     \left(x * \uppsi_{\vect{\lambda}}[\vect{n}]\right )_{\downarrow \vect{r}} = \left( \left(x[\vect{\lambda}] * \uppsi_{\lambda_i}[n_i]\right)_{\downarrow r_1} ...
-* \uppsi_{\lambda_m}[n_m]\right)_{\downarrow r_m}.
-% \end{equation}.
-
-% We abuse notation to indicate that $\downarrow r_i$ is an operator that downsamples the $i$'th dimension, and $\downarrow \vect{r}$ downsamples all dimensions each with their correponding downsample factor in $\vect{r} \in \mathbb{N}^m$.
-
-% For $m$-dimensional sequences of dimensional lengths $N_1, ..., N_m$, the forward and inverse \acp{fft} computation time is reduced from $2\sum_{i} T \log N_i$ (no frequency periodisation), with $T = \prod_i N_i$ to $\sum_{i} \frac{T}{K_i} \log \frac{N_i}{K_i} + P \log P$, where $K_1 = 1$, $K_i = \prod_{j < i} r_j$ and $P = \prod_i \frac{N_i}{r_i}$.
-However, the number of convolution multiplications in the frequency domain increases from $\prod_i N_i$ to $\sum_i \prod_j \frac{N_j}{K_i} \leq m \prod_i N_i$.
-
-% In contrast, optimal non-separable scattering transforms can achieve a computational cost of $\sum_{i} N_i \log N_i + P \log P$, since a full forward \ac{fft} is required to compute a convolution with a non-separable $m$-dimensional filter.
-
-% The greatest computational advantage is during direct convolution computations, as is present in \acp{cnn}.
-For an $m$-dimensional filter impulse response with filter lengths $L_1, ..., L_m$, the computational cost reduces from $P\prod_i L_i$ to $P \sum_i L_i$.
-
-\section{Results}
-
-\subsection{MNIST}
-
-% We benchmark separable \ac{ws} against 2D scattering by repeating the MNIST hand-written digit dataset experiment in \citep{2dscattering}.
-
-The MNIST dataset \citep{mnist} contains 60000 training and 10000 test samples.
-Unlike in \citep{2dscattering}, which decorrelates scattering coefficients with a discrete cosine transform prior to classification, we perform classification on the scattering coefficients directly.
-Kymatio \citep{kymatio} is used to produce the 2D scattering coefficients.
-Our implementation of separable scattering is implemented similarly to Kymatio, with PyTorch \citep{pytorch} as a backend for FFT convolutions and \ac{nn} models.
-
-Unless specified otherwise, all experiments have $\beta = \alpha = 2.5$.
-Scattering features are normalised prior to classification, according to the mean and \ac{std} calculated on the training set.
-No data augmentation is performed.
-
-\begin{table}
-\centering
-\caption{MNIST classification error rate (\%) of separable and \ac{2d} scattering coefficients using a \ac{nn} classifier} \label{tab:mnistnn}
-\begin{tabular}{|lr|l|} \hline
-\multirow{2}{*}{2D WS + NN}        & $l=1$, $J=2$  & $0.64 \pm 0.05$    \\
-          & $l=2$, $J=3$ & $0.50 \pm 0.03$     \\ \hline
-\multirow{2}{*}{Separable WS + NN}  & $l=1$, $d=(4,4)$ & $0.63 \pm 0.05 $    \\
-          & $l=2$, $d=(4,4)$  & $0.52 \pm 0.04$     \\ \hline
-\end{tabular}
-\end{table}
-
-% Table \ref{tab:mnistlda} shows the results for MNIST using a \ac{lda} classifier from the Scikit-learn python package \citep{sklearn}, with a covariance shrinkage factor of $5\cdot 10^{-3}$.
-Scattering is performed for $l=1, 2$ levels.
-
-% \begin{table}[!h] 
-% \centering
-% \caption{MNIST classification error rate (\%) of separable and \ac{2d} scattering coefficients  using an \ac{lda} classifier} \label{tab:mnistlda}
-% \begin{tabular}{|r|rr|rr|} \hline
-% \multirow{2}{*}{Training size}              & \multicolumn{2}{l}{2D}                               & \multicolumn{2}{l|}{Separable}                             \\
-%  & \multicolumn{1}{l}{$l=1$} & \multicolumn{1}{l}{$l=2$} & \multicolumn{1}{l}{$l=1$} & \multicolumn{1}{l|}{$l=2$} \\ \hline
-% 1000                              & 4.58                        & 2.69                        & 4.87                        & 4.58                        \\
-% 2000                              & 3.89                        & 1.62                        & 4.31                        & 2.6                         \\
-% 5000                              & 3.54                        & 1.12                        & 3.84                        & 1.7                         \\
-% 10000                             & 3.35                        & 1.05                        & 3.54                        & 1.5                         \\
-% 20000                             & 3.16                        & 0.87                        & 3.41                        & 1.41                        \\
-% 40000                             & 2.92                        & 0.86                        & 3.34                        & 1.25                        \\
-% 60000                             & 2.92                        & 0.81                        & 3.27                        & 1.34     \\ \hline                  
-% \end{tabular}
-% \end{table}
-
-Due to its separability and non-angularly spaced filters, separable scattering does not perform as well compared to 2D scattering when using simple classifiers, such as \ac{lda} \citep{lda}.
-To illustrate that this performance discrepancy is not of significant consequence in a \ac{nn} setting, we test performance on the full dataset utilising a simple architecture.
-The neural network architecture used has an input layer with 256 neurons, followed by two hidden layers with 128 and 64 neurons respectively.
-The output layer has 10 neurons - one for each digit.
-Input and hidden layers are followed by a batch norm layer \citep{batchnorm} and ReLU activation function \citep{relu}.
-The output layer is followed by a softmax function.
-The Adam optimiser \citep{adam} is used with cross-entropy loss, a batch size of 256 and learning rate of $3 \cdot 10^{-5}$.
-5000 of the 60000 training samples are reserved for validation and removed from the training set.
-Training is stopped when validation loss starts to increase.
-The \ac{nn} model is initialised with random weights, and the experiment is repeated 50 times.
-Different invariance scales were tested, and the best results are reported in table \ref{tab:mnistnn}.
-Tests are repeated for $l \in \{1, 2\}$ levels of scattering.
-
-\subsection{MedMNIST3D}
-
-The MedMNIST3D datasets are a subset of the MedMNIST dataset group \citep{medmnist}, where each 3D dataset contains $28\times 28 \times 28$ images with 2, 3 or 11 classes.
-Train, test and validation data partitions are provided by the authors.
-All datasets have on the order of 1000 training samples.
-We compare the baseline \ac{nn} results provided in \citep{medmnist} with separable scattering features combined with a simple \ac{nn} classifier.
-
-We use \ac{nn} classifier with an input layer containing 1024 neurons, followed by two hidden layers with 512 and 256 neurons respectively.
-Input and hidden layers layers are each followed by a batch norm layer \citep{batchnorm} and a ReLU non-linearity \citep{relu}.
-For datasets with two classes, the output layer is a single neuron followed by a sigmoid activation function.
-For datasets with more than two classes, the output layer has a size equal to the number of classes, followed by a softmax activation.
-Binary cross-entropy loss are used for datasets with two classes, otherwise cross-entropy loss is used.
-The Adam optimiser \citep{adam} with a learning rate of $1 \cdot 10^{-5}$ is used.
-All other configuration parameters are identical to the model used for the MNIST dataset.
-
-A single level of scattering coefficients are computed, with $Q = (2, 2, 2)$ and $d = (4, 4, 4)$.
-Many MedMNIST3D datasets tend to be unbalanced, implying that \ac{auc} is a more reliable metric to measure model performance.
-The results are shown in table \ref{tab:medmnist}, in which \ac{sota} \ac{auc} performance is achieved for the Organ, Adrenal and Vessel datasets.
-Table \ref{tab:medmnist} indicates the number of classes ($c$) for each of the datasets, with accuracy (ACC) also shown for reference.
-The performance of our method on non-\ac{sota} results are comparable to the other baseline \ac{nn} approaches presented in \citep{medmnist}.
-It is likely that better results can be achieved by the proposed method if the filters are made learnable and/or scattering parameters are uniquely optimised for each dataset.

### 2024-05-27 dissertation/main.tex
draft edit nearly complete, need to do refs

--- 2024-05-05 dissertation/main.tex
@@
 
-\newcommand\articleinfo[5]{%
+\newcommand\articleinfo[4]{%
 \noindent\rule{\textwidth}{0.5pt}
@@
 \noindent\rule{\textwidth}{0.25pt}%
-\section*{\centering \textsc{Abstract}}
-\begin{spacing}{1.0}    
-    \small  #5 \noindent
-\end{spacing}
-\noindent\rule{\textwidth}{0.5pt}\pagebreak
 }
@@
 \supervisor{D.J.J. Versfeld}
-\cosupervisor{J.A. Du Preez}
+\cosupervisor{J.A. du Preez}
 % \date{August}{2024} 
 \SetCopyrightOff
-\address{\AorE{Departement Siviele Ingenieurswese,\\ Stellenbosch Universiteit,\\
+\address{\AorE{Departement Elektriese en Elektroniese Ingenieurswese,\\ Stellenbosch Universiteit,\\
 Privaatsak X1, 7602 Matieland, Suid Afrika.
 }{%
-Department of Civil Engineering,\\ Stellenbosch University,\\
+Department of Electrical and Electronic Engineering,\\ Stellenbosch University,\\
 Private Bag X1, 7602 Matieland, South Africa.}}
@@
 
+\chapter*{Declaration of Publications}
+
+This dissertation includes three original papers published in peer-reviewed journals.
+The writing of these papers were the responsibility of myself and contain original work I have individually developed, under the supervision of professors Versfeld and Du Preez.
+Each paper includes a detailed statement of author contributions using the CRediT system where they appear, along with all necessary information to locate the original publications.
+
+\textit{Hierdie proefskrif sluit drie publikasies in, gepubliseer in eweknie-geëvalueerde joernale.
+Die skryf van hierdie publikasies was die hoofverantwoordelik van myself, waarvan ek die inhoud individueel ontwikkel het onder die leierskap van my toesighouers, Professore Versfeld en Du Preez.
+'n Deeglike bydraestelling word voorsien waar die publikasies in hierdie proefskrif verskyn in terme van die CRediT sisteem.}
+
+\chapter*{Acknowledgements}
+
+I would like to thank my supervisors, for encouraging me to upgrade my Master's degree to PhD, allowing me to pursue my love for academics.
+It is an honour to be able contribute something to a world filled with brilliant minds.
+
+A special thank you to my parents, who have supported me in various ways in my entire academic career.
+Without them, I would not have made it this far.
+A kind thank you to my mother, who has stood by side, and my father, who has helped me through writing countless impactful abstracts and dedicated many hours of his time to academic discussions.
+
+Finally, a heartfelt thanks to my partner, who has stood by me through it all and meticulously read every single word.
+My countless work hours and your endless patience have finally paid off.
+I love you dearly.
+
 \begin{abstract}
+
+In this dissertation by publication, we investigate the application of wavelet scattering transforms on whale vocalisations, which is presented as three published papers.
+
+We first consider how wavelet transforms can provide an alternative time-frequency (TF) decomposition for use in signal detectors.
+Detectors typically utilise the short-time Fourier transform, for which the continuous wavelet transform (CWT) is a suitable alternative.
+As an additional contribution, we improve upon the spectral entropy (SE)detector, which is known to be superior in terms of signal detection compared to energy detectors.
+
+In the second publication, we expand on the relatedness of the CWT to wavelet scattering.
+Wavelet scattering is a feature extraction method akin to the filters of a convolutional neural network (CNN) frontend and Mel-frequency cepstral coefficients (MFCCs).
+We show that wavelet scattering coefficients can be combined with a SE detector, which can be fed into a classification system.
+We further improve upon the SE detector presented in the first paper and provide a critical discussion of the challenges faced when evaluating performance of the proposed methods on a large real-world dataset.
+
+Finally, the concepts established in the previous papers are used to construct an entirely new and generalised form of the wavelet scattering transform -- separable wavelet scattering.
+We extend the definition of one-dimensional scattering to an arbitrary number of signal dimensions.
+Separable scattering benefits from various computational and conceptual advantages, which has not been generalised to this extent in the current literature.
 
@@
 \begin{abstract}[afrikaans]
+
+In hierdie publikasie proefskrif, ondersoek ons die toepassing van \textit{wavelet} verstrooiing transformasies op walvis vokaliserings, deur middel van drie gepubliseerde joernaal arikels.
+
+Eerste oorweeg on hoe \textit{wavelet} transformasies 'n alternatiewe tyd-frekwensie (TF) ont-binding vir seindetektors kan voorsien.
+Detektors maak gewoonlik gebruik van die kort-tyd Fourier transform (STFT) waarvoor die kontinu\"e \textit{wavelet} trasnform (CWT) 'n alternatief is.
+Ons verbeter ook die spektrale entropy (SE) detector, wetend dat dit beter presteer as energie-gebasseerde seindetektors.
+
+In die tweede publikasie brei ons uit op die verwantskap tussen die CWT en \textit{wavelet} verstrooiing.
+\textit{Wavelet} verstrooiing is metode wat kenmerke uit seine onttrek in 'n soorgelyke manier aan konvolusionele neurale netwerke (CNNs) en Mel-frekwensie kepstrale ko\"effisi-\"ente (MFCCs).
+Ons wys dat \textnormal{wavelet} verstrooiing met 'n SE detektor gekombineer kan word wat verder in 'n klasifikasiesisteem invloei.
+Ons bied 'n kritiese gesprek aan wat die uitdagings van sisteemevaluasie deur middle van 'n groot praktiese datastel bespreek.
+
+Laastens word die konsepte wat in die vorige publikasies vasgestel is gerbuik om 'n splinter-nuwe en veralgemene \textnormal{wavelet} verstrooiing metode te definieer -- skeibare \textnormal{wavelet} ver-strooiing.
+Ons veralgemeen die eendimensionele \textnormal{wavelet} verstrooiing to 'n arbitr\^ere aantal intree-dimensies wat nog nie tevore in die bestaande literatuur ondersoek is nie.
 

### 2024-06-22 dissertation/content/conclusion.tex
added jdp corrections

@@
 
-In this dissertation, we investigate the viability of wavelet transforms as alternative \ac{tf} decompositions in for detecting and classifying whale calls.
+In this dissertation, we investigate the viability of wavelet transforms as alternative \ac{tf} decompositions for detecting and classifying whale calls.
 
@@
 
-From the foundation established in chapter \ref*{chap:p2} and inspired by joint-\ac{tf} scattering, we generalise the notation of using separable wavelet filters in a multi-dimensional scattering transform in chapter \ref*{chap:p3}.
+From the foundation established in chapter \ref*{chap:p2} and inspired by joint-\ac{tf} scattering, we generalise the notion of using separable wavelet filters in a multi-dimensional scattering transform in chapter \ref*{chap:p3}.
 We show that, in a modern \ac{ml}-setting, it can be equally effective compared to other wavelet scattering filterbanks, yet provides significant advantages due to filter separability, which includes greater configuration freedoms and computational advantages.
 
-Further work includes deploying separable wavelet scattering and the improved \ac{se} detector from chapter \ref*{chap:p2} in a \ac{pam} detection and classification system, while evaluating its performance compared to \ac{nn}-based approaches.
+Further work includes deploying separable wavelet scattering and the improved \ac{se} detec\-tor from chapter \ref*{chap:p2} in a \ac{pam} detection and classification system, while evaluating its perfor\-mance compared to \ac{nn}-based approaches.
 
-The papers presented in this dissertation naturally introduce wavelet scattering to both signal detection and classification for \ac{pam} systems, thereby addressing some of the shortcomings of \acp{nn}.
-We develop a new type of scattering which is particularly well suited to the requirements of \ac{pam}.
+The papers presented in this dissertation introduces wavelet scattering to both signal detection and classification for \ac{pam} systems, thereby addressing some of the shortcomings of \acp{nn}, while also generalising an entirely new wavelet scattering transform.

### 2024-06-22 dissertation/content/introduction.tex
added jdp corrections

@@
 This dissertation focuses on \ac{tf} decompositions, and how they can be leveraged in order to reduce the usage or even entirely replace \acp{nn} in \ac{pam} systems.
-At the end of this investigative journey, we naturally arrive an entirely new generalisation of a feature extraction method based on \ac{ct} wavelets, which can be utilised as a drop-in replacement for a \ac{cnn} front-end.
+At the end of this investigative journey, we arrive at an entirely new generalisation of a feature extraction method based on \ac{ct} wavelets, which can be utilised as a drop-in replacement for a \ac{cnn} front-end.
 Wavelet scattering features can be further improved with deep learning across filter channels \cite{nnphasecollapse}, still making it useful in a deep learning context.
@@
     \item Poor \ac{snr}, often less than 0 dB.
-    \item Incomplete or mislabeled training data, due to poor \ac{snr} and human error.
+    \item Incomplete or mislabelled training data, due to poor \ac{snr} and human error.
     \item Small datasets, which only cover a portion of environment variability, making it unsuitable for a general model.
@@
 
-All the above factors seem demotivate the use of \acp{nn} for the following reasons: principle of operation can be unpredictable; computation time; it requires reliable training data or special techniques to mitigate problematic data.
+All the above factors arguably discourages the use of \acp{nn} for the following reasons: principle of operation can be unpredictable; computation time; it requires reliable training data or special techniques to mitigate problematic data.
 Although it is possible to address these challenges, it is worth exploring non-\ac{nn} methods.
@@
 Contextual chapters serve to unify the content of the published papers and show the natural progression of analysing the function and purpose \ac{tf} decompositions.
-All relevant literature is therefore gradually introduced, since each paper conceptually expanding on the previous paper's ideas and contributions.
+All relevant literature is therefore gradually introduced, seeing that each paper conceptually expands on the previous paper's ideas and contributions.
 
@@
 
-\section{Journal Papers and Contributions}
+\section{Journal Papers}
 
@@
 Each chapter containing a publication declares detailed author contributions in terms of the CRediT system.
+
+\clearpage
+\section{Contributions}
+This section serves as a summary of the contributions made by each paper. Contributions are ordered by most to least significant.
+
+\begin{enumerate}
+    \item \textbf{Soft-output Signal Detection for Cetacean Vocalizations using Spectral Entropy, K-means Clustering and the Continuous Wavelet Trans\-form}:
+    \begin{enumerate}
+        \item Consider the \ac{cwt} as an alternative \ac{tf} decomposition for \ac{se} calculation.
+        \item Introduce a method to scale threshold output into an interpretable and para\-meter agnostic pseudo-probability.
+        \item Introduce median filtering to improve \ac{se} detectors.
+    \end{enumerate}
+    \item \textbf{Detecting and Classifying Whale Calls with Wavelet Scattering and Spec\-tral Entropy}:
+    \begin{enumerate}
+        \item Introduce \ac{ws} to the field of \ac{pam} for classification and detection tasks.
+        \item Use \ac{ws} in combination with a \ac{se} detector, thereby demonstrating that classi\-fication and detection systems can coexist effectively.
+        \item Improve on the previous paper's threshold scaling method by providing true proba\-bilistic output from the \ac{se} detection step.
+        \item Evaluate the proposed system using practical data, providing critical discus\-sions on the reliability of typical \ac{pam} datasets. 
+    \end{enumerate}
+    \item \textbf{Separable Wavelet Scattering}: 
+    \begin{enumerate}
+        \item Introduce an entirely new family of \ac{ws} transforms, which generalises one-dimensional \ac{ws} to an arbitrary number of dimensions by using separable filters.
+        \item Demonstrate \ac{sota} results on some of the MedMNIST3D \cite{medmnist} datasets, which are a collection of 3D medical datasets to benchmark \ac{ml} models, in spirit of MNIST \cite{mnist}.
+    \end{enumerate}.
+\end{enumerate}

### 2024-06-22 dissertation/content/nomenclature.tex
added jdp corrections

@@
     $f(x)$ & a continuous function of $x$ \\
-    $f(x, y)$ & a continuous multi-variable function of $x$ and $y$ \\
+    $f(x, y)$ & a continuous multivariable function of $x$ and $y$ \\
     $f[n]$ & a discrete-time (sampled) function indexed by $n$ \\
@@
     $\mathcal{F} \{f\}$ & the Fourier transform operator acting on $f$\\
-    $\hat{f}$ & the Fourier transform of $f$ or an estimate of $f$, obvious from context\\
+    $\hat{f}$ & the Fourier transform of $f$ or an estimate of $f$, derived from context\\
     $x^*$ & the complex conjugate of $x$\\

### 2024-06-22 dissertation/content/p1_intro.tex
added jdp corrections

@@
     \label{eqn:dtft}
-    \text{DTFT}\{x\} \triangleq \hat{x}[k] = \sum_{n=0}^{N} x(t) e^{-j \frac{2\pi k n}{N}} = \left(x \otimes \upsilon_k\right)\bigg|_{n=0},
+    \text{DTFT}\{x\} \triangleq \hat{x}[k] = \sum_{n=0}^{N} x[n] e^{-j \frac{2\pi k n}{N}} = \left(x \otimes \upsilon_k\right)\bigg|_{n=0},
 \end{equation}
@@
 
-The \ac{dtft} is commonly used to quickly compute convolutions for which the \ac{fft} is a fast alorithm with $N \log N$ computational complexity.
+The \ac{dtft} is commonly used to quickly compute convolutions for which the \ac{fft} is a fast algorithm with $N \log N$ computational complexity.
 In particular, to circularly convolve signals $x$ and $y$, a fast computation is implemented as
@@
 
-Fourier transforms as decribed above are invertible operators, since they are orthogonal decompositions.
+Fourier transforms as described above are invertible operators, since they are orthogonal decompositions.
 However, for \ac{tf} analyses, invertibility is not necessarily a requirement of the application.
-As such, this disseration does not consider invertibility for the remainder of the content.
+As such, this dissertation does not consider invertibility for the remainder of the content.
 
@@
 Formally, given a signal with a fundamental frequency most accurately captured in the frequency bin $k_0$, any change in frequency by a factor $a$ will move the fundamental frequency content to the bin $a k_0$.
-This has dire consequences for lower frequency signals - the frequency resolution of lower frequency signals is significantly poorer compared to the frequency resolution of its own harmonics.
-As such, a small frequency shift of a lower frequency signal with ample harmonic content which cause large index shifts for its harmonics, i.e., the $b$'th harmonic will be shifted to index $a b k_0$.
+This has dire consequences for lower frequency signals - the frequency resolution of lower frequency signals is significantly poorer compared to the frequency resolution of that signal's harmonics.
+A small frequency shift of a low frequency signal with harmonic content which cause large index shifts for its harmonics, i.e., the $b$'th harmonic will be shifted to index $a b k_0$.
 
@@
 For example, if $a=2$, the fundamental's frequency index is shifted to $k_0 + Q$ (as opposed to $2 k_0$ for the \ac{stft}).
-This property is referred to a frequency-shift invariance, since small frequency deformations of the order $Q^{-1}$ do not significantly change the frequency index structure of the \ac{tf} decomposition \citep{1dscattering1}.
-The \ac{stft} is much more sensitive to such deformations across its entire structure for harmonically rich signals.
+This property is referred to as frequency-shift invariance, since small frequency deformations of the order $Q^{-1}$ do not significantly change the frequency index structure of the \ac{tf} decomposition \citep{1dscattering1}.
+The \ac{stft} of a signal with harmonic content is much more sensitive to such deformations across its entire structure.
 
@@
 The exponential filterbank is, in fact, the way wavelet filters are typically constructed.
-Additionally, the \ac{mfcc} filterbank attributes much of its sucess (compared to raw \ac{stft} magnitude coefficients) due to its exponentially placed filters.
+Additionally, the \ac{mfcc} filterbank attributes much of its success (compared to raw \ac{stft} magnitude coefficients) due to its exponentially placed filters.
 
@@
 
-To construct a wavelet filterbank with $\upsilon_k \in \Upsilon$ which has the exponentially-spaced frequency properties discussed in section \ref{sec:p1i:tfa}, we use a ``template'' wavelet $\uppsi(t)$, referred to as the mother wavelet \cite{waveletbook}.
+To construct a wavelet filterbank with $\upsilon_k \in \Upsilon$ which has the exponentially spaced fre\-quency properties discussed in section \ref{sec:p1i:tfa}, we use a ``template'' wavelet $\uppsi(t)$, referred to as the mother wavelet \cite{waveletbook}.
 Conventionally, wavelet filters are defined as \ac{ct} filters and then discretised in software implementations.
@@
 
-The multiplicate scaling, in addition to time scaling, ensures that $\uppsi_{\lambda_k}$ retains constant amplitude in the frequency domain, also referred to as L1 normalisation \cite{waveletbook}.
+The multiplicative scaling, in addition to time scaling, ensures that $\uppsi_{\lambda_k}$ retains constant amplitude in the frequency domain, also referred to as L1 normalisation \cite{waveletbook}.
 
@@
 Discretisations of the \ac{cwt} approximate the transform by limiting $s$ to the set $s_k \in \frac{1}{\lambda_k}, \ \lambda_k \in \Lambda$.
-Chapter \ref{chap:p1} specifically considers the application of the \ac{cwt} for signal detectors, while also using other techniques to improve traditional methods.
+Chapter \ref{chap:p1} specifically considers the application of the \ac{cwt} for signal detectors, alongside additional detector improvements.

### 2024-06-22 dissertation/content/p2_intro.tex
added jdp corrections

@@
 In audio, \acp{mfcc} are features calculated from the \ac{stft} of the signal.
-Note that a ``\ac{tf}'' decomposition, as referred to in this dissertation, does not necesarrily require a time variable: spatial variables are often present instead.
+Note that a ``\ac{tf}'' decomposition, as referred to in this dissertation, does not necessarily require a time variable: spatial variables are often present instead.
 In this sense, a \ac{cnn} is can also be considered as a ``\ac{tf}'' decomposition, where multi-dimensional index variables (or index vectors) are used for ``time'' and frequency:
@@
 \end{equation}
-The filters $\Phi_i$ are not necesarrily multi-dimensional, and may only operate along a single dimension of $X$.
+The filters $\Phi_i$ are not necessarily multi-dimensional, and may only operate along a single dimension of $X$.
 If $\Phi_i$ is an averaging (low-pass) filter, some additional stability can be observed in $\tilde{X_i}$ compared with $X$.
 This allows for the introduction of invariances to various deformations applied to $X$, which can be greatly beneficial for some applications \citep{cnninvariance}.
-In a sense, cascased structures of $X$ and $\tilde{X_i}$ are defined in a \ac{cnn} for which the network learns the required invariances \cite{2dscattering}.
+In a sense, cascaded structures of $X$ and $\tilde{X_i}$ are defined in a \ac{cnn} for which the network learns the required invariances \cite{2dscattering}.
 
@@
 Invariance filters ($\Phi_i$) can also have additional properties that makes it favourable for audio applications.
-For example, many applications ``blur'' (low-pass) the TF decomposition \citep{seyicwt, speechmodulation}, which can reduce the impact of noise fluctuations, and can also serve to suppress small frequency fluctuations across adjascent frequency bins (if the blurring kernel $\Phi_i$ filters across frequency as well).
+For example, many applications ``blur'' (low-pass) the TF decom\-position \citep{seyicwt, speechmodulation}, which can reduce the impact of noise fluctuations, and can also serve to suppress small frequency fluctuations across adjacent frequency bins (if the blurring kernel $\Phi_i$ filters across frequency as well).
 The blurring operation can improve the performance of signal detectors and classifiers.

### 2024-06-22 dissertation/content/p3_intro.tex
added jdp corrections

@@
     \item Harmonic scattering \citep{harmonicscattering} (3 dimensions).
-The spherical harmonics are used as filters which consitute an orthogonal basis in 3 dimensions.
+The spherical harmonics are used as filters which constitute an orthogonal basis in 3 dimensions.
 These filters have ``lobes'' that extend in the along various directions in the frequency domain, depending on the harmonic.
@@
 Separable filters have been shown to be effective in a \ac{cnn} settings, while also allowing for much quicker convolution operations \cite{separablecnn,separablecnn2} when computed directly.
-Separable filters also allow for the independance of each dimension, which can be particularly useful for optimisations such as downsampling.
+Separable filters also allow for the independence of each dimension, which can be particularly useful for optimisations such as downsampling.
 This type of scheme is illustrated by higher-dimensional dyadic wavelet transforms such as the \ac{dwt} \cite{waveletbook}.
@@
 
-The type of scattering transform used is usually highly dependent on the signal dimen\-sionality -- there is no unifiying multi-dimensional analogy of 1D scattering as is present with dyadic wavelet transforms \citep{waveletbook}.
+The type of scattering transform used is usually highly dependent on the signal dimen\-sionality -- there is no unifying multi-dimensional analogy of 1D scattering as is present with dyadic wavelet transforms \citep{waveletbook}.
 We construct this generalisation in the chapter \ref*{chap:p3} via separable filters.

### 2024-06-22 dissertation/content/paper1/main.tex
added jdp corrections

@@
 \articleinfo{Soft-output Signal Detection for Cetacean Vocalizations using
-Spectral \break Entropy, K-means Clustering and the Continuous Wavelet Transform}{Ecological Informatics}{15 May 2024}{doi}
+Spectral \break Entropy, K-means Clustering and the Continuous Wavelet Transform}{Ecological Informatics}{May 2023}{https://doi.org/10.1016/j.ecoinf.2023.101990}
 
@@
 
-\textbf{D.J.J. Verseld}: data collection; writing -- review and editing; supervision.
+\textbf{D.J.J. Versfeld}: data collection; writing -- review and editing; supervision.
 
-\textbf{J.A. Du Preez}: writing -- review and editing; supervision.
+\textbf{J.A. du Preez}: writing -- review and editing; supervision.
 
@@
 \begin{equation}
-\label{eqn:specdist}
+\label{eqn:specdistp1}
 	P[k,m] = \frac{S[k,m]}{\sum_{j=0}^{N-1} S[j,m]}.
@@
 \begin{equation}
-\label{eqn:entropy}
+\label{eqn:entropyp1}
 	H[m] = - \sum_{k=0}^{N/2} P[k,m] \log (P[k,m]).
@@
 
-SE is calculated by substituting the CWT equivalent sinusoidal power from equation (\ref{eqn:cwtps}) into equations (\ref{eqn:specdist}) and (\ref{eqn:entropy}).
+SE is calculated by substituting the CWT equivalent sinusoidal power from equation (\ref{eqn:cwtps}) into equations (\ref{eqn:specdistp1}) and (\ref{eqn:entropyp1}).
 

### 2024-06-22 dissertation/content/paper2/main.tex
added jdp corrections

@@
 \articleinfo{Detecting and classifying Whale Calls with Wavelet Scattering
-and Spectral Entropy}{the Journal of the Acoustical Society of America (JASA)}{15 May 2024}{doi}
+and Spectral Entropy}{the Journal of the Acoustical Society of America (JASA)}{15 May 2024}{TBD}
 
@@
 
-\textbf{D.J.J. Verseld}: writing -- review and editing; supervision
-
-\textbf{J.A. Du Preez}: writing -- review and editing; supervision.
+\textbf{D.J.J. Versfeld}: writing -- review and editing; supervision
+
+\textbf{J.A. du Preez}: writing -- review and editing; supervision.
 

### 2024-06-22 dissertation/content/paper3/main.tex
added jdp corrections

@@
 
-\articleinfo{Seperable Wavelet Scattering}{IEEE Signal Processing Letters}{15 May 2024}{doi}
+\articleinfo{Seperable Wavelet Scattering}{IEEE Signal Processing Letters}{15 May 2024}{TBD}
 
@@
 
-\textbf{D.J.J. Verseld}: writing -- review and editing; supervision.
+\textbf{D.J.J. Versfeld}: writing -- review and editing; supervision.
 
-\textbf{J.A. Du Preez}: writing -- review and editing; supervision.
+\textbf{J.A. du Preez}: writing -- review and editing; supervision.
 

### 2024-06-22 dissertation/main.tex
added jdp corrections

@@
 \usepackage{titlesec}
-\titleformat{\chapter}
-  {\LARGE\bfseries}{\thechapter}{1em}{}
-\titlespacing*{\chapter}{0pt}{3.5ex plus 1ex minus .2ex}{2.3ex plus .2ex}
+
+\titleformat{\chapter}[display]
+{\bfseries\Large}
+{\LARGE\chaptertitlename \ \thechapter}
+{0pt}
+{\vspace{0pt}%
+\LARGE}
+\titlespacing*{\chapter}{0pt}{0pt}{1em}
+
+\titleformat{\section}
+{\bfseries\Large}
+{\Large\thesection}
+{0.75em}
+{\Large}
+
+\titlespacing*{\section}{0pt}{1em}{0.75em}
 
@@
 \begin{document}
-\title{\AorE{Wavelet Scattering Transformasies Toegepas op Walvis Vokaliserings}{Wavelet Scattering Transforms Applied to Whale Vocalisations}}
+\title{\AorE{Wavelet Verstrooiing Transformasies Toegepas op Walvis Vokaliserings}{Wavelet Scattering Transforms Applied to Whale Vocalisations}}
 \author{M.W. Rademan}{Marco Wiehann Rademan}
@@
 
-I would like to thank my supervisors, for encouraging me to upgrade my Master's degree to PhD, allowing me to pursue my love for academics.
+I would like to thank my supervisors, Professors Versfeld and Du Preez, for encouraging me to upgrade my Master's degree to PhD, allowing me to pursue my love for academics.
 It is an honour to be able contribute something to a world filled with brilliant minds.
@@
 
-Finally, a heartfelt thanks to my partner, who has stood by me through it all and meticulously read every single word.
+Finally, a heartfelt thanks to my partner, who meticulously read every single word and has been my pillar and primary motivation since the day I started this journey.
 My countless work hours and your endless patience have finally paid off.
@@
 Detectors typically utilise the short-time Fourier transform, for which the continuous wavelet transform (CWT) is a suitable alternative.
-As an additional contribution, we improve upon the spectral entropy (SE)detector, which is known to be superior in terms of signal detection compared to energy detectors.
+As an additional contribution, we improve upon the spectral entropy (SE) detector, which is known to be superior in terms of signal detection compared to energy detectors.
 
@@
 
-Eerste oorweeg on hoe \textit{wavelet} transformasies 'n alternatiewe tyd-frekwensie (TF) ont-binding vir seindetektors kan voorsien.
-Detektors maak gewoonlik gebruik van die kort-tyd Fourier transform (STFT) waarvoor die kontinu\"e \textit{wavelet} trasnform (CWT) 'n alternatief is.
-Ons verbeter ook die spektrale entropy (SE) detector, wetend dat dit beter presteer as energie-gebasseerde seindetektors.
+Aanvanklink oorweeg ons hoe \textit{wavelet} transformasies 'n alternatiewe tyd-frekwensie (TF) ont-binding vir seindetektors kan voorsien.
+Detektors maak gewoonlik gebruik van die kort-tyd Fourier transform (STFT) waarvoor die kontinu\"e \textit{wavelet} transform (CWT) 'n alternatief is.
+Ons verbeter ook die spektrale entropie (SE) detektor, wetend dat dit beter presteer as energie-gebasseerde seindetektors.
 
 In die tweede publikasie brei ons uit op die verwantskap tussen die CWT en \textit{wavelet} verstrooiing.
-\textit{Wavelet} verstrooiing is metode wat kenmerke uit seine onttrek in 'n soorgelyke manier aan konvolusionele neurale netwerke (CNNs) en Mel-frekwensie kepstrale ko\"effisi-\"ente (MFCCs).
-Ons wys dat \textnormal{wavelet} verstrooiing met 'n SE detektor gekombineer kan word wat verder in 'n klasifikasiesisteem invloei.
+\textit{Wavelet} verstrooiing is metode wat kenmerke uit seine onttrek in 'n soorgelyke manier as konvolusionele neurale netwerke (CNNs) en Mel-frekwensie kepstrale ko\"effisi-\"ente (MFCCs).
+Ons wys dat \textnormal{wavelet} verstrooiing met 'n SE detektor gekombineer kan word wat verder in 'n klassifikasiesisteem invloei.
 Ons bied 'n kritiese gesprek aan wat die uitdagings van sisteemevaluasie deur middle van 'n groot praktiese datastel bespreek.
 
-Laastens word die konsepte wat in die vorige publikasies vasgestel is gerbuik om 'n splinter-nuwe en veralgemene \textnormal{wavelet} verstrooiing metode te definieer -- skeibare \textnormal{wavelet} ver-strooiing.
-Ons veralgemeen die eendimensionele \textnormal{wavelet} verstrooiing to 'n arbitr\^ere aantal intree-dimensies wat nog nie tevore in die bestaande literatuur ondersoek is nie.
+Laastens word die konsepte wat in die vorige publikasies gevestig is, gebruik om 'n splinter-nuwe en veralgemene \textnormal{wavelet} verstrooiing metode te definieer -- skeibare \textnormal{wavelet} ver-strooiing.
+Ons veralgemeen die eendimensionele \textnormal{wavelet} verstrooiing tot 'n arbitr\^ere aantal intree-dimensies, wat nie tevore in die bestaande literatuur ondersoek is nie.
 
